### 2023-08-18 header.tex
Upload all files

% **************GENERATED FILE, DO NOT EDIT**************

\title{Pigeons.jl: Distributed sampling from intractable distributions}

\author[1]{Nikola Surjanovic}
\author[1]{Miguel Biron-Lattes}
\author[2]{Paul Tiede}
\author[3]{Saifuddin Syed}
\author[1]{Trevor Campbell}
\author[1]{Alexandre Bouchard-Côté}
\affil[1]{University of British Columbia}
\affil[2]{Harvard University}
\affil[3]{University of Oxford}

\keywords{Distributed computation, Bayesian inference, parallelism invariance, 
multi-threading, Message Passing Interface, Markov chain Monte Carlo}

\hypersetup{
pdftitle = {My JuliaCon proceeding},
pdfsubject = {JuliaCon 2019 Proceedings},
pdfauthor = {1st author, 2nd author, 3rd author},
pdfkeywords = {Julia, Optimization, Game theory, Compiler},
}

### 2023-08-18 paper.tex
Upload all files

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}

\usepackage[autonum, blackhypersetup]{shortex}
\usepackage{xcolor}
\usepackage[font=scriptsize]{subcaption}
\usepackage{forest}

\newcommand{\NS}[1]{\textcolor{red}{#1}}
\def\bfx{\mathbf{x}}
\def\bfX{\mathbf{X}}

\definecolor{julia1}{rgb}{0, 0.60, 0.98}
\definecolor{julia2}{rgb}{0.89, 0.43, 0.28}
\definecolor{julia3}{rgb}{0.24, 0.64, 0.30}
\definecolor{julia4}{rgb}{0.76, 0.44, 0.82}
\definecolor{julia5}{rgb}{0.67, 0.55, 0.09}
\definecolor{julia6}{HTML}{00aaae}
\definecolor{julia7}{HTML}{ed5e93}
\definecolor{julia8}{HTML}{c68225}

\begin{document}

\input{header}

\maketitle

\begin{abstract}

We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides simple APIs to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.

\end{abstract}

\input{src/introduction.tex}
\input{src/examples.tex}
\input{src/PT.tex}
\input{src/SPI.tex}
\input{src/related.tex}
% \input{src/benchmarks.tex}
\input{src/conclusion.tex}

\section{Acknowledgements}
NS acknowledges the support of a Vanier Canada Graduate Scholarship. 
PT acknowledges the support of the Black Hole Initiative at Harvard University, 
which is funded by grants from the John Templeton Foundation and the 
Gordon and Betty Moore Foundation to Harvard University, 
as well as support by the National Science Foundation grants AST-1935980 
and AST-2034306 and the Gordon and Betty Moore Foundation (GBMF-10423).
SS acknowledges the support of EPSRC grant EP/R034710/1 CoSines.
ABC and TC acknowledge the support of an NSERC Discovery Grant.
We also acknowledge use of the ARC Sockeye computing platform from the
University of British Columbia, as well as cloud computing resources provided 
by Oracle.

\input{bib.tex}

\end{document}

% Inspired by the International Journal of Computer Applications template

### 2023-08-18 src/PT.tex
Upload all files

\section{Parallel tempering}
\label{sec:PT}
Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \cite{syed2021nrpt}, which we outline in Algorithm~\ref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

\medskip 
In this section we assume a basic understanding of 
Markov chain Monte Carlo (MCMC) methods.
For readers unfamiliar with MCMC, it is important to know that 
it is a method to obtain approximate samples from a distribution. 
More specifically, it is a class of algorithms for simulating a Markov chain of 
states that look like draws from the target distribution $\pi$.
However, with traditional 
MCMC methods, the samples may be heavily correlated instead of independent 
and may fail to sufficiently explore the 
full space of the distribution (see \cref{fig:bimodal}); 
PT is a method that aims to address these issues.
For a more in-depth review of PT, we refer readers to \cite{surjanovic2022vpt}.

\subsection{Overview of PT}
Suppose that we would like to estimate integrals involving $\pi$, such as 
$\int f(x) \pi(x) \, \dee x$. 
These integrals may be multivariate and even include combinations of 
continuous and discrete variables (where sums replace integrals in the discrete 
case).
One method is to obtain samples from $\pi$ to approximate such integrals.
Often, the distribution 
$\pi$ can be challenging for traditional MCMC methods---such 
as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
because of its structure.
For example, in a bimodal example such as the one illustrated in 
Figure~\ref{fig:bimodal},
traditional methods might remain in one of the two modes for an extremely 
long period of time.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.3\textwidth]{img/bimodal.pdf}
    \caption{A simple bimodal distribution from which traditional 
    MCMC methods would struggle to obtain samples. Blue lines display output 
    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
    The sampler in this figure is visibly trapped in one of the two modes.}
    \label{fig:bimodal}
\end{figure}

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in Figure~\ref{fig:path}.

\begin{figure}[t]
    \centering
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{subfigure}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.

\subsection{Local exploration and communication}
A struct that is useful for the implementation of PT is the \texttt{Replica}, 
which we will often refer to in the remainder of this section. 
A single \texttt{Replica} struct stores a \texttt{state} variable and a \texttt{chain} 
integer, among other entries. 
At the beginning of PT, $N$ \texttt{Replicas} are created, 
one for each distribution on the annealing path, and 
the chain entries in the $N$ replicas are initialized at $1,2,\ldots,N$, respectively. 
For a given \texttt{Replica}, if the \texttt{chain} number is $i$ and the \texttt{state}
is $x$, then this means that the sample corresponding to the $i$-th distribution in 
the sequence $\pi_1, \ldots, \pi_N$ is currently at location $x$.

\medskip 
In the local exploration phase,
each \texttt{Replica}'s state is modified using an MCMC move targeting $\pi_i$,
where $i$ is given by \texttt{Replica.chain}.
The MCMC move can either modify \texttt{Replica.state} in-place, or modify the 
\texttt{Replica}'s \texttt{state} field. 
This operation is indicated by the \texttt{local\_exploration} function in 
\cref{alg:distributed_PT}.

\medskip 
In the communication phase, PT proposes swaps between pairs of replicas. 
In principle, there are two equivalent ways to do a swap.
In the first implementation, the \texttt{Replica}s 
could exchange their \texttt{state} fields.
Alternatively, they could exchange their \texttt{chain} fields.
Because we provide distributed implementations, we use the latter as it ensures that 
the amount of data that needs to be exchanged between two machines during a swap 
can be made very small (two floating point numbers), resuling in 
an exchange of $O(N)$ messages of size $O(1)$. 
Note that this cost does not vary with the dimensionality of the state space, 
in contrast to the first implementation that would transmit 
$O(N)$ messages of size $O(d)$, where $d$ is the dimension of the state space,
incurring a potentially very high communication cost for large values of $d$.

\subsection{Distributed implementation}
A distributed implementation of PT is presented in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
We present the algorithm from the perspective that the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. 
However, Pigeons.jl also allows for the more general case where the number of 
machines is not necessarily equal to $N$.

\subsubsection{The \texttt{PermutedDistributedArray}}
\label{sec:permuted_dist_array}
Recall that our theoretical $O(1)$ message size for communication between two machines 
is achieved by exchanging \texttt{chain} indices between \texttt{Replica}s 
instead of their \texttt{state}s.
One difficulty can be encountered in a distributed implementation, which we 
illustrate with the following simple example. 
Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
machine 1 is exploring chain 2 while machine 4 is exploring chain 3.
At one point, the \texttt{Replica} at \texttt{chain} 2 (machine 1) might need to exchange
\texttt{chain} indices with the \texttt{Replica} that has \texttt{chain} index 3 (machine 4). 
However, a priori it is not clear how machine 1 should know that it should communicate 
with machine 4 because it has no knowledge about the chain indices on the other machines.

\medskip 
To resolve this issue, we introduce a special data structure called a 
\texttt{PermutedDistributedArray}. In the case where the number of replicas is 
equal to the number of available machines, the construction is quite simple
and effectively results in each machine storing one additional integer. 
Considering the same example above, the solution to the problem is to have
machine 1 (\texttt{chain} 2 wanting to communicate with \texttt{chain} 3) 
communicate with \textit{machine} 3. 
Machine 3 stores in its \texttt{dist\_array} variable of type 
\texttt{PermutedDistributedArray} the value 4, which is the machine number that 
stores \texttt{chain} 3. By updating these \texttt{PermutedDistributedArray} 
variables at each communication step, we can ensure that each machine $j$ is aware 
of which machine number currently stores \texttt{chain} $j$. 
Therefore, the machines act as keys in a dictionary for the communication 
permutations.
An illustration of communication between four machines is provided in 
\cref{fig:index_process}.
We note that with a \texttt{PermutedDistributedArray} we make special assumptions 
on how we access and write to the array elements. 
Several MPI processes cooperate, with each machine storing 
data for a slice of this distributed array, and at each time 
step an index of the array is manipulated by exactly one machine.
% With these assumptions we achieve read/write costs that are 
% near-constant in the number of machines participating.

\begin{figure*}[t] 
  \centering 
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_1_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_1.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_2_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_2.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_3_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_3.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_4_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_4.pdf}
  \end{subfigure}
  \caption{
    A summary of communication between four machines with $N=4$ PT chains.
    Circles represent machines with the machine number in the center of the circle. 
    Different colours are also used for different machines.
    Numbers in parantheses represent the \texttt{chain} 
    that is currently being explored by that machine. 
    Arrows indicate communication occuring between machines to exchange \texttt{chain} 
    indices.
    To keep track of which chains are stored on which machine, we introduce the 
    \texttt{dist\_array} of type \texttt{PermutedDistributedArray}, described in 
    \cref{sec:permuted_dist_array}, which is of length $N$. 
    The $j$-th element of \texttt{dist\_array} at a given time step indicates which 
    machine number is storing \texttt{chain} index $j$.
    The bolded curve in the figures to the right indicate the trajectory of the first 
    \texttt{Replica} over the course of each of the communication steps.      
    \textbf{Top left:} the PT replicas are initialized and the first communication step 
    is proposed. At this time step, $\texttt{dist\_array} = [1,2,3,4]$.
    \textbf{Top right:} after the first successful communication step, we now have 
    $\texttt{dist\_array} = [2,1,4,3]$.
    \textbf{Bottom left:} The second communication step is completed and 
    $\texttt{dist\_array} = [2,4,1,3]$.
    \textbf{Bottom right:} The third communication step is completed and 
    $\texttt{dist\_array} = [4,2,3,1]$.}
  \label{fig:index_process}
\end{figure*}

\subsubsection{MPI implementation details}
We use the MPI.jl \cite{byrne2021mpi} package to support communication between machines. 
In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
functions \texttt{send()}, \texttt{receive!()}, 
and \texttt{waitall()}. 
The function \texttt{send()} has three arguments in our pseudocode: the first argument 
is the object to be sent, the second is the machine number to which the object should 
be sent, and finally the third argument is a unique tag for this specific send/receive 
request at this time step and on this machine, generated by \texttt{tag()}.
The \texttt{tag()} function can be any function that allows one to uniquely identify 
a tag for a given send/receive request on a given machine at any given time step $t$.
The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
except that the first argument 
specifies the object to which data should be written directly. Finally, 
\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
of machines to complete.

\medskip 
In \cref{alg:distributed_PT} we also introduce the functions 
\texttt{permuted\_get()} and \texttt{permuted\_set!()}. 
Given the current machine's \texttt{dist\_array} and the 
\texttt{partner} chain with which it should communicate, 
\texttt{permuted\_get()} returns the machine number that holds the \texttt{partner} chain. 
Given the current machine's \texttt{dist\_array}, \texttt{chain} number, and machine number $j$,
\texttt{permuted\_set!()} updates the \texttt{PermutedDistributedArray} variables 
with the updated permutation.

\begin{algorithm}[t]
	\begin{algorithmic}[1]
    \Require Initial state $x_0$, sequence of distributions $\{\pi_i\}_{i=1}^N$, 
      number of iterations $T$, machine number $j$
    \State $\texttt{chain} \gets j$ \Comment{current \texttt{chain} number}
    \State $\texttt{dist\_array}[j] \gets j$ \Comment{\texttt{chain} $j$ is on machine $j$}

    \For{$t$ {\bf in} 1, 2, \dots, $T$}
		  \If{$t$ is even} \Comment{choose between even or odd swap}
		    \State $P \gets \{i: 1 \le i < N, i \text{ is even} \}$
		  \Else
		    \State $P \gets \{i: 1 \le i < N, i \text{ is odd} \}$
		  \EndIf

      \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
      \State $\texttt{partner} \gets \texttt{nothing}$
      \If{$\texttt{chain} \in P$}
        \State $\texttt{partner} \gets \texttt{chain}+1$
      \ElsIf{$\texttt{chain}-1 \in P$}
        \State $\texttt{partner} \gets \texttt{chain}-1$
      \EndIf
      \If{$\texttt{partner != nothing}$ \textbf{and} \text{swap is accepted}}
        \State $\texttt{to\_machine} \gets \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
          \Comment{retrieve \texttt{dist\_array}[\texttt{partner}]}
        \State $\texttt{send\_tag} \gets \texttt{tag}(t, \texttt{chain}, j)$
        \State $\texttt{send}(\texttt{chain}, \texttt{to\_machine}, \texttt{send\_tag})$
        \State $\texttt{receive\_tag} \gets \texttt{tag}(t, \texttt{chain}, \texttt{to\_machine})$
        \State $\texttt{receive!}(\texttt{chain}, \texttt{to\_machine}, \texttt{receive\_tag})$
        \State $\texttt{waitall(requests)}$ 
          \Comment{wait for MPI requests for this pair of machines to complete}
        \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$
          \Comment{set value of \texttt{dist\_array}[\texttt{chain}] to \texttt{to\_machine}}
      \EndIf
      \State $c_t \gets \texttt{chain}$ 
		\EndFor
    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
	\end{algorithmic}
  \caption{Distributed PT on machine $j$ (one replica per machine)}
  \label{alg:distributed_PT}
\end{algorithm}

### 2023-08-18 src/SPI.tex
Upload all files

\section{Strong parallelism invariance}
\label{sec:PI_causes}
In this section we describe potential violations of 
strong parallelism invariance (SPI) that can occur in a distributed setting. 
We also explain how we avoid these issues by using special 
distributed reduction schemes and splittable random number generators. 
Insights provided in this section can be applied 
to general distributed software beyond Julia.

\medskip 
We have identified two factors that can cause violations of our previously-defined 
SPI that standard Julia libraries 
do not automatically take care of:
\begin{enumerate}   
  \item \textbf{Non-associativity of floating point operations:} When 
  several workers perform distributed reduction of floating point values, 
  the output of this reduction will be slightly different depending on the order taken 
  during reduction. 
  When these reductions are then fed into further random operations, 
  this implies that two randomized algorithms with the same seed but using a 
  different number of workers will eventually arbitrarily diverge.

  \item \textbf{Global, thread-local, and task-local random number generators:} 
  These are the dominant approaches to parallel random number generators in current languages, 
  and an appropriate understanding of these RNGs is necessary. In particular, 
  in Julia it is important to understand the behaviour of the \texttt{@threads} macro.
\end{enumerate}
Our focus in the remainder of this section is to describe how our implementation 
solves the two above issues.

\subsection{Distributed reduction and floating point non-associativity}
After each round of PT, the machines need to exchange information such as the 
average swap acceptance probabilities, statistics to adapt a variational 
reference and adjust annealing parameters, and so on. 
For instance, suppose our \texttt{state} is univariate and real-valued and 
that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
is visited as well as the mean of the univariate \texttt{state}s from the target chain.
To obtain the final estimate of the mean of the target distribution, we would like 
to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
the number of times that each \texttt{Replica} visited the target chain.   
This process involves summing floating-point values that are located on each 
of the \texttt{Replica}s/machines and is an example of a reduction scheme.

\medskip 
To illustrate why distributed reduction with floating point values can violate 
strong parallelism invariance if not properly implemented, we consider the 
following toy example. 
Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
in the following examples.
In this case, if our reduction procedure is to sum the floating point numbers, 
we know that our final answer should be approximately $36x$. 
However, depending on the exact order in which floating point addition is carried out, 
the answers might not all be the same and exactly equal to $36x$.
For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
for eight machines would be given by 
\[
  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
  \approx 978.5814582452562.
\]
In contrast, with two machines, one possible order of operations might be 
\[
  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
  \approx 978.5814582452563.
\]
A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.

\medskip 
To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
are exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in the reduction tree, irrespective of the 
number of machines available to perform the reduction. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$3x$}
            [{$1x$}]
            [{$2x$}]
          ]
          [{$7x$}
            [{$3x$}] 
            [{$4x$}] 
          ]
        ]
        [{$26x$}
          [{$11x$}
            [{$5x$}] 
            [{$6x$}] 
          ]
          [{$15x$}
            [{$7x$}] 
            [{$8x$}] 
          ]
        ]
      ]
  \end{forest}
  \caption{Adding eight floating point numbers $\cbra{1x, 2x, \ldots, 8x}$ 
  across eight machines. 
  Additions in each row of the tree can be performed in parallel. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$, indicating 
  the order of operations.}
  \label{fig:reduction_tree_8}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$6x$}
            [{$3x$}
              [{$1x$}]
              [{$2x$}]
            ]
            [{$3x$}]
          ]
          [{$4x$}]
        ]
        [{$26x$}
          [{$8x$}]
          [{$18x$}
            [{$7x$}]
            [{$11x$}
              [{$6x$}]
              [{$5x$}]
            ]
          ]
        ]
      ]
  \end{forest}
  \caption{One possible way of adding eight floating point numbers 
  $\cbra{1x, 2x, \ldots, 8x}$ across two machines. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $(((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x)$.
  Note that the order of operations in this expression is different from the one 
  presented in \cref{fig:reduction_tree_8}.}
  \label{fig:reduction_tree_2}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia2]
          ]
          [{$7x$}, fill = julia3
            [{$3x$}, fill = julia3] 
            [{$4x$}, fill = julia4] 
          ]
        ]
        [{$26x$}, fill = julia5
          [{$11x$}, fill = julia5
            [{$5x$}, fill = julia5] 
            [{$6x$}, fill = julia6] 
          ]
          [{$15x$}, fill = julia7
            [{$7x$}, fill = julia7] 
            [{$8x$}, fill = julia8] 
          ]
        ]
      ]
  \end{forest}
  \caption{Addition of eight floating point numbers across eight machines with a 
  guarantee on SPI. Each machine is represented by a different colour. 
  The final result can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
  \label{fig:reduction_tree_colour_8}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia1]
          ]
          [{$7x$}, fill = julia1
            [{$3x$}, fill = julia1] 
            [{$4x$}, fill = julia1] 
          ]
        ]
        [{$26x$}, fill = julia2
          [{$11x$}, fill = julia2
            [{$5x$}, fill = julia2] 
            [{$6x$}, fill = julia2] 
          ]
          [{$15x$}, fill = julia2
            [{$7x$}, fill = julia2] 
            [{$8x$}, fill = julia2] 
          ]
        ]
      ]
  \end{forest}
  \caption{Addition of eight floating point numbers across two machines with a 
  guarantee on SPI. Each machine is represented by a different colour. 
  The final result can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$,
  which is the same as that given by eight machines in \cref{fig:reduction_tree_colour_8}.}
  \label{fig:reduction_tree_colour_2}
\end{figure}

\subsection{Splittable random streams}
\label{sec:splittable_randoms}
Another building block towards achieving SPI is a \emph{splittable random stream}
\cite{lecuyer1988splittable,burton1992splittable}. 
Julia uses \emph{task-local} random number generators, a notion that is 
related to (but does not necessarily imply) strong parallelism invariance. 
A \emph{task} is a unit of work on a machine.
A task-local RNG would then mean that a separate RNG is 
used for each unit of work, hopefully implying strong parallelism invariance
if the number of tasks is assumed to be constant. 
Unfortunately, this is not the case when a separate task is created for each thread 
of execution in Julia.
We note that the \texttt{@threads} macro in Julia creates exactly \texttt{nthreads()}
tasks and thus \texttt{nthreads()} pseudorandom number generators.
This can break strong parallelism invariance as the output may depend on the number of 
threads.

\medskip 
To motivate splittable random streams, consider the following toy example 
that violates our notion of SPI:
\begin{lstlisting}[language = Julia]
using Random
import Base.Threads.@threads

println("Num. of threads: $(Threads.nthreads())")

const n_iters = 10000
result = zeros(n_iters)
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand()
end
println("Result: $(last(result))")
\end{lstlisting}
With eight threads, this outputs:
\begin{lstlisting}[language = Julia]
Num. of threads: 8
Result: 0.25679999169092793
\end{lstlisting}
Julia guarantees that if we rerun this code, as long as we are using eight threads, 
we will always get the same result, irrespective of the multi-threading 
scheduling decisions implied by the \texttt{@threads}-loop. 
Internally, Julia works with task-local RNGs and the \texttt{@threads} macro 
spawns \texttt{nthreads()} number of task-local RNGs.
For this reason, with a different number of threads, the result is different:
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.8785201210435906
\end{lstlisting}

In this simple example above, the difference in output is perhaps not too concerning, 
but for our parallel tempering 
use case, the distributed version of the algorithm is significantly more complex 
and difficult to debug compared to the single-threaded one. We therefore take task-local 
random number generation one step further and achieve SPI, which 
guarantees that the output is not only reproducible with respect to repetitions 
for a fixed number of threads, but also for different numbers of threads or processes.

\medskip 
A first step to achieve this is to associate one random number 
generator to each \texttt{Replica}. 
To do so, we use our SplittableRandoms.jl package, which is a Julia implementation 
of Java SplittableRandoms. Our package offers an implementation of  
SplitMix64 \cite{steele2014fast},
which allows us to turn one seed into an arbitrary collection of pseudo-independent 
RNGs. 
A quick example of how to use the SplittableRandoms.jl library is given below. 
By splitting a master RNG using the \texttt{split()} function, 
we can achieve SPI even with the use of the \texttt{@threads} macro.

\begin{lstlisting}[language = Julia]
using Random
using SplittableRandoms: SplittableRandom, split
import Base.Threads.@threads

println("Num. of threads: $(Threads.nthreads())")

const n_iters = 10000
const master_rng = SplittableRandom(1)
result = zeros(n_iters)
rngs = [split(master_rng) for _ in 1:n_iters]
Random.seed!(1)
@threads for i in 1:n_iters
    result[i] = rand(rngs[i])
end
println("Result: $(last(result))")
\end{lstlisting}
With one and eight threads, the code above outputs
\begin{lstlisting}[language = Julia]
Num. of threads: 1
Result: 0.4394633333251359
\end{lstlisting}
\begin{lstlisting}[language = Julia]
Num. of threads: 8
Result: 0.4394633333251359
\end{lstlisting}

### 2023-08-18 src/examples.tex
Upload all files

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
For further reading, we also direct readers to our growing list of examples, which can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
We begin by installing the latest official release of Pigeons.jl: 
\begin{lstlisting}[language = Julia]
using Pkg; Pkg.add("Pigeons")
\end{lstlisting}

\subsection{Targets}
To use Pigeons.jl, we must specify a target distribution, given by $\gamma$ 
in \cref{eq:normalizing_constant}.
Numerous possible types of target distributions are supported, including 
custom probability densities (specified up to a normalizing constant) written in Julia.
We also allow to interface with models written in common probabilistic programming 
languages, including:
\begin{itemize}
    \item Turing.jl \cite{ge2018turing} models (\texttt{TuringLogPotential})
    \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
    \item Comrade.jl models for black hole imaging (\texttt{ComradeLogPotential})
    \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
    (e.g. Blang \cite{bouchard2022blang} code for phylogenetic inference over combinatorial spaces) 
\end{itemize}
Additional targets are currently being accommodated and will be 
introduced to Pigeons.jl in the near future.

\medskip 
In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
applied to a non-identifiable ``coinflip'' data set, modified from the example
at \url{https://turing.ml/v0.22/docs/using-turing/quick-start}. 
The Bayesian model can be formulated as 
\[
  \label{eq:coinflip}
  p_1, p_2 &\stackrel{iid}{\sim} U(0, 1) \\    
  Y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
\]
The random variable $Y$ is the number of heads observed on $n$ coin flips
where the probability of heads is $p_1 p_2$.  
This model is non-identifiable, meaning that it is not possible to distinguish 
the effects of the two different parameters $p_1$ and $p_2$. As a consequence, 
the target distribution exhibits a complicated structure, as displayed in 
\cref{fig:coinflip_posterior}.
The density of interest corresponding to this model is 
\[
  \pi(p_1, p_2) &= \gamma(p_1, p_2)/Z,
\] 
where 
\[ 
  \gamma(p_1, p_2) &= 
    {n \choose y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
    \label{eq:coinflip_density} \\
  Z &= \int_0^1 \int_0^1 \gamma(p_1, p_2) \, \dee p_1 \, \dee p_2. \label{eq:coinflip_normalization}
\]
The distribution $\pi$ is also known as the \emph{posterior distribution} in 
Bayesian statistics.

\begin{figure}[t]
    \centering 
    \includegraphics[width=0.5\textwidth]{img/coinflip_posterior.pdf}
    \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
    with $n=100,000$ coin flips and $y=50,000$ observed heads, 
    estimated using $2^{17}$ samples from Pigeons.jl. 
    We present the pairwise plot for $p_1$ and $p_2$, as well as the estimated 
    densities of the marginal of the posterior for each of the two parameters.
    Note that because the model is non-identifiable, as we collect more data the 
    posterior distribution concentrates around the curve $p_1 p_2 = 0.5$, instead 
    of a single point, assuming that the true probability of observing heads 
    is 0.5.}
    \label{fig:coinflip_posterior}
\end{figure}

\medskip 
Suppose that we perform $n=100,000$ coin tosses and observe 
$y=50,000$ heads.
We would like to obtain samples from our posterior, $\pi$, having collected this data.
We begin by installing Turing
\begin{lstlisting}[language = Julia]
Pkg.add("Turing")
\end{lstlisting}
and then defining our Turing model and storing it in the variable \texttt{model}:
\begin{lstlisting}[language = Julia]
using Turing
@model function coinflip(n, y)
    p1 ~ Uniform(0.0, 1.0)
    p2 ~ Uniform(0.0, 1.0)
    y ~ Binomial(n, p1 * p2)
    return y
end
model = coinflip(100000, 50000)
\end{lstlisting}

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.jl.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

\begin{figure*}[t]
    \centering
    \begin{lstlisting}
          --------------------------------------------------------------------------------------
            #scans    restarts      Λ        time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
          ---------- ---------- ---------- ---------- ---------- ---------- ---------- --------- 
                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
          --------------------------------------------------------------------------------------
    \end{lstlisting}
    \caption{Standard output provided by Pigeons.jl. 
    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
    such as the number of allocations per round, time (in seconds), and estimates 
    of the log of the normalization constants. 
    The output is described in greater detail in \cref{sec:PT_diagnostics}
    and has been modified to exclude columns that are not described in the paper.}
    \label{fig:standard_out}
\end{figure*}

\subsubsection{Other targets}
As mentioned previously, it is also possible to specify targets with custom 
probability densities, as well as Stan and Turing models. 
Additionally, suppose we have some code implementing vanilla MCMC, written 
in an arbitrary ``foreign'' language such as C++, Python, R, Java, etc. 
Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
The only requirement on the foreign language is that it supports 
reading the standard input and writing to the standard output, 
as described in our online documentation.

\subsection{Outputs}
Pigeons.jl provides many useful types of output, such as: plots of samples from 
the distribution, estimates of normalization constants, summary statistics of 
the target distribution, and various other diagnostics. 
We describe several examples of possible output below.

\subsubsection{Standard output}
An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
Each row of the table in the output indicates a new tuning round in parallel tempering, 
with the \texttt{\#scans} column indicating the number of scans/samples in 
that tuning round. 
During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
tuning parameters. Other outputs include: 
\begin{itemize}
    \item \texttt{restarts}: a higher number is better. 
    Informally, PT performs well at sampling from high-dimensional and/or multi-modal 
    distributions by pushing samples from an easy-to-sample distribution (the reference) 
    to the more difficult target distribution. 
    A tempered restart happens when a sample from the reference successfully 
    percolates to the target. (See the subsequent sections for a more detailed description 
    of parallel tempering.) 
    When the reference supports i.i.d.~sampling, tempered restarts 
    can enable large jumps in the state space.

    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
    which measures the inherent difficulty of the sampling problem.
    A rule of thumb to configure the number of PT chains is also 
    given by \cite{syed2021nrpt}, where they suggest that stable performance 
    should be achieved when the number of chains is set to roughly 2$\Lambda$.
    See \cref{sec:PT_diagnostics} for more information.

    \item \texttt{time} and \texttt{allc}: the time (in seconds) and number of 
    allocations (in bytes) used in each round.

    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
    log of the normalization constant. See \cref{sec:PT_diagnostics} for more information.

    \item \texttt{min($\alpha$)} and \texttt{mean($\alpha$)}: minimum and average swap 
    acceptance rates during the communication phase across the PT chains.
    See \cref{sec:PT} for a description of PT communication.
\end{itemize}

\subsubsection{Plots}
It is straightforward to obtain plots of samples from the target distribution, 
such as trace-plots, pairwise plots, and density plots of the marginals.

\medskip 
To obtain posterior densities and trace-plots, we first 
make sure that we have the third-party 
MCMCChains.jl\footnote{\url{https://github.com/TuringLang/MCMCChains.jl}}, 
StatsPlots.jl\footnote{\url{https://github.com/JuliaPlots/StatsPlots.jl}}, and 
PlotlyJS.jl\footnote{\url{https://github.com/JuliaPlots/PlotlyJS.jl}} 
packages installed via
\begin{lstlisting}[language=Julia]
Pkg.add("MCMCChains", "StatsPlots", "PlotlyJS")
\end{lstlisting}

With the \texttt{pt} output object from before for our non-identifiable coinflip model, 
we can run the following:
\begin{lstlisting}[language=Julia]
using MCMCChains, StatsPlots, PlotlyJS
plotlyjs()
samples = Chains(
    sample_array(pt), variable_names(pt))
my_plot = StatsPlots.plot(samples)
display(my_plot)
\end{lstlisting}
The output of the above code chunk is an interactive plot that can be zoomed in or out
and exported as an HTML webpage. 
A modified static version of the output for the first parameter, $p_1$, 
is displayed in the top panel of \cref{fig:trace_density_plots},
along with a comparison to the output from a single-chain algorithm in the bottom 
panel of the same figure.

\begin{figure*}[t]
    \centering
    \begin{subfigure}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/trace_density_plots_pt.pdf}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/trace_density_plots_single_chain.pdf}
    \end{subfigure}
    \caption{
        Trace-plots for the first parameter, $p_1$, in the 
        non-identifiable coinflip Turing model.
        \textbf{Left:} Samples from Pigeons.jl using PT with 10 chains. Note that the 
        trace-plot indicates fast mixing/exploration across the state space.
        \textbf{Right:} Single-chain Markov chain Monte Carlo. Note that the 
        trace-plot explores the state space much more slowly when we do not use PT.
    }
    \label{fig:trace_density_plots}
\end{figure*}

\medskip 
To obtain pair plots, we add the 
PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
packages:
\begin{lstlisting}[language=Julia]
Pkg.add("PairPlots", "CairoMakie")
\end{lstlisting}
and then run
\begin{lstlisting}[language=Julia]
using PairPlots, CairoMakie
my_plot = PairPlots.pairplot(samples) 
display(my_plot)
\end{lstlisting}
The output of the above code chunk with an increased number of samples 
is displayed in \cref{fig:coinflip_posterior}.
\footnote{The code chunk above only works with Julia 1.9.}

\subsubsection{Estimate of normalization constant}
The (typically unknown) constant $Z$ in \cref{eq:normalizing_constant} is 
referred to as the \emph{normalization constant}. 
In many applications, it is useful to approximate this constant. 
For example, in Bayesian statistics, this corresponds to the 
marginal likelihood and can be used for model selection.

\medskip 
As a side-product of PT, we automatically obtain an approximation to the natural 
logarithm of the normalization constant. This is done automatically using the 
stepping stone estimator \cite{xie2011improving}.
The estimate can be accessed using
\begin{lstlisting}[language=Julia]
stepping_stone(pt)
\end{lstlisting}
In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$ 
using a computer algebra system. Note that this is very close to the output provided in 
\cref{fig:standard_out}.

\subsubsection{Online statistics}
\label{sec:online_stats}
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

\subsubsection{Off-memory processing}
When either the dimensionality of the model or the number of samples is large,
the obtained samples may not fit in memory. 
In some cases it may be necessary to store samples to disk if our statistics of 
interest cannot be calculated online and with constant-memory
(see \cref{sec:online_stats}).
We show here how to save samples to disk when Pigeons.jl is run on a single 
machine. A similar interface can be used over MPI.

\medskip 
First, we make sure that we set \texttt{checkpoint = true}, which saves a 
snapshot at the end of each round in the directory \texttt{results/all/<unique directory>}
and is symlinked to \texttt{results/latest}.
Second, we make sure that we use the \texttt{disk} recorder 
by setting \texttt{record = [disk]}, along with possibly any other desired recorders.
Accessing the samples from disk can then be achieved in a simple way using the Pigeons.jl 
function \texttt{process\_sample()}.

\subsubsection{PT diagnostics}
\label{sec:PT_diagnostics}
We describe how to produce some key parallel tempering diagnostics from 
\cite{syed2021nrpt}.

\medskip 
The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 
used to approximately inform the appropriate number of chains. 
Based on \cite{syed2021nrpt}, stable PT performance should be achieved when  
the number of chains is set to roughly $2\Lambda$. This can be achieved by 
modifying the \texttt{n\_chains} argument in the call to \texttt{pigeons()}. 
The global communication barrier is shown at each round and can also be accessed 
with 
\begin{lstlisting}[language=Julia]
Pigeons.global_barrier(pt)
\end{lstlisting}

The number of restarts per round can be accessed with 
\begin{lstlisting}[language=Julia]
n_tempered_restarts(pt)
\end{lstlisting}
These quantities are also displayed in \cref{fig:standard_out}. 
Many other useful PT diagnostic statistics and plots can be obtained, as described 
in our full documentation.

\subsection{Parallel and distributed PT}
One of the main benefits of Pigeons.jl is that it allows users to easily parallelize 
and/or distribute their PT sampling efforts. We explain how to run MPI locally on 
one machine and also how to use MPI when a cluster is available.

\subsubsection{Running MPI locally}
To run MPI locally on one machine using four MPI processes and one thread per process,
use
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model), 
    on = ChildProcess(
            n_local_mpi_processes = 4,
            n_threads = 1))
\end{lstlisting}

\subsubsection{Running MPI on a cluster}
Often, MPI is available via a cluster scheduling system. To run MPI over 
several machines:
\begin{enumerate}
    \item In the cluster login node, follow the Pigeons.jl installation instructions
    in our online documentation. 
    \item Start Julia in the login node, and perform a one-time setup by 
    calling \texttt{Pigeons.setup\_mpi()}.
    \item In the Julia REPL running in the login node, run:
\end{enumerate}
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model), 
    n_chains = 1_000,
    on = MPI(n_mpi_processes = 1_000, 
             n_threads = 1))
\end{lstlisting}
The code above will start a distributed PT algorithm with 1,000 chains on 1,000 
MPI processes each using one thread.
Note that for the above code chunks involving \texttt{ChildProcess()} and 
\texttt{MPI()} to work, it may be necessary to specify dependencies in their 
function calls.

\subsection{Additional options}
\label{sec:additional_options}
In the preceding example we only specified the target distribution and let 
Pigeons.jl decide on default values for most other settings of the inference engine. 
There are various settings we can change, including: 
the random seed (\texttt{seed}), the number of PT chains (\texttt{n\_chains}), 
the number of PT tuning rounds/samples (\texttt{n\_rounds}), 
and a variational reference distribution family (\texttt{variational}), among other settings.
For instance, we can run 
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model),
    n_rounds = 10, 
    n_chains = 10,
    variational = GaussianReference(),
    seed = 2
)
\end{lstlisting}
which runs PT with the same Turing model target as before and explicitly states 
that we should use 10 PT tuning rounds with 10 chains (described below). 
In the above code chunk we also specify that we would like to use a 
Gaussian variational reference distribution.
That is, the reference distribution is chosen from a multivariate Gaussian family 
that lies as close as possible to the target distribution in order to improve 
the efficiency of PT. We refer readers to \cite{surjanovic2022vpt} for more details.
When only continuous parameters are of interest, we encourage users to consider 
using \texttt{variational = GaussianReference()} and setting \texttt{n\_chains\_variational = 10}, 
for example, as the number of restarts may substantially increase with these settings.

### 2023-08-18 src/introduction.tex
Upload all files

\section{Introduction}
In many scientific application domains, the ability to obtain samples from a 
distribution is crucial. 
For instance, sampling methods have been used to discover magnetic polarization 
in the black hole of galaxy M87 \cite{akiyama2021seven}
and to image the Sagittarius A* black hole \cite{akiyama2022first}.
They have also been used to 
model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
and to identify gerrymandering in Georgia's 2021 congressional districting plan 
\cite{zhao2022mathematically}.
Similarly, evaluating high-dimensional integrals or sums over complicated 
combinatorial spaces are related tasks that can also be solved with sampling 
via Markov chain Monte Carlo (MCMC) methods. 
However, such calculations can often be bottlenecks in the scientific process, with 
simulations that can last days or even weeks to finish.

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0 is currently compatible with Julia 1.8+.}
enables users to sample efficiently from high-dimensional and complex distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple API allows users to perform such computations 
single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
machines. Further, it comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density
\footnote{This density may also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

\medskip 
The output of Pigeons.jl can be used for two main tasks:
\begin{enumerate}
    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
    For example, the choice $f(x) = x$ computes the mean and 
    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
    where $I[\cdot]$ denotes the indicator function.

    \item Approximating the value of the normalization constant $Z$. For 
    example, in Bayesian statistics, this corresponds to the 
    marginal likelihood, which can be used for model selection. 
\end{enumerate}
Its implementation shines compared to traditional sampling packages in the 
following scenarios:
\begin{itemize}
    \item When the target density $\pi$ is challenging due to a complex structure 
    (e.g., high-dimensional, multi-modal, etc.).

    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
    the normalization constant $Z$. 
    Many existing tools focus on the former and struggle or fail to do the latter.

    \item When the target distribution $\pi$ is defined over a non-standard space, 
    e.g. a combinatorial object such as a phylogenetic tree.  
\end{itemize}

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is completely invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

\medskip 
The general Julia developer will be interested in: 
\begin{itemize}
    \item The main causes of a violation of strong parallelism invariance that we have identified 
    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
    them in Pigeons.jl.

    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
    package that was developed by our team 
    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
\end{itemize}

### 2023-08-18 scripts/MH_walk.jl

@@ -1,0 +1,32 @@
+using LinearAlgebra 
+using Distributions
+
+"""
+Compute the acceptance ratio for a MH random walk. 
+Assumes that the proposal is standard normal and that there are exactly two dimensions!
+"""
+function acceptance_ratio(x_0, x_1, target_pdf)
+    return min(1.0, target_pdf(x_1) * pdf(MvNormal([x_1[1], x_1[2]], I), x_0) / 
+           (target_pdf(x_0) * pdf(MvNormal([x_0[1], x_0[2]], I), x_1)))
+end
+
+"""
+Very primitive implementation of a Metropolis-Hastings random walk 
+with a standard normal proposal with exactly two dimensions. (Used only for plotting purposes.)
+"""
+function MH_walk(rng, x_0, n, target_pdf)
+    xs = Vector{Vector{Float64}}(undef, n+1)
+    xs[1] = x_0
+    for i in 2:(n+1)
+        x_prev = xs[i-1]
+        x_prop = x_prev + rand(rng, MvNormal([0.0, 0.0], I))
+        α = acceptance_ratio(x_prev, x_prop, target_pdf)
+        U = rand(rng, Uniform(0.0, 1.0))
+        if U ≤ α
+            xs[i] = x_prop 
+        else 
+            xs[i] = copy(x_prev)
+        end
+    end
+    return xs
+end

### 2023-08-18 scripts/bimodal.jl

@@ -1,0 +1,47 @@
+"""
+    Generate a figure for a bimodal distribution.
+"""
+    
+using Plots
+using Distributions
+using Random
+    
+include("distributions.jl")
+include("MH_walk.jl")
+
+function main()
+    a = 2.5
+    res = 1_000
+    n_walk = 1_000
+
+    xs = range((-a-4.0), (a+4.0), length = res)
+    ys = copy(xs)
+    z = zeros(res, res)
+    for i in 1:res
+        x = xs[i]
+        for j in 1:res
+            y = ys[j]
+            z[i, j] = target_pdf(x, y, a)
+        end
+    end
+    p = heatmap(
+        xs, ys, z,
+        colorbar = :none, border = :none, margin = (0.0,:mm), axis = nothing,
+        aspect_ratio = :equal, size = (500, 500)
+    )
+
+    function target_pdf_2(x) 
+        target_pdf(x[1], x[2], a)
+    end
+    rng = MersenneTwister(2)
+    walk = MH_walk(rng, [-a, -a], n_walk, target_pdf_2)
+    plot!(
+        p, [walk[i][1] for i in eachindex(walk)], 
+        [walk[i][2] for i in eachindex(walk)], color = 1,
+        linewidth = 2, legend_position = :none, alpha = 0.5
+    )
+    savefig(p, "img/bimodal.pdf")
+end
+
+
+main()

### 2023-08-18 scripts/coinflip_posterior.jl

@@ -1,0 +1,40 @@
+using Distributions 
+using Random 
+using Plots 
+
+function coinflip_posterior(n, n_success) 
+    n_fail = n - n_success 
+    x = range(0.0, 1.0, length = 1_000)
+    y = pdf.(Beta(n_success + 1, n_fail + 1), x)
+    p = Plots.plot(
+        x, y, ylims = (0.0, 10.0), xlims = (0.0, 1.0), linewidth = 2, 
+        xticks = 0.0:0.1:1.0, legend = false, xlabel = "θ", ylabel = "π(θ|x)"
+    )
+    return p 
+end 
+
+
+function main() 
+    # settings 
+    n_flips = 100 
+    true_prob = 0.75
+    rng = MersenneTwister(1)
+    plot_indices = vcat(0:10, n_flips) # which plots to save
+
+    y = []
+    for n in 0:n_flips 
+        if n == 0 
+            p = coinflip_posterior(0, 0) # plot uniform distribution 
+        else
+            y = vcat(y, rand(rng, Bernoulli(true_prob)))
+            n_success = sum(y)
+            p = coinflip_posterior(n, n_success)
+        end
+        if n in plot_indices
+            display(p)
+            savefig(p, "results/coinflip/coinflip_posterior_" * string(n) * ".png")
+        end
+    end 
+end 
+
+main()

### 2023-08-18 scripts/distributions.jl

@@ -1,0 +1,9 @@
+function target_pdf(x, y, a)
+    return 0.5 * (pdf(Normal(-a, 1.0), x) * pdf(Normal(-a, 1.0), y)) +
+           0.5 * (pdf(Normal(a, 1.0), x) * pdf(Normal(a, 1.0), y))
+end
+
+function reference_pdf(x, y, a)
+    return pdf(Normal(0.0, sqrt(a^2+1)), x) * pdf(Normal(0.0, sqrt(a^2+1)), y)
+end
+

### 2023-08-18 scripts/examples.jl

@@ -1,0 +1,99 @@
+using Turing
+@model function coinflip(n, y)
+    p1 ~ Uniform(0.0, 1.0)
+    p2 ~ Uniform(0.0, 1.0)
+    y ~ Binomial(n, p1 * p2)
+    return y
+end
+model = coinflip(1e5, 5e4)
+
+using Pigeons
+pt = pigeons(
+    target = TuringLogPotential(model), 
+    record = [traces, online, round_trip, Pigeons.timing_extrema, Pigeons.allocation_extrema]) 
+
+# pt_large = pigeons(
+#     target = TuringLogPotential(model), 
+#     record = [traces, online, round_trip, Pigeons.timing_extrema, Pigeons.allocation_extrema],
+#     n_rounds = 17) # not in JCON
+
+single_chain = pigeons(
+    target = TuringLogPotential(model), 
+    n_chains = 1, 
+    record = [traces])
+
+using MCMCChains, StatsPlots, PlotlyJS
+# plotlyjs()
+# samples = Chains(sample_array(pt), variable_names(pt))
+# my_plot = StatsPlots.plot(samples)
+# display(my_plot)
+
+# using PairPlots, CairoMakie
+# my_plot = PairPlots.pairplot(samples) 
+# display(my_plot)
+
+# Not a JCON code chunk ----------
+# Increase the number of samples to make the estimated posterior smoother
+# samples_large = Chains(sample_array(pt_large), variable_names(pt_large))
+# my_plot_large = PairPlots.pairplot(samples_large)
+# # CairoMakie.save("img/coinflip_posterior.pdf", my_plot_large) 
+# my_traceplot_large = StatsPlots.plot(samples_large)
+# display(my_traceplot_large)
+
+sa = sample_array(pt)[:, 1, :] # not a JCON code chunk ----------
+sa = reshape(sa, size(sa, 1), 1, 1)
+samples = Chains(sa, [variable_names(pt)[1]])
+# my_plot = StatsPlots.plot(samples)
+# display(my_plot)
+# turn off plotlyjs() for this plot!
+my_plot_v2 = Plots.plot(
+    1:(size(sa)[1]), sa[:,1,1],
+    xlabel = "Iteration", ylabel = "Sample value", legend = false
+)
+display(my_plot_v2)
+StatsPlots.savefig(my_plot_v2, "img/trace_density_plots_pt.pdf")
+
[…]

### 2023-08-18 scripts/heatmap_path.jl

@@ -1,0 +1,35 @@
+"""
+    Generate a path of heatmaps from a unimodal reference to a bimodal target.
+"""
+
+using Plots
+using Distributions
+
+include("distributions.jl")
+
+function main()
+    a = 2.5
+    res = 1_000
+    βs = [0.0, 0.05, 0.15, 0.25, 0.5, 1.0]
+
+    xs = range((-a-4.0), (a+4.0), length = res)
+    ys = copy(xs)
+    z = zeros(res, res)
+    k = 1
+    for β in βs
+        for i in 1:res
+            x = xs[i]
+            for j in 1:res
+                y = ys[j]
+                z[i, j] = reference_pdf(x, y, a)^(1-β) * target_pdf(x, y, a)^β
+            end
+        end
+        p = heatmap(xs, ys, z,
+                colorbar = :none, border = :none, margin = (0.0,:mm), axis = nothing,
+                aspect_ratio = :equal, size = (500, 500))
+        savefig(p, "img/heatmap_path_$k.pdf")
+        k += 1
+    end
+end
+
+main()

### 2023-08-18 scripts/index_process.jl

@@ -1,0 +1,67 @@
+using Plots
+using Pigeons
+
+function first_process_index(d::Dict{Int64, Vector{Int64}}) 
+    for i in eachindex(d)
+        if d[i][1] == 1 
+            return i 
+        end 
+    end
+    error()
+end
+
+
+function main()
+    # settings 
+    n_chains = 4
+    n_rounds = 3 # 2^n_rounds iterations
+    linewidth = 3
+    alpha = 0.2
+    acceptance = 1.0
+    seed = 1
+    n_iters = 5
+    padding = 0.15
+
+    # perform pigeons simulation 
+    pt = pigeons(
+        target            = Pigeons.TestSwapper(acceptance), 
+        record = [Pigeons.round_trip, Pigeons.index_process], 
+        n_chains          = n_chains, 
+        n_rounds          = n_rounds, 
+        seed              = seed
+    )
+    index_process = pt.reduced_recorders.index_process
+    first_process = first_process_index(index_process)
+
+    # make plots 
+    for n_iters_sub in 2:n_iters
+        p = Plots.plot(
+            1:n_iters_sub,
+            index_process[first_process][1:n_iters_sub], lc = 1, 
+            xlims = (1-padding, n_iters+padding), ylims = (1-padding, n_chains+padding), 
+            legend = false, linewidth = linewidth + 0.5, xticks = 1:1:n_iters, 
+            yticks = (1:1:n_chains, 1:n_chains), xlabel = "Iteration", ylabel = "Chain number")
+        if n_iters_sub != n_iters
+            Plots.plot!(
+                p, n_iters_sub:n_iters, index_process[first_process][n_iters_sub:n_iters], lc = 1, 
+                linewidth = linewidth, alpha = alpha^2) 
+        end
+        for i in eachindex(index_process)
+            if i != first_process
+                Plots.plot!(
+                    p, 1:n_iters_sub, index_process[i][1:n_iters_sub], lc = index_process[i][1], 
+                    linewidth = linewidth, alpha = alpha)
+                if n_iters_sub != n_iters 
+                    Plots.plot!(
+                        p, n_iters_sub:n_iters, index_process[i][n_iters_sub:n_iters], 
+                        lc = index_process[i][1], 
+                        linewidth = linewidth, alpha = alpha^2)
+                end
+            end 
+        end 
+        display(p)
+        savefig(p, "results/index_process_" * string(n_iters_sub - 1) * ".pdf")
+    end
+end 
+
+main()

### 2023-08-18 scripts/unimodal.jl

@@ -1,0 +1,52 @@
+"""
+    Generate MH-walk figures for a unimodal distribution.
+"""
+    
+using Plots
+using Distributions
+using Random
+    
+include("MH_walk.jl")
+
+target_pdf(x, y) = pdf(Normal(0.0, 1.0), x) * pdf(Normal(0.0, 1.0), y)
+
+function main()
+    res = 1_000
+    n_walk = 200
+
+    xs = range(-4.0, 4.0, length = res)
+    ys = copy(xs)
+    z = zeros(res, res)
+    for i in 1:res
+        x = xs[i]
+        for j in 1:res
+            y = ys[j]
+            z[i, j] = target_pdf(x, y)
+        end
+    end
+    
+
+    function target_pdf_2(x) 
+        target_pdf(x[1], x[2])
+    end
+
+    anim = @animate for n in 1:n_walk
+        rng = MersenneTwister(1)
+        walk = MH_walk(rng, [0.0, 0.0], n, target_pdf_2)
+        p = heatmap(
+            xs, ys, z,
+            colorbar = :none, border = :none, margin = (0.0,:mm), axis = nothing,
+            aspect_ratio = :equal, size = (500, 500)
+        )
+        plot!(
+            p, [walk[i][1] for i in eachindex(walk)], 
+            [walk[i][2] for i in eachindex(walk)], color = 1,
+            linewidth = 2, legend_position = :none, alpha = 0.5
+        )
+        # savefig(p, "results/unimodal/unimodal_" * string(n) * ".pdf")
+    end
+    gif(anim, "results/unimodal/unimodal.gif", fps = 15)
+end
+
+
+main()

### 2023-08-18 src/SPI.tex
Minor changes before submission

@@
 floating point arithmetic, we ensure that the order in which operations are performed 
-are exactly the same, irrespective of the number of processes/machines and threads. 
+is exactly the same, irrespective of the number of processes/machines and threads. 
 This is achieved by making sure that every value to be added---if addition is our 
@@
 of execution in Julia.
-We note that the \texttt{@threads} macro in Julia creates exactly \texttt{nthreads()}
+We note that the \texttt{@threads} macro in Julia creates \texttt{nthreads()}
 tasks and thus \texttt{nthreads()} pseudorandom number generators.

### 2023-08-18 src/examples.tex
Minor changes before submission

@@
     \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
-    \item Comrade.jl models for black hole imaging (\texttt{ComradeLogPotential})
+    \item Comrade.jl\footnote{\url{https://github.com/ptiede/Comrade.jl}} 
+      models for black hole imaging (\texttt{ComradeLogPotential})
     \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
@@
 diagnostic output by specifying 
-\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.jl.allocation\_extrema]}.
+\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
 It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
@@
     \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
-    log of the normalization constant. See \cref{sec:PT_diagnostics} for more information.
+    logarithm of the ratio of normalization constants between the target and the reference. 
+    In many cases, $Z_0 = 1$.
 

### 2023-08-18 src/introduction.tex
Add scripts for threads experiments

@@
 \url{https://github.com/Julia-Tempering/Pigeons.jl}. 
-Pigeons.jl v0.2.0 is currently compatible with Julia 1.8+.}
+Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
 enables users to sample efficiently from high-dimensional and complex distributions 

### 2023-08-18 scripts/SplittableRandoms_example.jl

@@ -1,0 +1,15 @@
+using Random
+using SplittableRandoms: SplittableRandom, split
+import Base.Threads.@threads
+
+println("Num. of threads: $(Threads.nthreads())")
+
+const n_iters = 10000
+const master_rng = SplittableRandom(1)
+result = zeros(n_iters)
+rngs = [split(master_rng) for _ in 1:n_iters]
+Random.seed!(1)
+@threads for i in 1:n_iters
+    result[i] = rand(rngs[i])
+end
+println("Result: $(last(result))")

### 2023-08-18 scripts/examples.jl

@@ -23,14 +23,14 @@ single_chain = pigeons(
     record = [traces])
 
 using MCMCChains, StatsPlots, PlotlyJS
-# plotlyjs()
-# samples = Chains(sample_array(pt), variable_names(pt))
-# my_plot = StatsPlots.plot(samples)
-# display(my_plot)
+plotlyjs()
+samples = Chains(sample_array(pt), variable_names(pt))
+my_plot = StatsPlots.plot(samples)
+display(my_plot)
 
-# using PairPlots, CairoMakie
-# my_plot = PairPlots.pairplot(samples) 
-# display(my_plot)
+using PairPlots, CairoMakie
+my_plot = PairPlots.pairplot(samples) 
+display(my_plot)
 
 # Not a JCON code chunk ----------
 # Increase the number of samples to make the estimated posterior smoother
@@ -40,39 +40,39 @@ using MCMCChains, StatsPlots, PlotlyJS
 # my_traceplot_large = StatsPlots.plot(samples_large)
 # display(my_traceplot_large)
 
-sa = sample_array(pt)[:, 1, :] # not a JCON code chunk ----------
-sa = reshape(sa, size(sa, 1), 1, 1)
-samples = Chains(sa, [variable_names(pt)[1]])
+# sa = sample_array(pt)[:, 1, :] # not a JCON code chunk ----------
+# sa = reshape(sa, size(sa, 1), 1, 1)
+# samples = Chains(sa, [variable_names(pt)[1]])
 # my_plot = StatsPlots.plot(samples)
 # display(my_plot)
 # turn off plotlyjs() for this plot!
-my_plot_v2 = Plots.plot(
-    1:(size(sa)[1]), sa[:,1,1],
-    xlabel = "Iteration", ylabel = "Sample value", legend = false
-)
-display(my_plot_v2)
-StatsPlots.savefig(my_plot_v2, "img/trace_density_plots_pt.pdf")
+# my_plot_v2 = Plots.plot(
+    # 1:(size(sa)[1]), sa[:,1,1],
+    # xlabel = "Iteration", ylabel = "Sample value", legend = false
+# )
+# display(my_plot_v2)
+# StatsPlots.savefig(my_plot_v2, "img/trace_density_plots_pt.pdf")
 
-sa = sample_array(single_chain)[:, 1, :] # not a JCON code chunk ----------
-sa = reshape(sa, size(sa, 1), 1, 1)
[…]

### 2023-08-18 scripts/threads.jl

@@ -1,0 +1,12 @@
+using Random
+import Base.Threads.@threads
+
+println("Num. of threads: $(Threads.nthreads())")
+
+const n_iters = 10000
+result = zeros(n_iters)
+Random.seed!(1)
+@threads for i in 1:n_iters
+    result[i] = rand()
+end
+println("Result: $(last(result))")

### 2023-10-30 paper.tex
Remove inputs and place in main tex

@@
 
-\input{src/introduction.tex}
-\input{src/examples.tex}
-\input{src/PT.tex}
-\input{src/SPI.tex}
-\input{src/related.tex}
-% \input{src/benchmarks.tex}
-\input{src/conclusion.tex}
+\section{Introduction}
+In many scientific application domains, the ability to obtain samples from a 
+distribution is crucial. 
+For instance, sampling methods have been used to discover magnetic polarization 
+in the black hole of galaxy M87 \cite{akiyama2021seven}
+and to image the Sagittarius A* black hole \cite{akiyama2022first}.
+They have also been used to 
+model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
+infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
+and to identify gerrymandering in Georgia's 2021 congressional districting plan 
+\cite{zhao2022mathematically}.
+Similarly, evaluating high-dimensional integrals or sums over complicated 
+combinatorial spaces are related tasks that can also be solved with sampling 
+via Markov chain Monte Carlo (MCMC) methods. 
+However, such calculations can often be bottlenecks in the scientific process, with 
+simulations that can last days or even weeks to finish.
+
+\medskip
+Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
+\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
+Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
+enables users to sample efficiently from high-dimensional and complex distributions 
+and solve integration problems by 
+implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
+that leverage distributed computation. Its simple API allows users to perform such computations 
+single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
+machines. Further, it comes with guarantees on strong parallelism invariance wherein 
+the output for a given seed is \emph{identical} irrespective 
+of the number of threads or processes. 
+Such a level of reproducibility is rare in distributed software but of 
+great use for the purposes of debugging in the context of sampling algorithms, 
+which produce stochastic output.
+Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
+\begin{enumerate}
+    \item one machine running on one thread;
+    \item one machine running on several threads;
+    \item several machines running, each using one thread, and
+    \item several machines running, each using several threads.
+\end{enumerate}
+
+\subsection{Problem formulation}
+We describe the class of problems that can be approached using Pigeons.jl.
+Let $\pi(x)$ denote a probability density
+\footnote{This density may also be a probability mass function 
+(discrete variables). We also allow for a combination of discrete/continuous variables.} 
+called the \emph{target}. 
+In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
+known only up to a normalizing constant, 
+\[
+\label{eq:normalizing_constant}
+  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
+\]  
+where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
+Pigeons.jl takes as input the function $\gamma$.
+
+\medskip 
+The output of Pigeons.jl can be used for two main tasks:
+\begin{enumerate}
+    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
+    For example, the choice $f(x) = x$ computes the mean and 
+    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
+    where $I[\cdot]$ denotes the indicator function.
+
+    \item Approximating the value of the normalization constant $Z$. For 
+    example, in Bayesian statistics, this corresponds to the 
+    marginal likelihood, which can be used for model selection. 
+\end{enumerate}
+Its implementation shines compared to traditional sampling packages in the 
+following scenarios:
+\begin{itemize}
+    \item When the target density $\pi$ is challenging due to a complex structure 
+    (e.g., high-dimensional, multi-modal, etc.).
+
+    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
+    the normalization constant $Z$. 
+    Many existing tools focus on the former and struggle or fail to do the latter.
+
+    \item When the target distribution $\pi$ is defined over a non-standard space, 
+    e.g. a combinatorial object such as a phylogenetic tree.  
+\end{itemize}
+
+\subsection{What is of interest to the general Julia developer?}
+Ensuring code correctness at the intersection of randomized, parallel, and 
+distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
+based on a principle that we refer to as \textit{strong parallelism invariance}
+(SPI).
+Namely, the output of Pigeons.jl is completely invariant to the number of machines 
+and/or threads.
+Without explicitly keeping SPI in mind during software construction, 
+the (random) output of the algorithm is only guaranteed to have the same distribution.
+This is a much weaker guarantee that, in particular, makes debugging difficult.
+However, with our notion of SPI we make debugging and 
+software validation considerably easier. This is because the developer can 
+first focus on the fully serial randomized algorithm, and then use it as an 
+easy-to-compare gold-standard reference for parallel/distributed implementations. 
+This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
+testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
+incurs additional false negatives due to statistical error.
+
+\medskip 
+The general Julia developer will be interested in: 
+\begin{itemize}
+    \item The main causes of a violation of strong parallelism invariance that we have identified 
+    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
+    them in Pigeons.jl.
+
+    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
+    package that was developed by our team 
+    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
+\end{itemize}
+
+\section{Examples}
+In this section we present a set of minimal examples 
+that demonstrate how to use Pigeons.jl for sampling. 
+For further reading, we also direct readers to our growing list of examples, which can be found at 
+\url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
+We begin by installing the latest official release of Pigeons.jl: 
+\begin{lstlisting}[language = Julia]
+using Pkg; Pkg.add("Pigeons")
+\end{lstlisting}
+
+\subsection{Targets}
+To use Pigeons.jl, we must specify a target distribution, given by $\gamma$ 
+in \cref{eq:normalizing_constant}.
+Numerous possible types of target distributions are supported, including 
+custom probability densities (specified up to a normalizing constant) written in Julia.
+We also allow to interface with models written in common probabilistic programming 
+languages, including:
+\begin{itemize}
+    \item Turing.jl \cite{ge2018turing} models (\texttt{TuringLogPotential})
+    \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
+    \item Comrade.jl\footnote{\url{https://github.com/ptiede/Comrade.jl}} 
+      models for black hole imaging (\texttt{ComradeLogPotential})
+    \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
+    (e.g. Blang \cite{bouchard2022blang} code for phylogenetic inference over combinatorial spaces) 
+\end{itemize}
+Additional targets are currently being accommodated and will be 
+introduced to Pigeons.jl in the near future.
+
+\medskip 
+In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
+applied to a non-identifiable ``coinflip'' data set, modified from the example
+at \url{https://turing.ml/v0.22/docs/using-turing/quick-start}. 
+The Bayesian model can be formulated as 
+\[
+  \label{eq:coinflip}
+  p_1, p_2 &\stackrel{iid}{\sim} U(0, 1) \\    
+  Y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
+\]
+The random variable $Y$ is the number of heads observed on $n$ coin flips
+where the probability of heads is $p_1 p_2$.  
+This model is non-identifiable, meaning that it is not possible to distinguish 
+the effects of the two different parameters $p_1$ and $p_2$. As a consequence, 
+the target distribution exhibits a complicated structure, as displayed in 
+\cref{fig:coinflip_posterior}.
+The density of interest corresponding to this model is 
+\[
+  \pi(p_1, p_2) &= \gamma(p_1, p_2)/Z,
+\] 
+where 
+\[ 
+  \gamma(p_1, p_2) &= 
+    {n \choose y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
+    \label{eq:coinflip_density} \\
+  Z &= \int_0^1 \int_0^1 \gamma(p_1, p_2) \, \dee p_1 \, \dee p_2. \label{eq:coinflip_normalization}
+\]
+The distribution $\pi$ is also known as the \emph{posterior distribution} in 
+Bayesian statistics.
+
+\begin{figure}[t]
+    \centering 
+    \includegraphics[width=0.5\textwidth]{img/coinflip_posterior.pdf}
+    \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
+    with $n=100,000$ coin flips and $y=50,000$ observed heads, 
+    estimated using $2^{17}$ samples from Pigeons.jl. 
+    We present the pairwise plot for $p_1$ and $p_2$, as well as the estimated 
+    densities of the marginal of the posterior for each of the two parameters.
+    Note that because the model is non-identifiable, as we collect more data the 
+    posterior distribution concentrates around the curve $p_1 p_2 = 0.5$, instead 
+    of a single point, assuming that the true probability of observing heads 
+    is 0.5.}
+    \label{fig:coinflip_posterior}
+\end{figure}
+
+\medskip 
+Suppose that we perform $n=100,000$ coin tosses and observe 
+$y=50,000$ heads.
+We would like to obtain samples from our posterior, $\pi$, having collected this data.
+We begin by installing Turing
+\begin{lstlisting}[language = Julia]
+Pkg.add("Turing")
+\end{lstlisting}
+and then defining our Turing model and storing it in the variable \texttt{model}:
+\begin{lstlisting}[language = Julia]
+using Turing
+@model function coinflip(n, y)
+    p1 ~ Uniform(0.0, 1.0)
+    p2 ~ Uniform(0.0, 1.0)
+    y ~ Binomial(n, p1 * p2)
+    return y
+end
+model = coinflip(100000, 50000)
+\end{lstlisting}
+
+From here, it is straightforward to sample from the density given by 
+\cref{eq:coinflip_density} up to a normalizing constant.
+We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
+state-of-the-art sampling algorithm, to sample from the target distribution.
+PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
+to select these parameters effectively, which Pigeons.jl implements under the hood.
+We also specify that we would like to store 
+the obtained samples in memory to be able to produce trace-plots, as well 
+as some basic online summary statistics of the target distribution and useful 
+diagnostic output by specifying 
+\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
+It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
+will be selected for the user.
+The code below runs Pigeons.jl on one machine with one thread.
+We use the default values for most settings, however 
+we explain later how one can obtain improved performance by setting 
+arguments more carefully (see \cref{sec:additional_options}).
+\begin{lstlisting}[language = Julia]
+using Pigeons
+pt = pigeons(
+    target = TuringLogPotential(model), 
+    record = [
+        traces, online, round_trip, 
+        Pigeons.timing_extrema, 
+        Pigeons.allocation_extrema])
+\end{lstlisting}
+Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
+we pass the model as an argument to \texttt{TuringLogPotential()}.
+Once we have stored the PT output in the variable \texttt{pt} we can 
+access the results, as described in the following section. 
+The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
+and explained in the next section. 
+For purposes of comparison, we also run a traditional 
+(single-chain Markov chain Monte Carlo) method.
+
+\begin{figure*}[t]
+    \centering
+    \begin{lstlisting}
+          --------------------------------------------------------------------------------------
+            #scans    restarts      Λ        time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
+          ---------- ---------- ---------- ---------- ---------- ---------- ---------- --------- 
+                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
+                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
+                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
+                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
+                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
+                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
+                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
+                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
+                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
+           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
+          --------------------------------------------------------------------------------------
+    \end{lstlisting}
+    \caption{Standard output provided by Pigeons.jl. 
+    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
+    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
+    such as the number of allocations per round, time (in seconds), and estimates 
+    of the log of the normalization constants. 
+    The output is described in greater detail in \cref{sec:PT_diagnostics}
+    and has been modified to exclude columns that are not described in the paper.}
+    \label{fig:standard_out}
+\end{figure*}
+
+\subsubsection{Other targets}
+As mentioned previously, it is also possible to specify targets with custom 
+probability densities, as well as Stan and Turing models. 
+Additionally, suppose we have some code implementing vanilla MCMC, written 
+in an arbitrary ``foreign'' language such as C++, Python, R, Java, etc. 
+Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
+The only requirement on the foreign language is that it supports 
+reading the standard input and writing to the standard output, 
+as described in our online documentation.
+
+\subsection{Outputs}
+Pigeons.jl provides many useful types of output, such as: plots of samples from 
+the distribution, estimates of normalization constants, summary statistics of 
+the target distribution, and various other diagnostics. 
+We describe several examples of possible output below.
+
+\subsubsection{Standard output}
+An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
+Each row of the table in the output indicates a new tuning round in parallel tempering, 
+with the \texttt{\#scans} column indicating the number of scans/samples in 
+that tuning round. 
+During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
+tuning parameters. Other outputs include: 
+\begin{itemize}
+    \item \texttt{restarts}: a higher number is better. 
+    Informally, PT performs well at sampling from high-dimensional and/or multi-modal 
+    distributions by pushing samples from an easy-to-sample distribution (the reference) 
+    to the more difficult target distribution. 
+    A tempered restart happens when a sample from the reference successfully 
+    percolates to the target. (See the subsequent sections for a more detailed description 
+    of parallel tempering.) 
+    When the reference supports i.i.d.~sampling, tempered restarts 
+    can enable large jumps in the state space.
+
+    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
+    which measures the inherent difficulty of the sampling problem.
+    A rule of thumb to configure the number of PT chains is also 
+    given by \cite{syed2021nrpt}, where they suggest that stable performance 
+    should be achieved when the number of chains is set to roughly 2$\Lambda$.
+    See \cref{sec:PT_diagnostics} for more information.
+
+    \item \texttt{time} and \texttt{allc}: the time (in seconds) and number of 
+    allocations (in bytes) used in each round.
+
+    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
+    logarithm of the ratio of normalization constants between the target and the reference. 
+    In many cases, $Z_0 = 1$.
+
+    \item \texttt{min($\alpha$)} and \texttt{mean($\alpha$)}: minimum and average swap 
+    acceptance rates during the communication phase across the PT chains.
+    See \cref{sec:PT} for a description of PT communication.
+\end{itemize}
+
+\subsubsection{Plots}
+It is straightforward to obtain plots of samples from the target distribution, 
+such as trace-plots, pairwise plots, and density plots of the marginals.
+
+\medskip 
+To obtain posterior densities and trace-plots, we first 
+make sure that we have the third-party 
+MCMCChains.jl\footnote{\url{https://github.com/TuringLang/MCMCChains.jl}}, 
+StatsPlots.jl\footnote{\url{https://github.com/JuliaPlots/StatsPlots.jl}}, and 
+PlotlyJS.jl\footnote{\url{https://github.com/JuliaPlots/PlotlyJS.jl}} 
+packages installed via
+\begin{lstlisting}[language=Julia]
+Pkg.add("MCMCChains", "StatsPlots", "PlotlyJS")
+\end{lstlisting}
+
+With the \texttt{pt} output object from before for our non-identifiable coinflip model, 
+we can run the following:
+\begin{lstlisting}[language=Julia]
+using MCMCChains, StatsPlots, PlotlyJS
+plotlyjs()
+samples = Chains(
+    sample_array(pt), variable_names(pt))
+my_plot = StatsPlots.plot(samples)
+display(my_plot)
+\end{lstlisting}
+The output of the above code chunk is an interactive plot that can be zoomed in or out
+and exported as an HTML webpage. 
+A modified static version of the output for the first parameter, $p_1$, 
+is displayed in the top panel of \cref{fig:trace_density_plots},
+along with a comparison to the output from a single-chain algorithm in the bottom 
+panel of the same figure.
+
+\begin{figure*}[t]
+    \centering
+    \begin{subfigure}{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/trace_density_plots_pt.pdf}
+    \end{subfigure}
+    \begin{subfigure}{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/trace_density_plots_single_chain.pdf}
+    \end{subfigure}
+    \caption{
+        Trace-plots for the first parameter, $p_1$, in the 
+        non-identifiable coinflip Turing model.
+        \textbf{Left:} Samples from Pigeons.jl using PT with 10 chains. Note that the 
+        trace-plot indicates fast mixing/exploration across the state space.
+        \textbf{Right:} Single-chain Markov chain Monte Carlo. Note that the 
+        trace-plot explores the state space much more slowly when we do not use PT.
+    }
+    \label{fig:trace_density_plots}
+\end{figure*}
+
+\medskip 
+To obtain pair plots, we add the 
+PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
+and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
+packages:
+\begin{lstlisting}[language=Julia]
+Pkg.add("PairPlots", "CairoMakie")
+\end{lstlisting}
+and then run
+\begin{lstlisting}[language=Julia]
+using PairPlots, CairoMakie
+my_plot = PairPlots.pairplot(samples) 
+display(my_plot)
+\end{lstlisting}
+The output of the above code chunk with an increased number of samples 
+is displayed in \cref{fig:coinflip_posterior}.
+\footnote{The code chunk above only works with Julia 1.9.}
+
+\subsubsection{Estimate of normalization constant}
+The (typically unknown) constant $Z$ in \cref{eq:normalizing_constant} is 
+referred to as the \emph{normalization constant}. 
+In many applications, it is useful to approximate this constant. 
+For example, in Bayesian statistics, this corresponds to the 
+marginal likelihood and can be used for model selection.
+
+\medskip 
+As a side-product of PT, we automatically obtain an approximation to the natural 
+logarithm of the normalization constant. This is done automatically using the 
+stepping stone estimator \cite{xie2011improving}.
+The estimate can be accessed using
+\begin{lstlisting}[language=Julia]
+stepping_stone(pt)
+\end{lstlisting}
+In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
+for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$ 
+using a computer algebra system. Note that this is very close to the output provided in 
+\cref{fig:standard_out}.
+
+\subsubsection{Online statistics}
+\label{sec:online_stats}
+Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
+we can output some basic summary statistics of the marginals of our target distribution. 
+For instance, it is straightforward to estimate the mean and variance of each 
+of the marginals of the target with 
+\begin{lstlisting}[language=Julia]
+using Statistics
+mean(pt); var(pt)
+\end{lstlisting}
+Other constant-memory statistic accumulators are made available in the 
+OnlineStats.jl \cite{day2020onlinestats} package. 
+To add additional constant-memory statistic accumulators, we 
+can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
+online documentation. For instance, we can also compute constant-memory estimates 
+of extrema of our distribution.
+
+\subsubsection{Off-memory processing}
+When either the dimensionality of the model or the number of samples is large,
+the obtained samples may not fit in memory. 
+In some cases it may be necessary to store samples to disk if our statistics of 
+interest cannot be calculated online and with constant-memory
+(see \cref{sec:online_stats}).
+We show here how to save samples to disk when Pigeons.jl is run on a single 
+machine. A similar interface can be used over MPI.
+
+\medskip 
+First, we make sure that we set \texttt{checkpoint = true}, which saves a 
+snapshot at the end of each round in the directory \texttt{results/all/<unique directory>}
+and is symlinked to \texttt{results/latest}.
+Second, we make sure that we use the \texttt{disk} recorder 
+by setting \texttt{record = [disk]}, along with possibly any other desired recorders.
+Accessing the samples from disk can then be achieved in a simple way using the Pigeons.jl 
+function \texttt{process\_sample()}.
+
+\subsubsection{PT diagnostics}
+\label{sec:PT_diagnostics}
+We describe how to produce some key parallel tempering diagnostics from 
+\cite{syed2021nrpt}.
+
+\medskip 
+The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 
+used to approximately inform the appropriate number of chains. 
+Based on \cite{syed2021nrpt}, stable PT performance should be achieved when  
+the number of chains is set to roughly $2\Lambda$. This can be achieved by 
+modifying the \texttt{n\_chains} argument in the call to \texttt{pigeons()}. 
+The global communication barrier is shown at each round and can also be accessed 
+with 
+\begin{lstlisting}[language=Julia]
+Pigeons.global_barrier(pt)
+\end{lstlisting}
+
+The number of restarts per round can be accessed with 
+\begin{lstlisting}[language=Julia]
+n_tempered_restarts(pt)
+\end{lstlisting}
+These quantities are also displayed in \cref{fig:standard_out}. 
+Many other useful PT diagnostic statistics and plots can be obtained, as described 
+in our full documentation.
+
+\subsection{Parallel and distributed PT}
+One of the main benefits of Pigeons.jl is that it allows users to easily parallelize 
+and/or distribute their PT sampling efforts. We explain how to run MPI locally on 
+one machine and also how to use MPI when a cluster is available.
+
+\subsubsection{Running MPI locally}
+To run MPI locally on one machine using four MPI processes and one thread per process,
+use
+\begin{lstlisting}[language = Julia]
+pigeons(
+    target = TuringLogPotential(model), 
+    on = ChildProcess(
+            n_local_mpi_processes = 4,
+            n_threads = 1))
+\end{lstlisting}
+
+\subsubsection{Running MPI on a cluster}
+Often, MPI is available via a cluster scheduling system. To run MPI over 
+several machines:
+\begin{enumerate}
+    \item In the cluster login node, follow the Pigeons.jl installation instructions
+    in our online documentation. 
+    \item Start Julia in the login node, and perform a one-time setup by 
+    calling \texttt{Pigeons.setup\_mpi()}.
+    \item In the Julia REPL running in the login node, run:
+\end{enumerate}
+\begin{lstlisting}[language = Julia]
+pigeons(
+    target = TuringLogPotential(model), 
+    n_chains = 1_000,
+    on = MPI(n_mpi_processes = 1_000, 
+             n_threads = 1))
+\end{lstlisting}
+The code above will start a distributed PT algorithm with 1,000 chains on 1,000 
+MPI processes each using one thread.
+Note that for the above code chunks involving \texttt{ChildProcess()} and 
+\texttt{MPI()} to work, it may be necessary to specify dependencies in their 
+function calls.
+
+\subsection{Additional options}
+\label{sec:additional_options}
+In the preceding example we only specified the target distribution and let 
+Pigeons.jl decide on default values for most other settings of the inference engine. 
+There are various settings we can change, including: 
+the random seed (\texttt{seed}), the number of PT chains (\texttt{n\_chains}), 
+the number of PT tuning rounds/samples (\texttt{n\_rounds}), 
+and a variational reference distribution family (\texttt{variational}), among other settings.
+For instance, we can run 
+\begin{lstlisting}[language = Julia]
+pigeons(
+    target = TuringLogPotential(model),
+    n_rounds = 10, 
+    n_chains = 10,
+    variational = GaussianReference(),
+    seed = 2
+)
+\end{lstlisting}
+which runs PT with the same Turing model target as before and explicitly states 
+that we should use 10 PT tuning rounds with 10 chains (described below). 
+In the above code chunk we also specify that we would like to use a 
+Gaussian variational reference distribution.
+That is, the reference distribution is chosen from a multivariate Gaussian family 
+that lies as close as possible to the target distribution in order to improve 
+the efficiency of PT. We refer readers to \cite{surjanovic2022vpt} for more details.
+When only continuous parameters are of interest, we encourage users to consider 
+using \texttt{variational = GaussianReference()} and setting \texttt{n\_chains\_variational = 10}, 
+for example, as the number of restarts may substantially increase with these settings.
+
+\section{Parallel tempering}
+\label{sec:PT}
+Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
+in \cite{syed2021nrpt}, which we outline in Algorithm~\ref{alg:distributed_PT}.
+This section gives both a brief overview of PT and some details of 
+its distributed implementation.
+
+\medskip 
+In this section we assume a basic understanding of 
+Markov chain Monte Carlo (MCMC) methods.
+For readers unfamiliar with MCMC, it is important to know that 
+it is a method to obtain approximate samples from a distribution. 
+More specifically, it is a class of algorithms for simulating a Markov chain of 
+states that look like draws from the target distribution $\pi$.
+However, with traditional 
+MCMC methods, the samples may be heavily correlated instead of independent 
+and may fail to sufficiently explore the 
+full space of the distribution (see \cref{fig:bimodal}); 
+PT is a method that aims to address these issues.
+For a more in-depth review of PT, we refer readers to \cite{surjanovic2022vpt}.
+
+\subsection{Overview of PT}
+Suppose that we would like to estimate integrals involving $\pi$, such as 
+$\int f(x) \pi(x) \, \dee x$. 
+These integrals may be multivariate and even include combinations of 
+continuous and discrete variables (where sums replace integrals in the discrete 
+case).
+One method is to obtain samples from $\pi$ to approximate such integrals.
+Often, the distribution 
+$\pi$ can be challenging for traditional MCMC methods---such 
+as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
+because of its structure.
+For example, in a bimodal example such as the one illustrated in 
+Figure~\ref{fig:bimodal},
+traditional methods might remain in one of the two modes for an extremely 
+long period of time.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.3\textwidth]{img/bimodal.pdf}
+    \caption{A simple bimodal distribution from which traditional 
+    MCMC methods may struggle to obtain samples. Blue lines display output 
+    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
+    The sampler in this figure is visibly trapped in one of the two modes.}
+    \label{fig:bimodal}
+\end{figure}
+
+To resolve this issue, PT constructs a sequence of $N$ distributions,  
+$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
+The distributions are chosen so that it is easy to obtain samples from $\pi_1$
+with the sampling difficulty increasing as one approches $\pi_N$. An example 
+of such a sequence of distributions, referred to as an \textit{annealing path},
+is shown in Figure~\ref{fig:path}.
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_1.pdf}
+      \caption*{$\pi_1$}
+    \end{subfigure}
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_2.pdf}
+      \caption*{$\pi_2$}
+    \end{subfigure}
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_3.pdf}
+      \caption*{$\pi_3$}
+    \end{subfigure}
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_4.pdf}
+      \caption*{$\pi_4$}
+    \end{subfigure}
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_5.pdf}
+      \caption*{$\pi_5$}
+    \end{subfigure}
+    \begin{subfigure}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/heatmap_path_6.pdf}
+      \caption*{$\pi_6$}
+    \end{subfigure}
+    \caption{Heatmaps of six distributions lying on an annealing path 
+    from a unimodal reference distribution  
+    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
+    $\pi_N$, which is in this case the bimodal distribution from 
+    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
+    are not directly comparable because the densities of intermediate distributions 
+    are not normalized.)}
+    \label{fig:path}
+\end{figure}
+
+We now turn to explain how this path of distributions can be used to enhance 
+sampling from the target distribution, $\pi$. 
+PT operates by first obtaining samples from each distribution on the path in parallel 
+(referred to as an \textit{exploration phase}). 
+Then, samples between adjacent distributions are swapped (referred to as 
+a \textit{communication phase}). The communication phase in PT is crucial: it allows 
+for the discovery of new regions of the space of the target distribution such as 
+the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.
+
+\subsection{Local exploration and communication}
+A struct that is useful for the implementation of PT is the \texttt{Replica}, 
+which we will often refer to in the remainder of this section. 
+A single \texttt{Replica} struct stores a \texttt{state} variable and a \texttt{chain} 
+integer, among other entries. 
+At the beginning of PT, $N$ \texttt{Replicas} are created, 
+one for each distribution on the annealing path, and 
+the chain entries in the $N$ replicas are initialized at $1,2,\ldots,N$, respectively. 
+For a given \texttt{Replica}, if the \texttt{chain} number is $i$ and the \texttt{state}
+is $x$, then this means that the sample corresponding to the $i$-th distribution in 
+the sequence $\pi_1, \ldots, \pi_N$ is currently at location $x$.
+
+\medskip 
+In the local exploration phase,
+each \texttt{Replica}'s state is modified using an MCMC move targeting $\pi_i$,
+where $i$ is given by \texttt{Replica.chain}.
+The MCMC move can either modify \texttt{Replica.state} in-place, or modify the 
+\texttt{Replica}'s \texttt{state} field. 
+This operation is indicated by the \texttt{local\_exploration} function in 
+\cref{alg:distributed_PT}.
+
+\medskip 
+In the communication phase, PT proposes swaps between pairs of replicas. 
+In principle, there are two equivalent ways to do a swap.
+In the first implementation, the \texttt{Replica}s 
+could exchange their \texttt{state} fields.
+Alternatively, they could exchange their \texttt{chain} fields.
+Because we provide distributed implementations, we use the latter as it ensures that 
+the amount of data that needs to be exchanged between two machines during a swap 
+can be made very small (two floating point numbers), resuling in 
+an exchange of $O(N)$ messages of size $O(1)$. 
+Note that this cost does not vary with the dimensionality of the state space, 
+in contrast to the first implementation that would transmit 
+$O(N)$ messages of size $O(d)$, where $d$ is the dimension of the state space,
+incurring a potentially very high communication cost for large values of $d$.
+
+\subsection{Distributed implementation}
+A distributed implementation of PT is presented in \cref{alg:distributed_PT}
+and we describe the details of the implementation below.
+We present the algorithm from the perspective that the number of machines available 
+is equal to the number of \texttt{Replica}s and distributions, $N$. 
+However, Pigeons.jl also allows for the more general case where the number of 
+machines is not necessarily equal to $N$.
+
+\subsubsection{The \texttt{PermutedDistributedArray}}
+\label{sec:permuted_dist_array}
+Recall that our theoretical $O(1)$ message size for communication between two machines 
+is achieved by exchanging \texttt{chain} indices between \texttt{Replica}s 
+instead of their \texttt{state}s.
+One difficulty can be encountered in a distributed implementation, which we 
+illustrate with the following simple example. 
+Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
+machine 1 is exploring chain 2 while machine 4 is exploring chain 3.
+At one point, the \texttt{Replica} at \texttt{chain} 2 (machine 1) might need to exchange
+\texttt{chain} indices with the \texttt{Replica} that has \texttt{chain} index 3 (machine 4). 
+However, a priori it is not clear how machine 1 should know that it should communicate 
+with machine 4 because it has no knowledge about the chain indices on the other machines.
+
+\medskip 
+To resolve this issue, we introduce a special data structure called a 
+\texttt{PermutedDistributedArray}. In the case where the number of replicas is 
+equal to the number of available machines, the construction is quite simple
+and effectively results in each machine storing one additional integer. 
+Considering the same example above, the solution to the problem is to have
+machine 1 (\texttt{chain} 2 wanting to communicate with \texttt{chain} 3) 
+communicate with \textit{machine} 3. 
+Machine 3 stores in its \texttt{dist\_array} variable of type 
+\texttt{PermutedDistributedArray} the value 4, which is the machine number that 
+stores \texttt{chain} 3. By updating these \texttt{PermutedDistributedArray} 
+variables at each communication step, we can ensure that each machine $j$ is aware 
+of which machine number currently stores \texttt{chain} $j$. 
+Therefore, the machines act as keys in a dictionary for the communication 
+permutations.
+An illustration of communication between four machines is provided in 
+\cref{fig:index_process}.
+We note that with a \texttt{PermutedDistributedArray} we make special assumptions 
+on how we access and write to the array elements. 
+Several MPI processes cooperate, with each machine storing 
+data for a slice of this distributed array, and at each time 
+step an index of the array is manipulated by exactly one machine.
+% With these assumptions we achieve read/write costs that are 
+% near-constant in the number of machines participating.
+
+\begin{figure*}[t] 
+  \centering 
+  \begin{subfigure}{0.48\textwidth}
+    \centering
+    \includegraphics[width=0.39\textwidth]{img/Communication_1_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{img/index_process_1.pdf}
+  \end{subfigure}
+  \begin{subfigure}{0.48\textwidth}
+    \centering
+    \includegraphics[width=0.39\textwidth]{img/Communication_2_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{img/index_process_2.pdf}
+  \end{subfigure}
+  \begin{subfigure}{0.48\textwidth}
+    \centering
+    \includegraphics[width=0.39\textwidth]{img/Communication_3_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{img/index_process_3.pdf}
+  \end{subfigure}
+  \begin{subfigure}{0.48\textwidth}
+    \centering
+    \includegraphics[width=0.39\textwidth]{img/Communication_4_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{img/index_process_4.pdf}
+  \end{subfigure}
+  \caption{
+    A summary of communication between four machines with $N=4$ PT chains.
+    Circles represent machines with the machine number in the center of the circle. 
+    Different colours are also used for different machines.
+    Numbers in parantheses represent the \texttt{chain} 
+    that is currently being explored by that machine. 
+    Arrows indicate communication occuring between machines to exchange \texttt{chain} 
+    indices.
+    To keep track of which chains are stored on which machine, we introduce the 
+    \texttt{dist\_array} of type \texttt{PermutedDistributedArray}, described in 
+    \cref{sec:permuted_dist_array}, which is of length $N$. 
+    The $j$-th element of \texttt{dist\_array} at a given time step indicates which 
+    machine number is storing \texttt{chain} index $j$.
+    The bolded curve in the figures to the right indicate the trajectory of the first 
+    \texttt{Replica} over the course of each of the communication steps.      
+    \textbf{Top left:} the PT replicas are initialized and the first communication step 
+    is proposed. At this time step, $\texttt{dist\_array} = [1,2,3,4]$.
+    \textbf{Top right:} after the first successful communication step, we now have 
+    $\texttt{dist\_array} = [2,1,4,3]$.
+    \textbf{Bottom left:} The second communication step is completed and 
+    $\texttt{dist\_array} = [2,4,1,3]$.
+    \textbf{Bottom right:} The third communication step is completed and 
+    $\texttt{dist\_array} = [4,2,3,1]$.}
+  \label{fig:index_process}
+\end{figure*}
+
+\subsubsection{MPI implementation details}
+We use the MPI.jl \cite{byrne2021mpi} package to support communication between machines. 
+In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
+functions \texttt{send()}, \texttt{receive!()}, 
+and \texttt{waitall()}. 
+The function \texttt{send()} has three arguments in our pseudocode: the first argument 
+is the object to be sent, the second is the machine number to which the object should 
+be sent, and finally the third argument is a unique tag for this specific send/receive 
+request at this time step and on this machine, generated by \texttt{tag()}.
+The \texttt{tag()} function can be any function that allows one to uniquely identify 
+a tag for a given send/receive request on a given machine at any given time step $t$.
+The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
+except that the first argument 
+specifies the object to which data should be written directly. Finally, 
+\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
+of machines to complete.
+
+\medskip 
+In \cref{alg:distributed_PT} we also introduce the functions 
+\texttt{permuted\_get()} and \texttt{permuted\_set!()}. 
+Given the current machine's \texttt{dist\_array} and the 
+\texttt{partner} chain with which it should communicate, 
+\texttt{permuted\_get()} returns the machine number that holds the \texttt{partner} chain. 
+Given the current machine's \texttt{dist\_array}, \texttt{chain} number, and machine number $j$,
+\texttt{permuted\_set!()} updates the \texttt{PermutedDistributedArray} variables 
+with the updated permutation.
+
+\begin{algorithm}[t]
+	\begin{algorithmic}[1]
+    \Require Initial state $x_0$, sequence of distributions $\{\pi_i\}_{i=1}^N$, 
+      number of iterations $T$, machine number $j$
+    \State $\texttt{chain} \gets j$ \Comment{current \texttt{chain} number}
+    \State $\texttt{dist\_array}[j] \gets j$ \Comment{\texttt{chain} $j$ is on machine $j$}
+
+    \For{$t$ {\bf in} 1, 2, \dots, $T$}
+		  \If{$t$ is even} \Comment{choose between even or odd swap}
+		    \State $P \gets \{i: 1 \le i < N, i \text{ is even} \}$
+		  \Else
+		    \State $P \gets \{i: 1 \le i < N, i \text{ is odd} \}$
+		  \EndIf
+
+      \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
+      \State $\texttt{partner} \gets \texttt{nothing}$
+      \If{$\texttt{chain} \in P$}
+        \State $\texttt{partner} \gets \texttt{chain}+1$
+      \ElsIf{$\texttt{chain}-1 \in P$}
+        \State $\texttt{partner} \gets \texttt{chain}-1$
+      \EndIf
+      \If{$\texttt{partner != nothing}$ \textbf{and} \text{swap is accepted}}
+        \State $\texttt{to\_machine} \gets \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
+          \Comment{retrieve \texttt{dist\_array}[\texttt{partner}]}
+        \State $\texttt{send\_tag} \gets \texttt{tag}(t, \texttt{chain}, j)$
+        \State $\texttt{send}(\texttt{chain}, \texttt{to\_machine}, \texttt{send\_tag})$
+        \State $\texttt{receive\_tag} \gets \texttt{tag}(t, \texttt{chain}, \texttt{to\_machine})$
+        \State $\texttt{receive!}(\texttt{chain}, \texttt{to\_machine}, \texttt{receive\_tag})$
+        \State $\texttt{waitall(requests)}$ 
+          \Comment{wait for MPI requests for this pair of machines to complete}
+        \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$
+          \Comment{set value of \texttt{dist\_array}[\texttt{chain}] to \texttt{to\_machine}}
+      \EndIf
+      \State $c_t \gets \texttt{chain}$ 
+		\EndFor
+    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
+	\end{algorithmic}
+  \caption{Distributed PT on machine $j$ (one replica per machine)}
+  \label{alg:distributed_PT}
+\end{algorithm}
+
+\section{Strong parallelism invariance}
+\label{sec:PI_causes}
+In this section we describe potential violations of 
+strong parallelism invariance (SPI) that can occur in a distributed setting. 
+We also explain how we avoid these issues by using special 
+distributed reduction schemes and splittable random number generators. 
+Insights provided in this section can be applied 
+to general distributed software beyond Julia.
+
+\medskip 
+We have identified two factors that can cause violations of our previously-defined 
+SPI that standard Julia libraries 
+do not automatically take care of:
+\begin{enumerate}   
+  \item \textbf{Non-associativity of floating point operations:} When 
+  several workers perform distributed reduction of floating point values, 
+  the output of this reduction will be slightly different depending on the order taken 
+  during reduction. 
+  When these reductions are then fed into further random operations, 
+  this implies that two randomized algorithms with the same seed but using a 
+  different number of workers will eventually arbitrarily diverge.
+
+  \item \textbf{Global, thread-local, and task-local random number generators:} 
+  These are the dominant approaches to parallel random number generators in current languages, 
+  and an appropriate understanding of these RNGs is necessary. In particular, 
+  in Julia it is important to understand the behaviour of the \texttt{@threads} macro.
+\end{enumerate}
+Our focus in the remainder of this section is to describe how our implementation 
+solves the two above issues.
+
+\subsection{Distributed reduction and floating point non-associativity}
+After each round of PT, the machines need to exchange information such as the 
+average swap acceptance probabilities, statistics to adapt a variational 
+reference and adjust annealing parameters, and so on. 
+For instance, suppose our \texttt{state} is univariate and real-valued and 
+that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
+is visited as well as the mean of the univariate \texttt{state}s from the target chain.
+To obtain the final estimate of the mean of the target distribution, we would like 
+to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
+the number of times that each \texttt{Replica} visited the target chain.   
+This process involves summing floating-point values that are located on each 
+of the \texttt{Replica}s/machines and is an example of a reduction scheme.
+
+\medskip 
+To illustrate why distributed reduction with floating point values can violate 
+strong parallelism invariance if not properly implemented, we consider the 
+following toy example. 
+Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
+as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
+in the following examples.
+In this case, if our reduction procedure is to sum the floating point numbers, 
+we know that our final answer should be approximately $36x$. 
+However, depending on the exact order in which floating point addition is carried out, 
+the answers might not all be the same and exactly equal to $36x$.
+For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
+for eight machines would be given by 
+\[
+  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
+  \approx 978.5814582452562.
+\]
+In contrast, with two machines, one possible order of operations might be 
+\[
+  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
+  \approx 978.5814582452563.
+\]
+A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.
+
+\medskip 
+To avoid the issue of non-associativity of 
+floating point arithmetic, we ensure that the order in which operations are performed 
+is exactly the same, irrespective of the number of processes/machines and threads. 
+This is achieved by making sure that every value to be added---if addition is our 
+reduction operation---is a leaf node in the reduction tree, irrespective of the 
+number of machines available to perform the reduction. 
+For instance, if $N$ values are to be reduced, then the reduction tree would have 
+$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
+in such a way that the order of operations is as if there were $N$ machines available. 
+To do so, it may be necessary for a machine to ``communicate with itself'', 
+imitating the behaviour that would be present if there were $N$ machines available.
+\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
+illustrate the reduction procedure for eight and two machines, respectively.
+
+\begin{figure}[t]
+  \centering
+  \begin{forest}
+    for tree = {circle, draw, minimum size = 0.8cm}
+      [{$36x$}
+        [{$10x$}
+          [{$3x$}
+            [{$1x$}]
+            [{$2x$}]
+          ]
+          [{$7x$}
+            [{$3x$}] 
+            [{$4x$}] 
+          ]
+        ]
+        [{$26x$}
+          [{$11x$}
+            [{$5x$}] 
+            [{$6x$}] 
+          ]
+          [{$15x$}
+            [{$7x$}] 
+            [{$8x$}] 
+          ]
+        ]
+      ]
+  \end{forest}
+  \caption{Adding eight floating point numbers $\cbra{1x, 2x, \ldots, 8x}$ 
+  across eight machines. 
+  Additions in each row of the tree can be performed in parallel. 
+  The final result is stored in the root node of the tree and can be represented 
+  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$, indicating 
+  the order of operations.}
+  \label{fig:reduction_tree_8}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \begin{forest}
+    for tree = {circle, draw, minimum size = 0.8cm}
+      [{$36x$}
+        [{$10x$}
+          [{$6x$}
+            [{$3x$}
+              [{$1x$}]
+              [{$2x$}]
+            ]
+            [{$3x$}]
+          ]
+          [{$4x$}]
+        ]
+        [{$26x$}
+          [{$8x$}]
+          [{$18x$}
+            [{$7x$}]
+            [{$11x$}
+              [{$6x$}]
+              [{$5x$}]
+            ]
+          ]
+        ]
+      ]
+  \end{forest}
+  \caption{One possible way of adding eight floating point numbers 
+  $\cbra{1x, 2x, \ldots, 8x}$ across two machines. 
+  The final result is stored in the root node of the tree and can be represented 
+  by the expression $(((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x)$.
+  Note that the order of operations in this expression is different from the one 
+  presented in \cref{fig:reduction_tree_8}.}
+  \label{fig:reduction_tree_2}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \begin{forest}
+    for tree = {circle, draw, minimum size = 0.8cm}
+      [{$36x$}, fill = julia1
+        [{$10x$}, fill = julia1
+          [{$3x$}, fill = julia1
+            [{$1x$}, fill = julia1]
+            [{$2x$}, fill = julia2]
+          ]
+          [{$7x$}, fill = julia3
+            [{$3x$}, fill = julia3] 
+            [{$4x$}, fill = julia4] 
+          ]
+        ]
+        [{$26x$}, fill = julia5
+          [{$11x$}, fill = julia5
+            [{$5x$}, fill = julia5] 
+            [{$6x$}, fill = julia6] 
+          ]
+          [{$15x$}, fill = julia7
+            [{$7x$}, fill = julia7] 
+            [{$8x$}, fill = julia8] 
+          ]
+        ]
+      ]
+  \end{forest}
+  \caption{Addition of eight floating point numbers across eight machines with a 
+  guarantee on SPI. Each machine is represented by a different colour. 
+  The final result can be represented 
+  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
+  \label{fig:reduction_tree_colour_8}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \begin{forest}
+    for tree = {circle, draw, minimum size = 0.8cm}
+      [{$36x$}, fill = julia1
+        [{$10x$}, fill = julia1
+          [{$3x$}, fill = julia1
+            [{$1x$}, fill = julia1]
+            [{$2x$}, fill = julia1]
+          ]
+          [{$7x$}, fill = julia1
+            [{$3x$}, fill = julia1] 
+            [{$4x$}, fill = julia1] 
+          ]
+        ]
+        [{$26x$}, fill = julia2
+          [{$11x$}, fill = julia2
+            [{$5x$}, fill = julia2] 
+            [{$6x$}, fill = julia2] 
+          ]
+          [{$15x$}, fill = julia2
+            [{$7x$}, fill = julia2] 
+            [{$8x$}, fill = julia2] 
+          ]
+        ]
+      ]
+  \end{forest}
+  \caption{Addition of eight floating point numbers across two machines with a 
+  guarantee on SPI. Each machine is represented by a different colour. 
+  The final result can be represented 
+  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$,
+  which is the same as that given by eight machines in \cref{fig:reduction_tree_colour_8}.}
+  \label{fig:reduction_tree_colour_2}
+\end{figure}
+
+\subsection{Splittable random streams}
+\label{sec:splittable_randoms}
+Another building block towards achieving SPI is a \emph{splittable random stream}
+\cite{lecuyer1988splittable,burton1992splittable}. 
+Julia uses \emph{task-local} random number generators, a notion that is 
+related to (but does not necessarily imply) strong parallelism invariance. 
+A \emph{task} is a unit of work on a machine.
+A task-local RNG would then mean that a separate RNG is 
+used for each unit of work, hopefully implying strong parallelism invariance
+if the number of tasks is assumed to be constant. 
+Unfortunately, this is not the case when a separate task is created for each thread 
+of execution in Julia.
+We note that the \texttt{@threads} macro in Julia creates \texttt{nthreads()}
+tasks and thus \texttt{nthreads()} pseudorandom number generators.
+This can break strong parallelism invariance as the output may depend on the number of 
+threads.
+
+\medskip 
+To motivate splittable random streams, consider the following toy example 
+that violates our notion of SPI:
+\begin{lstlisting}[language = Julia]
+using Random
+import Base.Threads.@threads
+
+println("Num. of threads: $(Threads.nthreads())")
+
+const n_iters = 10000
+result = zeros(n_iters)
+Random.seed!(1)
+@threads for i in 1:n_iters
+    result[i] = rand()
+end
+println("Result: $(last(result))")
+\end{lstlisting}
+With eight threads, this outputs:
+\begin{lstlisting}[language = Julia]
+Num. of threads: 8
+Result: 0.25679999169092793
+\end{lstlisting}
+Julia guarantees that if we rerun this code, as long as we are using eight threads, 
+we will always get the same result, irrespective of the multi-threading 
+scheduling decisions implied by the \texttt{@threads}-loop. 
+Internally, Julia works with task-local RNGs and the \texttt{@threads} macro 
+spawns \texttt{nthreads()} number of task-local RNGs.
+For this reason, with a different number of threads, the result is different:
+\begin{lstlisting}[language = Julia]
+Num. of threads: 1
+Result: 0.8785201210435906
+\end{lstlisting}
+
+In this simple example above, the difference in output is perhaps not too concerning, 
+but for our parallel tempering 
+use case, the distributed version of the algorithm is significantly more complex 
+and difficult to debug compared to the single-threaded one. We therefore take task-local 
+random number generation one step further and achieve SPI, which 
+guarantees that the output is not only reproducible with respect to repetitions 
+for a fixed number of threads, but also for different numbers of threads or processes.
+
+\medskip 
+A first step to achieve this is to associate one random number 
+generator to each \texttt{Replica}. 
+To do so, we use our SplittableRandoms.jl package, which is a Julia implementation 
+of Java SplittableRandoms. Our package offers an implementation of  
+SplitMix64 \cite{steele2014fast},
+which allows us to turn one seed into an arbitrary collection of pseudo-independent 
+RNGs. 
+A quick example of how to use the SplittableRandoms.jl library is given below. 
+By splitting a master RNG using the \texttt{split()} function, 
+we can achieve SPI even with the use of the \texttt{@threads} macro.
+
+\begin{lstlisting}[language = Julia]
+using Random
+using SplittableRandoms: SplittableRandom, split
+import Base.Threads.@threads
+
+println("Num. of threads: $(Threads.nthreads())")
+
+const n_iters = 10000
+const master_rng = SplittableRandom(1)
+result = zeros(n_iters)
+rngs = [split(master_rng) for _ in 1:n_iters]
+Random.seed!(1)
+@threads for i in 1:n_iters
+    result[i] = rand(rngs[i])
+end
+println("Result: $(last(result))")
+\end{lstlisting}
+With one and eight threads, the code above outputs
+\begin{lstlisting}[language = Julia]
+Num. of threads: 1
+Result: 0.4394633333251359
+\end{lstlisting}
+\begin{lstlisting}[language = Julia]
+Num. of threads: 8
+Result: 0.4394633333251359
+\end{lstlisting}
+
+\section{Related work}
+Automated software packages for Bayesian inference have revolutionized Bayesian data 
+analysis in the past two decades, and are now a core part of a typical applied statistics 
+workflow. For example, packages such as BUGS 
+\cite{lunn2013bugs, lunn2009bugs, lunn2000winbugs}, 
+JAGS \cite{hornik2003jags}, Stan \cite{carpenter2017stan}, PyMC3 
+\cite{salvatier2016probabilistic}, and Turing.jl \cite{ge2018turing} have been widely used in 
+many scientific applications.
+
+\medskip 
+These software packages often provide two key user-facing components: (1) a 
+probabilistic programming language (PPL), which allows users to specify Bayesian 
+statistical models in code with a familiar, mathematics-like syntax; and (2) an 
+inference engine, which is responsible for performing computational Bayesian
+inference once the model and data have been specified. 
+Pigeons.jl focuses on the development 
+of a new inference engine that employs distributed, high-performance computing.
+
+\medskip
+Inference engines available in existing software packages are varied in their 
+capabilities and limitations. For instance, the widely-used Stan inference engine 
+is only capable of handling real-valued (i.e., continuous) parameters. 
+Other tools, such as JAGS, are capable of handling discrete-valued parameters but 
+are limited in their capability to handle custom data-types 
+(e.g. phylogenetic trees). 
+Of those that have the capability to model arbitrary data types, 
+none have an automatically distributed implementation to our knowledge.
+Turing.jl \cite{ge2018turing} offers another popular inference engine and PPL, 
+however Pigeons.jl allows one to interface with several different PPLs as well as 
+to easily perform distributed computation.
+
+\medskip
+There is also a vast literature on distributed and parallel 
+Bayesian inference algorithms 
+\cite{bardenet2017markov, brockwell2006parallel, calderhead2014general,
+jacob2020unbiased, jacob2011using, lee2010utility, scott2016bayes, 
+wang2015parallelizing, wu2017average, zhu2017big}.
+These methods unfortunately do not lead to widely 
+usable software packages because they either introduce unknown amounts of 
+approximation error, involve significant communication cost, or reduce the 
+generality of Bayesian inference.
+
+\section{Conclusion}
+Pigeons.jl is a Julia package that enables users with no experience in distributed 
+computing to efficiently approximate posterior distributions and solve challenging 
+Lebesgue integration problems over a distributed computing environment.
+The core algorithm behind Pigeons.jl is distributed, non-reversible parallel tempering 
+\cite{syed2021nrpt,surjanovic2022vpt}. Pigeons.jl can be used in a multi-threaded context, 
+as well as distributed over up to thousands of MPI-communicating machines. 
+Further, Pigeons.jl is designed so that for a given seed, the output is \textit{identical}
+regardless of the number of threads or processes used.
+
+% \input{src/introduction.tex}
+% \input{src/examples.tex}
+% \input{src/PT.tex}
+% \input{src/SPI.tex}
+% \input{src/related.tex}
+% \input{src/conclusion.tex}
 
@@
 which is funded by grants from the John Templeton Foundation and the 
-Gordon and Betty Moore Foundation to Harvard University, 
-as well as support by the National Science Foundation grants AST-1935980 
+Gordon and Betty Moore Foundation to Harvard University.
+PT also acknowledges support by the National Science Foundation grants AST-1935980 
 and AST-2034306 and the Gordon and Betty Moore Foundation (GBMF-10423).

### 2023-10-31 paper/conclusion.tex
use paper folder

\section{Conclusion}
Pigeons.jl is a Julia package that enables users with no experience in distributed 
computing to efficiently approximate posterior distributions and solve challenging 
Lebesgue integration problems over a distributed computing environment.
The core algorithm behind Pigeons.jl is distributed, non-reversible parallel tempering 
\cite{syed2021nrpt,surjanovic2022vpt}. Pigeons.jl can be used in a multi-threaded context, 
as well as distributed over up to thousands of MPI-communicating machines. 
Further, Pigeons.jl is designed so that for a given seed, the output is \textit{identical}
regardless of the number of threads or processes used.

### 2023-10-31 paper/related.tex
use paper folder

\section{Related work}
Automated software packages for Bayesian inference have revolutionized Bayesian data 
analysis in the past two decades, and are now a core part of a typical applied statistics 
workflow. For example, packages such as BUGS 
\cite{lunn2013bugs, lunn2009bugs, lunn2000winbugs}, 
JAGS \cite{hornik2003jags}, Stan \cite{carpenter2017stan}, PyMC3 
\cite{salvatier2016probabilistic}, and Turing.jl \cite{ge2018turing} have been widely used in 
many scientific applications.

\medskip 
These software packages often provide two key user-facing components: (1) a 
probabilistic programming language (PPL), which allows users to specify Bayesian 
statistical models in code with a familiar, mathematics-like syntax; and (2) an 
inference engine, which is responsible for performing computational Bayesian
inference once the model and data have been specified. 
Pigeons.jl focuses on the development 
of a new inference engine that employs distributed, high-performance computing.

\medskip
Inference engines available in existing software packages are varied in their 
capabilities and limitations. For instance, the widely-used Stan inference engine 
is only capable of handling real-valued (i.e., continuous) parameters. 
Other tools, such as JAGS, are capable of handling discrete-valued parameters but 
are limited in their capability to handle custom data-types 
(e.g. phylogenetic trees). 
Of those that have the capability to model arbitrary data types, 
none have an automatically distributed implementation to our knowledge.
Turing.jl \cite{ge2018turing} offers another popular inference engine and PPL, 
however Pigeons.jl allows one to interface with several different PPLs as well as 
to easily perform distributed computation.

\medskip
There is also a vast literature on distributed and parallel 
Bayesian inference algorithms 
\cite{bardenet2017markov, brockwell2006parallel, calderhead2014general,
jacob2020unbiased, jacob2011using, lee2010utility, scott2016bayes, 
wang2015parallelizing, wu2017average, zhu2017big}.
These methods unfortunately do not lead to widely 
usable software packages because they either introduce unknown amounts of 
approximation error, involve significant communication cost, or reduce the 
generality of Bayesian inference.

### 2023-10-31 paper/header.tex
fix

--- 2023-08-18 header.tex
@@
 
-\keywords{Distributed computation, Bayesian inference, parallelism invariance, 
-multi-threading, Message Passing Interface, Markov chain Monte Carlo}
+\keywords{distributed computation, Bayesian inference, parallelism invariance, multi-threading, Message Passing Interface, Markov chain Monte Carlo}
 
 \hypersetup{
-pdftitle = {My JuliaCon proceeding},
-pdfsubject = {JuliaCon 2019 Proceedings},
-pdfauthor = {1st author, 2nd author, 3rd author},
-pdfkeywords = {Julia, Optimization, Game theory, Compiler},
+pdftitle = {Pigeons.jl: Distributed sampling from intractable distributions},
+pdfsubject = {JuliaCon 2022 Proceedings},
+pdfauthor = {Nikola Surjanovic, Miguel Biron-Lattes, Paul Tiede, Saifuddin Syed, Trevor Campbell, Alexandre Bouchard-Côté},
+pdfkeywords = {distributed computation, Bayesian inference, parallelism invariance, multi-threading, Message Passing Interface, Markov chain Monte Carlo},
 }

### 2023-11-01 paper/header.tex
try debug

@@
 \author[1]{Trevor Campbell}
-\author[1]{Alexandre Bouchard-Côté}
+\author[1]{Alexandre Bouchard-C\^{o}t\'{e}}
 \affil[1]{University of British Columbia}
@@
 pdfsubject = {JuliaCon 2022 Proceedings},
-pdfauthor = {Nikola Surjanovic, Miguel Biron-Lattes, Paul Tiede, Saifuddin Syed, Trevor Campbell, Alexandre Bouchard-Côté},
+pdfauthor = {Nikola Surjanovic, Miguel Biron-Lattes, Paul Tiede, Saifuddin Syed, Trevor Campbell, Alexandre Bouchard-C\^{o}t\'{e}},
 pdfkeywords = {distributed computation, Bayesian inference, parallelism invariance, multi-threading, Message Passing Interface, Markov chain Monte Carlo},

### 2024-03-23 paper/PT.tex
use paper folder
First updates based on reviewer comments

--- 2023-08-18 src/PT.tex
@@
     \centering
-    \includegraphics[width=0.3\textwidth]{img/bimodal.pdf}
+    \includegraphics[width=0.3\textwidth]{../img/bimodal.pdf}
     \caption{A simple bimodal distribution from which traditional 
-    MCMC methods would struggle to obtain samples. Blue lines display output 
+    MCMC methods may struggle to obtain samples. Blue lines display output 
     from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
@@
     \centering
-    \begin{subfigure}{0.15\textwidth}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_1.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
       \caption*{$\pi_1$}
-    \end{subfigure}
-    \begin{subfigure}{0.15\textwidth}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_2.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
       \caption*{$\pi_2$}
-    \end{subfigure}
-    \begin{subfigure}{0.15\textwidth}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_3.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
       \caption*{$\pi_3$}
-    \end{subfigure}
-    \begin{subfigure}{0.15\textwidth}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_4.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
       \caption*{$\pi_4$}
-    \end{subfigure}
-    \begin{subfigure}{0.15\textwidth}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_5.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
       \caption*{$\pi_5$}
-    \end{subfigure}
-    \begin{subfigure}{0.15\textwidth}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/heatmap_path_6.pdf}
+      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
       \caption*{$\pi_6$}
-    \end{subfigure}
+    \end{minipage}
     \caption{Heatmaps of six distributions lying on an annealing path 
@@
   \centering 
-  \begin{subfigure}{0.48\textwidth}
+  \begin{minipage}{0.48\textwidth}
     \centering
-    \includegraphics[width=0.39\textwidth]{img/Communication_1_v3.pdf}
-    \includegraphics[width=0.59\textwidth]{img/index_process_1.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=0.39\textwidth]{../img/Communication_1_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{../img/index_process_1.pdf}
+  \end{minipage}
+  \begin{minipage}{0.48\textwidth}
     \centering
-    \includegraphics[width=0.39\textwidth]{img/Communication_2_v3.pdf}
-    \includegraphics[width=0.59\textwidth]{img/index_process_2.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=0.39\textwidth]{../img/Communication_2_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{../img/index_process_2.pdf}
+  \end{minipage}
+  \begin{minipage}{0.48\textwidth}
     \centering
-    \includegraphics[width=0.39\textwidth]{img/Communication_3_v3.pdf}
-    \includegraphics[width=0.59\textwidth]{img/index_process_3.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=0.39\textwidth]{../img/Communication_3_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{../img/index_process_3.pdf}
+  \end{minipage}
+  \begin{minipage}{0.48\textwidth}
     \centering
-    \includegraphics[width=0.39\textwidth]{img/Communication_4_v3.pdf}
-    \includegraphics[width=0.59\textwidth]{img/index_process_4.pdf}
-  \end{subfigure}
+    \includegraphics[width=0.39\textwidth]{../img/Communication_4_v3.pdf}
+    \includegraphics[width=0.59\textwidth]{../img/index_process_4.pdf}
+  \end{minipage}
   \caption{
@@
 \end{algorithm}
+
+\subsection{Strong and weak scaling of PT}
+
+We study the strong and weak scaling properties of Pigeons' implementation of PT
+as given by Amdahl's law and Gustafson's law, respectively.
+For the strong scaling results, we consider a fixed problem size and compare the 
+compute time as we vary the number of processes.  
+For weak scaling, we consider a varying problem size proportional 
+to the number of processes used. 
+In both cases we use the number of chains as the definition of the problem size; 
+this is in line with previous heuristics that establish that the generaly problem 
+difficulty is proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
+
+\medskip
+We emphasize again that Pigeons supports both shared memory 
+parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
+might be useful to run parallel tempering with very many chains or 
+to run several instances of parallel tempering at once. In such cases, 
+the DMP setup is the most natural. Our implementation in Pigeons ensures that the 
+communication cost in this setup is minimal. In most cases, the 
+communication cost is negligible compared to other computations performed on each process.
+
+\medskip
+The results for strong and weak scaling are presented in \cref{fig:PT_scaling}. 
+For strong scaling, we see that ...
+For weak scaling, We see that ...
+\NS{Comment on plots}.
+
+\begin{figure*}[t]
+    \centering
+    \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
+    \end{minipage}
+    \caption{
+        \NS{Add appropriate plots.}
+        Strong and weak scaling results for PT.  
+        \textbf{Left:} Strong scaling. 
+        \textbf{Right:} Weak scaling. 
+    }
+    \label{fig:PT_scaling}
+\end{figure*}

### 2024-03-23 paper/introduction.tex
First updates based on reviewer comments

--- 2023-08-18 src/introduction.tex
@@
 single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
-machines. Further, it comes with guarantees on strong parallelism invariance wherein 
+machines (as tested on local compute clusters). Further, it comes with guarantees on strong parallelism invariance wherein 
 the output for a given seed is \emph{identical} irrespective 

### 2024-03-23 paper/examples.tex
use paper folder
First updates based on reviewer comments
Update examples.tex

--- 2023-08-18 src/examples.tex
@@
 \url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
+\NS{Add InferenceHub description and link.}
+
+\medskip
 We begin by installing the latest official release of Pigeons.jl: 
@@
   \gamma(p_1, p_2) &= 
-    {n \choose y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
+    \binom{n}{y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
     \label{eq:coinflip_density} \\
@@
     \centering 
-    \includegraphics[width=0.5\textwidth]{img/coinflip_posterior.pdf}
+    \includegraphics[width=0.5\textwidth]{../img/coinflip_posterior.pdf}
     \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
@@
     \centering
-    \begin{subfigure}{0.45\textwidth}
+    \begin{minipage}{0.45\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/trace_density_plots_pt.pdf}
-    \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{img/trace_density_plots_single_chain.pdf}
-    \end{subfigure}
+      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
+    \end{minipage}
     \caption{

### 2024-05-20 paper/PT.tex
Add weak scaling plots

@@
 
-\subsection{Strong and weak scaling of PT}
+\subsection{Weak scaling of PT}
 
-We study the strong and weak scaling properties of Pigeons' implementation of PT
-as given by Amdahl's law and Gustafson's law, respectively.
-For the strong scaling results, we consider a fixed problem size and compare the 
-compute time as we vary the number of processes.  
-For weak scaling, we consider a varying problem size proportional 
-to the number of processes used. 
-In both cases we use the number of chains as the definition of the problem size; 
-this is in line with previous heuristics that establish that the generaly problem 
-difficulty is proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
+We study the weak scaling properties of Pigeons' implementation of PT
+as given by Gustafson's law.
+In this setup, we consider a varying problem size proportional to the number of processes used. 
+The number of chains is our definition of the problem size; 
+this is in line with previous heuristics that establish that the problem 
+difficulty is generally proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
+In particular, we consider a multivariate normal target distribution 
+and vary the dimension $d \in \cbra{2^1, 2^2, \ldots, 2^{12}}$ for the distributed setup and 
+$d \in \cbra{2^1, 2^2, \ldots, 2^{10}}$ for the parallel (shared memory) setup. 
+The number of chains is set to $\ceil{\sqrt{d}}$ according to PT tuning suggestions \cite{syed2021nrpt} 
+and for the distributed and parallel settings we set the number of processes or threads equal 
+to the number of chains.
+Within each chain we use the autoMALA sampler \cite{biron2024automala}, which is a 
+gradient-based sampler with automatic step size selection in the Metropolis-adjusted Langevin algorithm.
+Ten different seeds are used in order to obtain approximate 95\% confidence intervals of 
+the speedup.
 
@@
 \medskip
-The results for strong and weak scaling are presented in \cref{fig:PT_scaling}. 
-For strong scaling, we see that ...
-For weak scaling, We see that ...
-\NS{Comment on plots}.
+The weak scaling results for both the DMP and SMP settings are presented in \cref{fig:PT_scaling}. 
+We see that we obtain an approximately linear speedup as we increase the number of processes.
 
@@
     \centering
-    \begin{minipage}{0.45\textwidth}
+    \begin{minipage}{0.4\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
+      \includegraphics[width=\textwidth]{../img/speedup_distributed.pdf}
     \end{minipage}
-    \begin{minipage}{0.45\textwidth}
+    \begin{minipage}{0.4\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
+      \includegraphics[width=\textwidth]{../img/speedup_parallel.pdf}
     \end{minipage}
     \caption{
-        \NS{Add appropriate plots.}
-        Strong and weak scaling results for PT.  
-        \textbf{Left:} Strong scaling. 
-        \textbf{Right:} Weak scaling. 
+        Weak scaling results for PT in both the distributed (MPI) and parallel settings 
+        with approximate 95\% confidence intervals for the speedup.  
+        The blue dotted line represents a hypothetical speedup equal to the 
+        number of processes/threads requested. 
+        The red dashed line indicates a speedup of one for reference.
+        \textbf{Left:} Distributed setup. 
+        \textbf{Right:} Parallel (shared) setup. 
     }

### 2024-05-20 paper/examples.tex
Add weak scaling plots

@@
 that demonstrate how to use Pigeons.jl for sampling. 
-For further reading, we also direct readers to our growing list of examples, which can be found at 
-\url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
-\NS{Add InferenceHub description and link.}
+We also direct readers to our growing list of examples at InferHub 
+(\url{https://julia-tempering.github.io/InferHub/}), which hosts a growing collection of posterior 
+distributions with an emphasis on difficult (non-log-concave) problems.
 
@@
 \cref{eq:coinflip_density} up to a normalizing constant.
-We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
+We use non-reversible parallel tempering \NS{add newest ergodicity cite?} 
+\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt} (PT), Pigeons.jl's 
 state-of-the-art sampling algorithm, to sample from the target distribution.

### 2024-05-21 paper/examples.tex
Add new cite and fix typo

@@
 \cref{eq:coinflip_density} up to a normalizing constant.
-We use non-reversible parallel tempering \NS{add newest ergodicity cite?} 
-\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt} (PT), Pigeons.jl's 
+We use non-reversible parallel tempering 
+\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity} (PT), Pigeons.jl's 
 state-of-the-art sampling algorithm, to sample from the target distribution.

### 2024-05-21 paper/PT.tex
Add new cite and fix typo
New tweaks based on Alex's comments

@@
 \medskip
-We emphasize again that Pigeons supports both shared memory 
-parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
-might be useful to run parallel tempering with very many chains or 
-to run several instances of parallel tempering at once. In such cases, 
-the DMP setup is the most natural. Our implementation in Pigeons ensures that the 
-communication cost in this setup is minimal. In most cases, the 
-communication cost is negligible compared to other computations performed on each process.
-
-\medskip
-The weak scaling results for both the DMP and SMP settings are presented in \cref{fig:PT_scaling}. 
+The weak scaling results for both the distributed memory parallelism (DMP) and 
+shared memory parallelism (SMP) settings are presented in \cref{fig:PT_scaling}. 
 We see that we obtain an approximately linear speedup as we increase the number of processes.
@@
         The blue dotted line represents a hypothetical speedup equal to the 
-        number of processes/threads requested. 
+        number of processes/threads. 
         The red dashed line indicates a speedup of one for reference.
         \textbf{Left:} Distributed setup. 
-        \textbf{Right:} Parallel (shared) setup. 
+        \textbf{Right:} Parallel (shared memory) setup. 
     }
@@
 \end{figure*}
+
+\subsection{Relevance of DMP versus SMP}
+
+We emphasize again that Pigeons supports both shared memory 
+parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
+might be useful to run parallel tempering with very many chains or 
+to run several instances of parallel tempering at once. In such cases, 
+the DMP setup is the most natural if the requested number of chains is very large. 
+Another notable example of the benefit of DMP is for the multi-GPU use case. 
+For instance, in some situations the likelihood evaluation of a single chain 
+requires its own GPU (e.g., for high-resolution black hole imaging, which relies 
+on running a Fast Fourier Transform on a GPU). 
+Additionally, DMP can be useful when the user's code is not thread safe. 
+In such cases, concurrency can still be achieved via DMP.
+Our implementation in Pigeons ensures that the 
+communication cost in the DMP setup is minimal. In most cases, the 
+communication cost is negligible compared to other computations performed on each process.

### 2024-05-21 paper/examples.tex
New tweaks based on Alex's comments

@@
 \label{sec:online_stats}
+Pigeons has facilities to support cases requiring large memory. For instance, we
+allow for the computation of online statistics, as well as off-memory sample storage.
 Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 

### 2024-05-22 paper/introduction.tex
Changes based on Trevor's comments

--- 2023-08-18 src/introduction.tex
@@
 that leverage distributed computation. Its simple API allows users to perform such computations 
-single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
-machines. Further, it comes with guarantees on strong parallelism invariance wherein 
+single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
+We have stress-tested Pigeons successfully with up to 1000 machines running 
+concurrently on a compute cluster (Sockeye) at the University of British Columbia.
+Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
 the output for a given seed is \emph{identical} irrespective 

### 2024-05-22 paper/PT.tex
Changes based on Trevor's comments
Update based on Trevor's comments

@@
 
+\medskip
 To resolve this issue, PT constructs a sequence of $N$ distributions,  
@@
 
+\medskip
 We now turn to explain how this path of distributions can be used to enhance 
@@
 
-We study the weak scaling properties of Pigeons' implementation of PT
-as given by Gustafson's law.
+We study the weak scaling properties of Pigeons' implementation of PT.
 In this setup, we consider a varying problem size proportional to the number of processes used. 
@@
 Another notable example of the benefit of DMP is for the multi-GPU use case. 
-For instance, in some situations the likelihood evaluation of a single chain 
-requires its own GPU (e.g., for high-resolution black hole imaging, which relies 
-on running a Fast Fourier Transform on a GPU). 
+For instance, in some situations the likelihood evaluation of a single chain may 
+require its own GPU (e.g., if one is running a Fast Fourier Transform on a GPU). 
 Additionally, DMP can be useful when the user's code is not thread safe. 

### 2024-05-22 paper/examples.tex
Update based on Trevor's comments

@@
 In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
-for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$ 
-using a computer algebra system. Note that this is very close to the output provided in 
-\cref{fig:standard_out}.
+for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$.
+Note that this is very close to the output provided in \cref{fig:standard_out}.
 

### 2024-05-27 paper/examples.tex
Starting to go through paper...

@@
 We also direct readers to our growing list of examples at InferHub 
-(\url{https://julia-tempering.github.io/InferHub/}), which hosts a growing collection of posterior 
+(\url{https://julia-tempering.github.io/InferHub/}), which hosts a  collection of posterior 
 distributions with an emphasis on difficult (non-log-concave) problems.
@@
 In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
-applied to a non-identifiable ``coinflip'' data set, modified from the example
-at \url{https://turing.ml/v0.22/docs/using-turing/quick-start}. 
+applied to a non-identifiable ``coinflip'' data set.
 The Bayesian model can be formulated as 
@@
   \label{eq:coinflip}
-  p_1, p_2 &\stackrel{iid}{\sim} U(0, 1) \\    
-  Y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
+  p_1, p_2 &\stackrel{\text{i.i.d.}}{\sim} U(0, 1), \\    
+  y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
 \]
-The random variable $Y$ is the number of heads observed on $n$ coin flips
+The random variable $y$ is the number of heads observed on $n$ coin flips
 where the probability of heads is $p_1 p_2$.  
@@
     \begin{lstlisting}
-          --------------------------------------------------------------------------------------
             #scans    restarts        Λ      time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
-          ---------- ---------- ---------- ---------- ---------- ---------- ---------- --------- 
+
                   2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
@@
            1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
-          --------------------------------------------------------------------------------------
+
     \end{lstlisting}
@@
     of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
-    such as the number of allocations per round, time (in seconds), and estimates 
+    such as the amount of memory allocations per round (in bytes), time (in seconds), and estimates 
     of the log of the normalization constants. 
@@
 Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
-The only requirement on the foreign language is that it supports 
-reading the standard input and writing to the standard output, 
-as described in our online documentation.
+See \url{https://pigeons.run/stable/input-overview/}.
 
@@
     \item \texttt{restarts}: a higher number is better. 
-    Informally, PT performs well at sampling from high-dimensional and/or multi-modal 
-    distributions by pushing samples from an easy-to-sample distribution (the reference) 
+    Informally, PT propagates samples from an easy-to-sample distribution (the reference) 
     to the more difficult target distribution. 
@@
 is displayed in \cref{fig:coinflip_posterior}.
-\footnote{The code chunk above only works with Julia 1.9.}
+\footnote{The code chunk above requires Julia 1.9.}
 

### 2024-05-27 paper/introduction.tex
Starting to go through paper...

@@
 In many scientific application domains, the ability to obtain samples from a 
-distribution is crucial. 
+probability distribution is of central interest. 
 For instance, sampling methods have been used to discover magnetic polarization 
@@
 Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
-enables users to sample efficiently from high-dimensional and complex distributions 
+enables users to sample efficiently from challenging probability distributions 
 and solve integration problems by 
 implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
-that leverage distributed computation. Its simple API allows users to perform such computations 
+that leverage distributed computation. Its simple user interface facilitates performing such computations 
 single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
@@
 \end{enumerate}
+
+\subsection{Latest documentation}
+
+More in-depth documentation and additional examples can be found at: \\ \url{https://pigeons.run/}.
 
@@
 \end{enumerate}
-Its implementation shines compared to traditional sampling packages in the 
+The Pigeons package particularly shines compared to traditional sampling approaches in the 
 following scenarios:
@@
 (SPI).
-Namely, the output of Pigeons.jl is completely invariant to the number of machines 
+Namely, the output of Pigeons.jl is  invariant to the number of machines 
 and/or threads.

### 2024-05-27 paper/paper.tex
use paper folder
fix
Starting to go through paper...

--- 2023-10-30 paper.tex
@@
 
+\usepackage{expl3} % add missing functionality in the old texlive version running on the server 
 \usepackage[autonum, blackhypersetup]{shortex}
 \usepackage{xcolor}
-\usepackage[font=scriptsize]{subcaption}
+%\let\subcaption\undefined % overwrite 'subcaption' from the style file
+%\usepackage[font=scriptsize]{subcaption}
 \usepackage{forest}
@@
 
+\pdfsuppresswarningpagegroup=1 % avoid warning due to pdfs made with photoshop
+
 \begin{document}
@@
 multimodal posteriors arising in Bayesian inference and high-dimensional 
-distributions in statistical mechanics. Pigeons.jl provides simple APIs to 
+distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
 perform such computations single-threaded, multi-threaded,
@@
 
-\input{src/introduction.tex}
-\input{src/examples.tex}
-\input{src/PT.tex}
-\input{src/SPI.tex}
-\input{src/related.tex}
-% \input{src/benchmarks.tex}
-\input{src/conclusion.tex}
+\input{introduction.tex}
+\input{examples.tex}
+\input{PT.tex}
+\input{SPI.tex}
+\input{related.tex}
+\input{conclusion.tex}
 

### 2024-05-28 paper/PT.tex
Pass on the distributed pseudo-code

@@
 it is a method to obtain approximate samples from a distribution. 
-More specifically, it is a class of algorithms for simulating a Markov chain of 
-states that look like draws from the target distribution $\pi$.
 However, with traditional 
@@
 \subsection{Local exploration and communication}
-A struct that is useful for the implementation of PT is the \texttt{Replica}, 
+A central data structure in Pigeons is the \texttt{Replica}, a Julia struct 
 which we will often refer to in the remainder of this section. 
-A single \texttt{Replica} struct stores a \texttt{state} variable and a \texttt{chain} 
+A single \texttt{Replica} stores a \texttt{state} variable and a \texttt{chain} 
 integer, among other entries. 
@@
 \subsection{Distributed implementation}
-A distributed implementation of PT is presented in \cref{alg:distributed_PT}
+A distributed implementation of PT is shown in \cref{alg:distributed_PT}
 and we describe the details of the implementation below.
-We present the algorithm from the perspective that the number of machines available 
+For simplicity, we present the algorithm from the perspective that the number of machines available 
 is equal to the number of \texttt{Replica}s and distributions, $N$. 
@@
 instead of their \texttt{state}s. 
-One difficulty can be encountered in a distributed implementation, which we 
-illustrate with the following simple example. 
+However to achieve this low cost a specialized distributed data structure is 
+needed to maintain a mapping from chain indices to machine identifiers. 
+To motivate it, we start with an illustrative example. 
 Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
@@
 step an index of the array is manipulated by exactly one machine. 
-% With these assumptions we achieve read/write costs that are 
-% near-constant in the number of machines participating.
+Thanks to this, we do not need to resort to global MPI synchronization steps.
 
@@
     Different colours are also used for different machines.
-    Numbers in parantheses represent the \texttt{chain} 
+    Numbers in parentheses represent the \texttt{chain} 
     that is currently being explored by that machine. 
@@
 We use the MPI.jl \cite{byrne2021mpi} package to support communication between machines. 
-In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
-functions \texttt{send()}, \texttt{receive!()}, 
-and \texttt{waitall()}. 
-The function \texttt{send()} has three arguments in our pseudocode: the first argument 
-is the object to be sent, the second is the machine number to which the object should 
-be sent, and finally the third argument is a unique tag for this specific send/receive 
-request at this time step and on this machine, generated by \texttt{tag()}.
-The \texttt{tag()} function can be any function that allows one to uniquely identify 
-a tag for a given send/receive request on a given machine at any given time step $t$.
-The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
-except that the first argument 
-specifies the object to which data should be written directly. Finally, 
-\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
-of machines to complete.
+In  \cref{alg:distributed_PT}, the function \texttt{swap\_stat()} extracts the 
+sufficient statistic required by each partner of a swap. The swap
+sufficient statistic contains two elements:
+(1) a uniformly distributed random number, and (2) the log of the ratio 
+$\pi_c(x_t) / \pi_{c'}(x_t)$, up to a normalization constant which may 
+depend on $c$ and $c'$ but not $x_t$.  
+The function \texttt{transmit()} is used to exchange these statistics 
+between the two partners of a proposed swap. In order to make an 
+accept-reject decision, the two machines use the same uniform sample (using an arbitrary, fixed 
+convention, namely the shared sufficient statistic's uniform from the lowest 
+interacting chain index) to make a common swap decision (where the 
+acceptance ratio is the sum of the two exchanged log ratios, (2)).
+
+%In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
+%functions \texttt{send()}, \texttt{receive!()}, 
+%and \texttt{waitall()}. 
+%The function \texttt{send()} has three arguments in our pseudocode: the first argument 
+%is the object to be sent, the second is the machine number to which the object should 
+%be sent, and finally the third argument is a unique tag for this specific send/receive 
+%request at this time step and on this machine, generated by \texttt{tag()}.
+%The \texttt{tag()} function can be any function that allows one to uniquely identify 
+%a tag for a given send/receive request on a given machine at any given time step $t$.
+%The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
+%except that the first argument 
+%specifies the object to which data should be written directly. Finally, 
+%\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
+%of machines to complete.
 
@@
 In \cref{alg:distributed_PT} we also introduce the functions 
-\texttt{permuted\_get()} and \texttt{permuted\_set!()}. 
+\texttt{permuted\_get()} and \texttt{permuted\_set!()}, which operate 
+on a  \texttt{PermutedDistributedArray} as described  in \cref{sec:permuted_dist_array}.
 Given the current machine's \texttt{dist\_array} and the 
@@
       \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
-      \State $\texttt{partner} \gets \texttt{nothing}$
+      \State $c_t \gets \texttt{chain}$ 
       \If{$\texttt{chain} \in P$}
@@
         \State $\texttt{partner} \gets \texttt{chain}-1$
+      \Else 
+      	\State $\texttt{partner} \gets \texttt{chain}$
       \EndIf
-      \If{$\texttt{partner != nothing}$ \textbf{and} \text{swap is accepted}}
       \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
-          \Comment{retrieve \texttt{dist\_array}[\texttt{partner}]}
-        \State $\texttt{send\_tag} \gets \texttt{tag}(t, \texttt{chain}, j)$
-        \State $\texttt{send}(\texttt{chain}, \texttt{to\_machine}, \texttt{send\_tag})$
-        \State $\texttt{receive\_tag} \gets \texttt{tag}(t, \texttt{chain}, \texttt{to\_machine})$
-        \State $\texttt{receive!}(\texttt{chain}, \texttt{to\_machine}, \texttt{receive\_tag})$
-        \State $\texttt{waitall(requests)}$ 
-          \Comment{wait for MPI requests for this pair of machines to complete}
+      \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
+      \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
+      \If{$\text{swap\_accepted}(\texttt{my\_stat}, \texttt{partner\_stat})$}
+        \State $\texttt{chain} \gets \texttt{partner\_chain}$
+      \EndIf
       \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$ 
-          \Comment{set value of \texttt{dist\_array}[\texttt{chain}] to \texttt{to\_machine}}
-      \EndIf
-      \State $c_t \gets \texttt{chain}$ 
 		\EndFor
@@
     \caption{
-        Weak scaling results for PT in both the distributed (MPI) and parallel settings 
-        with approximate 95\% confidence intervals for the speedup.  
+        Weak scaling results for PT in both the distributed (MPI, {\bf left}) and parallel settings (Julia threads, {\bf right})  
+        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
         The blue dotted line represents a hypothetical speedup equal to the 
@@
         The red dashed line indicates a speedup of one for reference.
-        \textbf{Left:} Distributed setup. 
-        \textbf{Right:} Parallel (shared memory) setup. 
     }
@@
 parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
-might be useful to run parallel tempering with very many chains or 
+might be useful to run parallel tempering with  many chains or 
 to run several instances of parallel tempering at once. In such cases, 

### 2024-05-28 paper/examples.tex
Pass on the distributed pseudo-code

@@
     calling \texttt{Pigeons.setup\_mpi()}.
-    \item In the Julia REPL running in the login node, run:
+    \item In the Julia REPL running in the login node, run:\footnote{In 
+    version 0.4.0 of Pigeons, the function \text{MPI} has been renamed \text{MPIProcesses} 
+    to avoid a clash with the library \text{MPI.jl}.}
 \end{enumerate}
@@
 \texttt{MPI()} to work, it may be necessary to specify dependencies in their 
-function calls.
+function calls. See \url{https://pigeons.run/stable/mpi/} for details.
 

### 2024-05-28 paper/SPI.tex
Releasing token

--- 2023-08-18 src/SPI.tex
@@
 This is achieved by making sure that every value to be added---if addition is our 
-reduction operation---is a leaf node in the reduction tree, irrespective of the 
-number of machines available to perform the reduction. 
+reduction operation---is a leaf node in a reduction tree which is invariant to the 
+number of machines available to perform the computation. 
 For instance, if $N$ values are to be reduced, then the reduction tree would have 

### 2024-05-28 paper/PT.tex
Releasing token
Draft of multi thread + distributed combo

@@
 and we describe the details of the implementation below.
-For simplicity, we present the algorithm from the perspective that the number of machines available 
-is equal to the number of \texttt{Replica}s and distributions, $N$. 
+For pedagogy, we first present the algorithm in the special case where the number of machines available 
+is equal to the number of \texttt{Replica}s and distributions, $N$. We outline 
+how Pigeons removes this restriction in \cref{sec:relax}.
 However, Pigeons.jl also allows for the more general case where the number of 
@@
 
-%In our pseudocode in \cref{alg:distributed_PT} we define the MPI-style 
-%functions \texttt{send()}, \texttt{receive!()}, 
-%and \texttt{waitall()}. 
-%The function \texttt{send()} has three arguments in our pseudocode: the first argument 
-%is the object to be sent, the second is the machine number to which the object should 
-%be sent, and finally the third argument is a unique tag for this specific send/receive 
-%request at this time step and on this machine, generated by \texttt{tag()}.
-%The \texttt{tag()} function can be any function that allows one to uniquely identify 
-%a tag for a given send/receive request on a given machine at any given time step $t$.
-%The function \texttt{receive!()} has the same last two arguments as \texttt{send()} 
-%except that the first argument 
-%specifies the object to which data should be written directly. Finally, 
-%\texttt{waitall(requests)} waits for all initialized MPI \texttt{requests} for a given pair 
-%of machines to complete.
-
 \medskip 
@@
     \For{$t$ {\bf in} 1, 2, \dots, $T$}
-		  \If{$t$ is even} \Comment{choose between even or odd swap}
-		    \State $P \gets \{i: 1 \le i < N, i \text{ is even} \}$
-		  \Else
-		    \State $P \gets \{i: 1 \le i < N, i \text{ is odd} \}$
-		  \EndIf
 
@@
     \State $c_t \gets \texttt{chain}$
-      \If{$\texttt{chain} \in P$}
-        \State $\texttt{partner} \gets \texttt{chain}+1$
-      \ElsIf{$\texttt{chain}-1 \in P$}
-        \State $\texttt{partner} \gets \texttt{chain}-1$
+
+		  \If{$t$ is even} \Comment{choose between even or odd swap}
+		    \State $\texttt{direction} \gets +1$
 		  \Else
-      	\State $\texttt{partner} \gets \texttt{chain}$
+		    \State $\texttt{direction} \gets -1$
 		  \EndIf
+		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$
+
+      \If{$\texttt{partner} = 0$} 
+        \State $\texttt{partner} \gets 1$ \Comment{Encode boundaries as self-interaction}
+      \ElsIf{$\texttt{partner} = N + 1$}
+        \State $\texttt{partner} \gets N$
+      \EndIf
+	    \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
       \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
-      \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
       \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
@@
 \end{algorithm}
+
+\subsubsection{Load balancing}
+\label{sec:relax}
+
+So far we have assumed, for pedagogy, that the number of 
+machines $M$ is equal to the number of chains $N$.
+This restriction is not present in Pigeons, so we now describe how 
+the code base generalizes \cref{alg:distributed_PT}. 
+First, the $N$ replicas are shared between the $M$ machines so that 
+each machine contains either $\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ 
+replicas. Each machine's replicas are called the local replicas. 
+The allocation of replicas to machine is fixed over the whole execution of the 
+algorithm and maintained using the \texttt{LoadBalance} struct. 
+Second, the code performing local exploration is enclosed in a 
+loop over the local replicas, and if the options ``\texttt{multithreaded = true}'' 
+is used in the \texttt{pigeons(...)} call, the exploration loop is performed in a 
+\texttt{@threads for} loop. 
+Finally, the functions used in the communication, 
+\texttt{permuted\_get}, \texttt{permuted\_set} \texttt{transmit}, etc., 
+are vectorized to accommodate that \texttt{chain} and 
+\texttt{partner} are no longer integers but rather sets of integers. 
+In particular, the permuted distributed array is designed so that each 
+machine is responsible for tracking machine allocations for 
+$\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ chains instead of a single one.
 

### 2024-05-29 paper/PT.tex
Fix bug
Tweaks
Minor tweaks

@@
 A central data structure in Pigeons is the \texttt{Replica}, a Julia struct 
-which we will often refer to in the remainder of this section. 
+that we will often refer to in the remainder of this section. 
 A single \texttt{Replica} stores a \texttt{state} variable and a \texttt{chain} 
@@
 For pedagogy, we first present the algorithm in the special case where the number of machines available 
-is equal to the number of \texttt{Replica}s and distributions, $N$. We outline 
-how Pigeons removes this restriction in \cref{sec:relax}.
-However, Pigeons.jl also allows for the more general case where the number of 
+is equal to the number of \texttt{Replica}s and distributions, $N$. 
+However, Pigeons also allows for the more general case where the number of 
 machines is not necessarily equal to $N$.
+We outline how Pigeons achieves this in \cref{sec:relax}.
 
-\subsubsection{The \texttt{PermutedDistributedArray}}
+\subsubsection{Distributed mapping from chains to machines}
 \label{sec:permuted_dist_array}
@@
 instead of their \texttt{state}s. 
-However to achieve this low cost a specialized distributed data structure is 
+However, in order to achieve this low cost, a specialized distributed data structure is 
 needed to maintain a mapping from chain indices to machine identifiers. 
@@
 
-		  \If{$t$ is even} \Comment{choose between even or odd swap}
+		  \If{$\texttt{is\_even}(t) = \texttt{is\_even}(\texttt{chain})$} 
 		    \State $\texttt{direction} \gets +1$
@@
 Second, the code performing local exploration is enclosed in a 
-loop over the local replicas, and if the options ``\texttt{multithreaded = true}'' 
+loop over the local replicas, and if the option ``\texttt{multithreaded = true}'' 
 is used in the \texttt{pigeons(...)} call, the exploration loop is performed in a 
@@
 Finally, the functions used in the communication, 
-\texttt{permuted\_get}, \texttt{permuted\_set} \texttt{transmit}, etc., 
+\texttt{permuted\_get}, \texttt{permuted\_set!}, \texttt{transmit}, etc., 
 are vectorized to accommodate that \texttt{chain} and 
@@
     \caption{
-        Weak scaling results for PT in both the distributed (MPI, {\bf left}) and parallel settings (Julia threads, {\bf right})  
+        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
+        and parallel (Julia threads, {\bf right}) settings 
         with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   

### 2024-05-29 paper/SPI.tex
Minor tweaks

--- 2023-08-18 src/SPI.tex
@@
 This is achieved by making sure that every value to be added---if addition is our 
-reduction operation---is a leaf node in the reduction tree, irrespective of the 
-number of machines available to perform the reduction. 
+reduction operation---is a leaf node in a reduction tree that is invariant to the 
+number of machines available to perform the computation. 
 For instance, if $N$ values are to be reduced, then the reduction tree would have 

### 2024-05-29 paper/introduction.tex
Tweaks
Minor tweaks

@@
 \url{https://github.com/Julia-Tempering/Pigeons.jl}. 
-Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
+Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
 enables users to sample efficiently from challenging probability distributions 
@@
 
-More in-depth documentation and additional examples can be found at: \\ \url{https://pigeons.run/}.
+More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.
 
@@
 We describe the class of problems that can be approached using Pigeons.jl.
-Let $\pi(x)$ denote a probability density
-\footnote{This density may also be a probability mass function 
+Let $\pi(x)$ denote a probability density\footnote{This density may 
+also be a probability mass function 
 (discrete variables). We also allow for a combination of discrete/continuous variables.} 
@@
     \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
-    the normalization constant $Z$. 
+    the normalizing constant $Z$. 
     Many existing tools focus on the former and struggle or fail to do the latter.

### 2024-05-29 paper/paper.tex
Minor tweaks

@@
 \begin{abstract}
-
 We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
@@
 a distributed and randomized algorithm that satisfies strong parallelism invariance.
-
 \end{abstract}

### 2025-05-06 workshop/abstract.tex
first attempt @ icml workshop (broken)

--- 2024-05-29 paper/paper.tex
@@
-% JuliaCon proceedings template
-\documentclass{juliacon}
-\setcounter{page}{1}
-
-\usepackage{expl3} % add missing functionality in the old texlive version running on the server 
-\usepackage[autonum, blackhypersetup]{shortex}
-\usepackage{xcolor}
-%\let\subcaption\undefined % overwrite 'subcaption' from the style file
-%\usepackage[font=scriptsize]{subcaption}
-\usepackage{forest}
-
-\newcommand{\NS}[1]{\textcolor{red}{#1}}
-\def\bfx{\mathbf{x}}
-\def\bfX{\mathbf{X}}
-
-\definecolor{julia1}{rgb}{0, 0.60, 0.98}
-\definecolor{julia2}{rgb}{0.89, 0.43, 0.28}
-\definecolor{julia3}{rgb}{0.24, 0.64, 0.30}
-\definecolor{julia4}{rgb}{0.76, 0.44, 0.82}
-\definecolor{julia5}{rgb}{0.67, 0.55, 0.09}
-\definecolor{julia6}{HTML}{00aaae}
-\definecolor{julia7}{HTML}{ed5e93}
-\definecolor{julia8}{HTML}{c68225}
-
-\pdfsuppresswarningpagegroup=1 % avoid warning due to pdfs made with photoshop
-
-\begin{document}
-
-\input{header}
-
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\input{introduction.tex}
-\input{examples.tex}
-\input{PT.tex}
-\input{SPI.tex}
-\input{related.tex}
-\input{conclusion.tex}
-
-\section{Acknowledgements}
-NS acknowledges the support of a Vanier Canada Graduate Scholarship. 
-PT acknowledges the support of the Black Hole Initiative at Harvard University, 
-which is funded by grants from the John Templeton Foundation and the 
-Gordon and Betty Moore Foundation to Harvard University.
-PT also acknowledges support by the National Science Foundation grants AST-1935980 
-and AST-2034306 and the Gordon and Betty Moore Foundation (GBMF-10423).
-SS acknowledges the support of EPSRC grant EP/R034710/1 CoSines.
-ABC and TC acknowledge the support of an NSERC Discovery Grant.
-We also acknowledge use of the ARC Sockeye computing platform from the
-University of British Columbia, as well as cloud computing resources provided 
-by Oracle.
-
-\input{bib.tex}
-
-\end{document}
-
-% Inspired by the International Journal of Computer Applications template

### 2025-05-06 workshop/ack.tex
first attempt @ icml workshop (broken)

--- 2023-10-30 paper.tex
@@
-% JuliaCon proceedings template
-\documentclass{juliacon}
-\setcounter{page}{1}
-
-\usepackage[autonum, blackhypersetup]{shortex}
-\usepackage{xcolor}
-\usepackage[font=scriptsize]{subcaption}
-\usepackage{forest}
-
-\newcommand{\NS}[1]{\textcolor{red}{#1}}
-\def\bfx{\mathbf{x}}
-\def\bfX{\mathbf{X}}
-
-\definecolor{julia1}{rgb}{0, 0.60, 0.98}
-\definecolor{julia2}{rgb}{0.89, 0.43, 0.28}
-\definecolor{julia3}{rgb}{0.24, 0.64, 0.30}
-\definecolor{julia4}{rgb}{0.76, 0.44, 0.82}
-\definecolor{julia5}{rgb}{0.67, 0.55, 0.09}
-\definecolor{julia6}{HTML}{00aaae}
-\definecolor{julia7}{HTML}{ed5e93}
-\definecolor{julia8}{HTML}{c68225}
-
-\begin{document}
-
-\input{header}
-
-\maketitle
-
-\begin{abstract}
-
-We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
-obtain samples from complicated probability distributions, such as 
-multimodal posteriors arising in Bayesian inference and high-dimensional 
-distributions in statistical mechanics. Pigeons.jl provides simple APIs to 
-perform such computations single-threaded, multi-threaded,
-and/or distributed over thousands of MPI-communicating machines.
-In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
-output for a given seed is identical irrespective of the number of threads and processes, 
-which is crucial for scientific reproducibility and software validation. 
-We describe the key features of Pigeons.jl and the approach taken to implement 
-a distributed and randomized algorithm that satisfies strong parallelism invariance.
-
-\end{abstract}
-
-\input{src/introduction.tex}
-\input{src/examples.tex}
-\input{src/PT.tex}
-\input{src/SPI.tex}
-\input{src/related.tex}
-% \input{src/benchmarks.tex}
-\input{src/conclusion.tex}
-
 \section{Acknowledgements}
@@
 by Oracle.
-
-\input{bib.tex}
-
-\end{document}
-
-% Inspired by the International Journal of Computer Applications template

### 2025-05-06 workshop/paper.tex
first attempt @ icml workshop (broken)

--- 2024-05-27 paper/paper.tex
@@
-% JuliaCon proceedings template
-\documentclass{juliacon}
+\documentclass{article}
+
+\usepackage[accepted]{icml2025}
+\usepackage{listings}
+\usepackage{jlcode}
+\usepackage{natbib}
 \setcounter{page}{1}
@@
 
+% The \icmltitle you define below is probably too long as a header.
+% Therefore, a short form for the running title is supplied here:
+\icmltitlerunning{Pigeons.jl: Distributed sampling from intractable distributions}
+
 \begin{document}
 
-\input{header}
+\twocolumn[
+\icmltitle{Pigeons.jl: Distributed sampling from intractable distributions}
 
-\maketitle
+\icmlsetsymbol{equal}{*}
 
-\begin{abstract}
+\begin{icmlauthorlist}
+\icmlauthor{Nikola Surjanovic}{ubc}
+\icmlauthor{Miguel Biron-Lattes}{sfu}
+\icmlauthor{Paul Tiede}{harvard}
+\icmlauthor{Saifuddin Syed}{oxford}
+\icmlauthor{Alexandre Bouchard-C{\^o}t{\'e}}{ubc}
+\icmlauthor{Trevor Campbell}{ubc}
+\end{icmlauthorlist}
 
-We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
-obtain samples from complicated probability distributions, such as 
-multimodal posteriors arising in Bayesian inference and high-dimensional 
-distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
-perform such computations single-threaded, multi-threaded,
-and/or distributed over thousands of MPI-communicating machines.
-In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
-output for a given seed is identical irrespective of the number of threads and processes, 
-which is crucial for scientific reproducibility and software validation. 
-We describe the key features of Pigeons.jl and the approach taken to implement 
-a distributed and randomized algorithm that satisfies strong parallelism invariance.
+\icmlaffiliation{harvard}{Center for Astrophysics, Harvard University, United States}
+\icmlaffiliation{oxford}{Department of Statistics, University of Oxford, United Kingdom}
+\icmlaffiliation{sfu}{Department of Statistics and Actuarial Sciences, Simon Fraser University, Canada}
+\icmlaffiliation{ubc}{Department of Statistics, University of British Columbia, Canada}
 
-\end{abstract}
+\icmlcorrespondingauthor{Miguel Biron-Lattes}{mbironla@sfu.ca}
+\icmlkeywords{Distributed computation, Bayesian inference, Markov chain Monte Carlo, Parallelism Invariance, Message Passing Interface, Multi-Threading}
 
+\vskip 0.3in
+]
+
+\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+
+\input{abstract}
 \input{introduction.tex}
@@
 \input{conclusion.tex}
+\input{ack}
 
-\section{Acknowledgements}
-NS acknowledges the support of a Vanier Canada Graduate Scholarship. 
-PT acknowledges the support of the Black Hole Initiative at Harvard University, 
-which is funded by grants from the John Templeton Foundation and the 
-Gordon and Betty Moore Foundation to Harvard University.
-PT also acknowledges support by the National Science Foundation grants AST-1935980 
-and AST-2034306 and the Gordon and Betty Moore Foundation (GBMF-10423).
-SS acknowledges the support of EPSRC grant EP/R034710/1 CoSines.
-ABC and TC acknowledge the support of an NSERC Discovery Grant.
-We also acknowledge use of the ARC Sockeye computing platform from the
-University of British Columbia, as well as cloud computing resources provided 
-by Oracle.
-
-\input{bib.tex}
+\bibliographystyle{icml2025}
+\bibliography{../paper/ref.bib}
 

### 2025-05-06 workshop/PT.tex
some fixes
formatting

--- 2024-05-29 paper/PT.tex
@@
 
-\medskip 
 In this section we assume a basic understanding of 
@@
 
-\medskip
 To resolve this issue, PT constructs a sequence of $N$ distributions,  
@@
 
-\medskip
 We now turn to explain how this path of distributions can be used to enhance 
@@
 
-\medskip 
 In the local exploration phase,
@@
 
-\medskip 
 In the communication phase, PT proposes swaps between pairs of replicas. 
@@
 
-\medskip 
 To resolve this issue, we introduce a special data structure called a 
@@
 
-\medskip 
 In \cref{alg:distributed_PT} we also introduce the functions 
@@
 \begin{algorithm}[t]
+\footnotesize
 	\begin{algorithmic}[1]
@@
 
-\medskip
 The weak scaling results for both the distributed memory parallelism (DMP) and 

### 2025-05-06 workshop/SPI.tex
formatting

--- 2024-05-29 paper/SPI.tex
@@
 
-\medskip 
 We have identified two factors that can cause violations of our previously-defined 
@@
 
-\medskip 
 To illustrate why distributed reduction with floating point values can violate 
@@
 
-\medskip 
 To avoid the issue of non-associativity of 
@@
 
-\medskip 
 To motivate splittable random streams, consider the following toy example 
@@
 
-\medskip 
 A first step to achieve this is to associate one random number 

### 2025-05-06 workshop/examples.tex
formatting

--- 2024-05-28 paper/examples.tex
@@
 
-\medskip
 We begin by installing the latest official release of Pigeons.jl: 
@@
 
-\medskip 
 In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
@@
 
-\medskip 
 Suppose that we perform $n=100,000$ coin tosses and observe 
@@
 
-\medskip 
 To obtain posterior densities and trace-plots, we first 
@@
 
-\medskip 
 To obtain pair plots, we add the 
@@
 
-\medskip 
 As a side-product of PT, we automatically obtain an approximation to the natural 
@@
 
-\medskip 
 First, we make sure that we set \texttt{checkpoint = true}, which saves a 
@@
 
-\medskip 
 The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 

### 2025-05-06 workshop/introduction.tex
formatting

--- 2024-05-29 paper/introduction.tex
@@
 
-\medskip
 Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
@@
 
-\medskip 
 The output of Pigeons.jl can be used for two main tasks:
@@
 
-\medskip 
 The general Julia developer will be interested in: 

### 2025-05-06 workshop/related.tex
formatting

--- 2023-10-31 paper/related.tex
@@
 
-\medskip 
 These software packages often provide two key user-facing components: (1) a 
@@
 
-\medskip
 Inference engines available in existing software packages are varied in their 
@@
 
-\medskip
 There is also a vast literature on distributed and parallel 

### 2025-05-07 workshop/SPI.tex
almost 4 pages but needs work

@@
 imitating the behaviour that would be present if there were $N$ machines available.
-\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
-illustrate the reduction procedure for eight and two machines, respectively.
+\cref{fig:reduction_tree_colour} illustrates the reduction procedure for eight and 
+two machines.
 
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
   \centering
+  \begin{minipage}{0.5\linewidth}
   \begin{forest}
@@
   \end{forest}
-  \caption{Addition of eight floating point numbers across eight machines with a 
-  guarantee on SPI. Each machine is represented by a different colour. 
-  The final result can be represented 
-  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
-  \label{fig:reduction_tree_colour_8}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
+  \end{minipage}%
+  \begin{minipage}{0.5\linewidth}
   \begin{forest}
@@
   \end{forest}
-  \caption{Addition of eight floating point numbers across two machines with a 
-  guarantee on SPI. Each machine is represented by a different colour. 
-  The final result can be represented 
-  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$,
-  which is the same as that given by eight machines in \cref{fig:reduction_tree_colour_8}.}
-  \label{fig:reduction_tree_colour_2}
-\end{figure}
+  \end{minipage}
+  \caption{Addition of eight floating point numbers across eight (\textbf{left}) 
+  and two (\textbf{right}) machines with a guarantee on SPI. Each machine is 
+  represented by a different colour. In both cases, the final result can be 
+  represented by the same expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
+  \label{fig:reduction_tree_colour}
+\end{figure*}
 

### 2025-05-07 workshop/abstract.tex
almost 4 pages but needs work

@@
 \begin{abstract}
-We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
-obtain samples from complicated probability distributions, such as 
+Sampling from complicated probability distributions---such as 
 multimodal posteriors arising in Bayesian inference and high-dimensional 
-distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
-perform such computations single-threaded, multi-threaded,
-and/or distributed over thousands of MPI-communicating machines.
-In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
-output for a given seed is identical irrespective of the number of threads and processes, 
-which is crucial for scientific reproducibility and software validation. 
-We describe the key features of Pigeons.jl and the approach taken to implement 
-a distributed and randomized algorithm that satisfies strong parallelism invariance.
+distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
+a simple user interface to perform such computations single-threaded, multi-threaded,
+and/or distributed over thousands of MPI-communicating machines. In this paper,
+we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
+a guarantee that the output for a given seed is identical irrespective of the number 
+of threads and processes. This feature is crucial for scientific reproducibility and 
+software validation. We describe the key features of Pigeons.jl that enable a 
+distributed and randomized algorithm satisfying SPI.
 \end{abstract}

### 2025-05-07 workshop/examples.tex
almost 4 pages but needs work

@@
     \centering 
-    \includegraphics[width=0.5\textwidth]{../img/coinflip_posterior.pdf}
+    \includegraphics[width=0.8\linewidth]{../img/coinflip_posterior.pdf}
     \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
@@
 access the results, as described in the following section. 
-The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
-and explained in the next section. 
 For purposes of comparison, we also run a traditional 
 (single-chain Markov chain Monte Carlo) method.
-
-\begin{figure*}[t]
-    \centering
-    \begin{lstlisting}
-            #scans    restarts        Λ      time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
-
-                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
-                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
-                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
-                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
-                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
-                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
-                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
-                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
-                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
-           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
-
-    \end{lstlisting}
-    \caption{Standard output provided by Pigeons.jl. 
-    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
-    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
-    such as the amount of memory allocations per round (in bytes), time (in seconds), and estimates 
-    of the log of the normalization constants. 
-    The output is described in greater detail in \cref{sec:PT_diagnostics}
-    and has been modified to exclude columns that are not described in the paper.}
-    \label{fig:standard_out}
-\end{figure*}
 
@@
 See \url{https://pigeons.run/stable/input-overview/}.
-
-\subsection{Outputs}
-Pigeons.jl provides many useful types of output, such as: plots of samples from 
-the distribution, estimates of normalization constants, summary statistics of 
-the target distribution, and various other diagnostics. 
-We describe several examples of possible output below.
-
-\subsubsection{Standard output}
-An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
-Each row of the table in the output indicates a new tuning round in parallel tempering, 
-with the \texttt{\#scans} column indicating the number of scans/samples in 
-that tuning round. 
-During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
-tuning parameters. Other outputs include: 
-\begin{itemize}
-    \item \texttt{restarts}: a higher number is better. 
-    Informally, PT propagates samples from an easy-to-sample distribution (the reference) 
-    to the more difficult target distribution. 
-    A tempered restart happens when a sample from the reference successfully 
-    percolates to the target. (See the subsequent sections for a more detailed description 
-    of parallel tempering.) 
-    When the reference supports i.i.d.~sampling, tempered restarts 
-    can enable large jumps in the state space.
-
-    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
-    which measures the inherent difficulty of the sampling problem.
-    A rule of thumb to configure the number of PT chains is also 
-    given by \cite{syed2021nrpt}, where they suggest that stable performance 
-    should be achieved when the number of chains is set to roughly 2$\Lambda$.
-    See \cref{sec:PT_diagnostics} for more information.
-
-    \item \texttt{time} and \texttt{allc}: the time (in seconds) and number of 
-    allocations (in bytes) used in each round.
-
-    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
-    logarithm of the ratio of normalization constants between the target and the reference. 
-    In many cases, $Z_0 = 1$.
-
-    \item \texttt{min($\alpha$)} and \texttt{mean($\alpha$)}: minimum and average swap 
-    acceptance rates during the communication phase across the PT chains.
-    See \cref{sec:PT} for a description of PT communication.
-\end{itemize}
-
-\subsubsection{Plots}
-It is straightforward to obtain plots of samples from the target distribution, 
-such as trace-plots, pairwise plots, and density plots of the marginals.
-
-To obtain posterior densities and trace-plots, we first 
-make sure that we have the third-party 
-MCMCChains.jl\footnote{\url{https://github.com/TuringLang/MCMCChains.jl}}, 
-StatsPlots.jl\footnote{\url{https://github.com/JuliaPlots/StatsPlots.jl}}, and 
-PlotlyJS.jl\footnote{\url{https://github.com/JuliaPlots/PlotlyJS.jl}} 
-packages installed via
-\begin{lstlisting}[language=Julia]
-Pkg.add("MCMCChains", "StatsPlots", "PlotlyJS")
-\end{lstlisting}
-
-With the \texttt{pt} output object from before for our non-identifiable coinflip model, 
-we can run the following:
-\begin{lstlisting}[language=Julia]
-using MCMCChains, StatsPlots, PlotlyJS
-plotlyjs()
-samples = Chains(
-    sample_array(pt), variable_names(pt))
-my_plot = StatsPlots.plot(samples)
-display(my_plot)
-\end{lstlisting}
-The output of the above code chunk is an interactive plot that can be zoomed in or out
-and exported as an HTML webpage. 
-A modified static version of the output for the first parameter, $p_1$, 
-is displayed in the top panel of \cref{fig:trace_density_plots},
-along with a comparison to the output from a single-chain algorithm in the bottom 
-panel of the same figure.
-
-\begin{figure*}[t]
-    \centering
-    \begin{minipage}{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
-    \end{minipage}
-    \begin{minipage}{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
-    \end{minipage}
-    \caption{
-        Trace-plots for the first parameter, $p_1$, in the 
-        non-identifiable coinflip Turing model.
-        \textbf{Left:} Samples from Pigeons.jl using PT with 10 chains. Note that the 
-        trace-plot indicates fast mixing/exploration across the state space.
-        \textbf{Right:} Single-chain Markov chain Monte Carlo. Note that the 
-        trace-plot explores the state space much more slowly when we do not use PT.
-    }
-    \label{fig:trace_density_plots}
-\end{figure*}
-
-To obtain pair plots, we add the 
-PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
-and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
-packages:
-\begin{lstlisting}[language=Julia]
-Pkg.add("PairPlots", "CairoMakie")
-\end{lstlisting}
-and then run
-\begin{lstlisting}[language=Julia]
-using PairPlots, CairoMakie
-my_plot = PairPlots.pairplot(samples) 
-display(my_plot)
-\end{lstlisting}
-The output of the above code chunk with an increased number of samples 
-is displayed in \cref{fig:coinflip_posterior}.
-\footnote{The code chunk above requires Julia 1.9.}
-
-\subsubsection{Estimate of normalization constant}
-The (typically unknown) constant $Z$ in \cref{eq:normalizing_constant} is 
-referred to as the \emph{normalization constant}. 
-In many applications, it is useful to approximate this constant. 
-For example, in Bayesian statistics, this corresponds to the 
-marginal likelihood and can be used for model selection.
-
-As a side-product of PT, we automatically obtain an approximation to the natural 
-logarithm of the normalization constant. This is done automatically using the 
-stepping stone estimator \cite{xie2011improving}.
-The estimate can be accessed using
-\begin{lstlisting}[language=Julia]
-stepping_stone(pt)
-\end{lstlisting}
-In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
-for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$.
-Note that this is very close to the output provided in \cref{fig:standard_out}.
-
-\subsubsection{Online statistics}
-\label{sec:online_stats}
-Pigeons has facilities to support cases requiring large memory. For instance, we
-allow for the computation of online statistics, as well as off-memory sample storage.
-Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
-we can output some basic summary statistics of the marginals of our target distribution. 
-For instance, it is straightforward to estimate the mean and variance of each 
-of the marginals of the target with 
-\begin{lstlisting}[language=Julia]
-using Statistics
-mean(pt); var(pt)
-\end{lstlisting}
-Other constant-memory statistic accumulators are made available in the 
-OnlineStats.jl \cite{day2020onlinestats} package. 
-To add additional constant-memory statistic accumulators, we 
-can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
-online documentation. For instance, we can also compute constant-memory estimates 
-of extrema of our distribution.
 

### 2025-05-07 workshop/introduction.tex
almost 4 pages but needs work

@@
 concurrently on a compute cluster (Sockeye) at the University of British Columbia.
-Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
-the output for a given seed is \emph{identical} irrespective 
-of the number of threads or processes. 
-Such a level of reproducibility is rare in distributed software but of 
-great use for the purposes of debugging in the context of sampling algorithms, 
-which produce stochastic output.
+Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
+wherein the output for a given seed is \emph{identical} irrespective of the number 
+of threads or processes. Such a level of reproducibility is rare in distributed 
+software but of great use for the purposes of debugging in the context of sampling 
+algorithms, which produce stochastic output.
 Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
@@
 
-\subsection{Latest documentation}
-
-More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.
-
-\subsection{Problem formulation}
-We describe the class of problems that can be approached using Pigeons.jl.
-Let $\pi(x)$ denote a probability density\footnote{This density may 
-also be a probability mass function 
+\section{Overview of Pigeons.jl}
+A typical instance of the class of problems that Pigeons.jl tackles can be described by
+a probability density\footnote{This density may also be a probability mass function 
 (discrete variables). We also allow for a combination of discrete/continuous variables.} 
-called the \emph{target}. 
-In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
-known only up to a normalizing constant, 
+$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
+density $\pi$ is typically known only up to a normalizing constant, 
 \[
@@
 \end{itemize}
-
-\subsection{What is of interest to the general Julia developer?}
-Ensuring code correctness at the intersection of randomized, parallel, and 
-distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
-based on a principle that we refer to as \textit{strong parallelism invariance}
-(SPI).
-Namely, the output of Pigeons.jl is  invariant to the number of machines 
-and/or threads.
-Without explicitly keeping SPI in mind during software construction, 
-the (random) output of the algorithm is only guaranteed to have the same distribution.
-This is a much weaker guarantee that, in particular, makes debugging difficult.
-However, with our notion of SPI we make debugging and 
-software validation considerably easier. This is because the developer can 
-first focus on the fully serial randomized algorithm, and then use it as an 
-easy-to-compare gold-standard reference for parallel/distributed implementations. 
-This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
-testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
-incurs additional false negatives due to statistical error.
-
-The general Julia developer will be interested in: 
-\begin{itemize}
-    \item The main causes of a violation of strong parallelism invariance that we have identified 
-    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
-    them in Pigeons.jl.
-
-    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
-    package that was developed by our team 
-    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
-\end{itemize}

### 2025-05-07 workshop/paper.tex
some fixes
almost 4 pages but needs work

@@
 
-\usepackage[accepted]{icml2025}
-\usepackage{listings}
-\usepackage{jlcode}
-\usepackage{natbib}
-\setcounter{page}{1}
-
-\usepackage{expl3} % add missing functionality in the old texlive version running on the server 
-\usepackage[autonum, blackhypersetup]{shortex}
-\usepackage{xcolor}
-%\let\subcaption\undefined % overwrite 'subcaption' from the style file
-%\usepackage[font=scriptsize]{subcaption}
-\usepackage{forest}
-
-\newcommand{\NS}[1]{\textcolor{red}{#1}}
-\def\bfx{\mathbf{x}}
-\def\bfX{\mathbf{X}}
-
-\definecolor{julia1}{rgb}{0, 0.60, 0.98}
-\definecolor{julia2}{rgb}{0.89, 0.43, 0.28}
-\definecolor{julia3}{rgb}{0.24, 0.64, 0.30}
-\definecolor{julia4}{rgb}{0.76, 0.44, 0.82}
-\definecolor{julia5}{rgb}{0.67, 0.55, 0.09}
-\definecolor{julia6}{HTML}{00aaae}
-\definecolor{julia7}{HTML}{ed5e93}
-\definecolor{julia8}{HTML}{c68225}
-
-\pdfsuppresswarningpagegroup=1 % avoid warning due to pdfs made with photoshop
+\input{macros}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Pigeons.jl: Distributed sampling from intractable distributions}
+\newcommand{\thetitle}{Reproducible sampling from intractable distributions with Pigeons.jl}
+\icmltitlerunning{\thetitle}
 
@@
 \twocolumn[
-\icmltitle{Pigeons.jl: Distributed sampling from intractable distributions}
+\icmltitle{\thetitle}
 
@@
 \input{introduction.tex}
-\input{examples.tex}
-\input{PT.tex}
 \input{SPI.tex}
 \input{related.tex}
-\input{conclusion.tex}
-\input{ack}
+%\input{conclusion.tex}
+%\input{ack}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% APPENDIX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\appendix
+\onecolumn
+
+\input{examples}
+\input{PT}
+
 \end{document}

### 2025-05-13 workshop/introduction.tex
fix

@@
 single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
-We have stress-tested Pigeons successfully with up to 1000 machines running 
+We have stress-tested Pigeons successfully with up to 1000 processes running 
 concurrently on a compute cluster (Sockeye) at the University of British Columbia.

### 2025-05-13 workshop/PT.tex
almost 4 pages but needs work
TPG section + reorder

@@
 \label{sec:PT}
+
+\begin{algorithm}[t]
+	\begin{algorithmic}[1]
+    \Require Initial state $x_0$, sequence of distributions $\{\pi_i\}_{i=1}^N$, 
+      number of iterations $T$, machine number $j$
+    \State $\texttt{chain} \gets j$ \Comment{current \texttt{chain} number}
+    \State $\texttt{dist\_array}[j] \gets j$ \Comment{\texttt{chain} $j$ is on machine $j$}
+
+    \For{$t$ {\bf in} 1, 2, \dots, $T$}
+
+     \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
+    \State $c_t \gets \texttt{chain}$
+
+		  \If{$\texttt{is\_even}(t) = \texttt{is\_even}(\texttt{chain})$} 
+		    \State $\texttt{direction} \gets +1$
+		  \Else
+		    \State $\texttt{direction} \gets -1$
+		  \EndIf
+		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$
+
+      \If{$\texttt{partner} = 0$} 
+        \State $\texttt{partner} \gets 1$ \Comment{Encode boundaries as self-interaction}
+      \ElsIf{$\texttt{partner} = N + 1$}
+        \State $\texttt{partner} \gets N$
+      \EndIf
+	    \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
+      \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
+      \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
+      \If{$\text{swap\_accepted}(\texttt{my\_stat}, \texttt{partner\_stat})$}
+        \State $\texttt{chain} \gets \texttt{partner\_chain}$
+      \EndIf
+      \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$ 
+		\EndFor
+    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
+	\end{algorithmic}
+  \caption{Distributed PT on machine $j$ (one replica per machine)}
+  \label{alg:distributed_PT}
+\end{algorithm}
+
 Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
@@
 its distributed implementation.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.25\textwidth]{../img/bimodal.pdf}
+    \caption{A simple bimodal distribution from which traditional 
+    MCMC methods may struggle to obtain samples. Blue lines display output 
+    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
+    The sampler in this figure is visibly trapped in one of the two modes.}
+    \label{fig:bimodal}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.3\textwidth]{../img/bimodal.pdf}
-    \caption{A simple bimodal distribution from which traditional 
-    MCMC methods may struggle to obtain samples. Blue lines display output 
-    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
-    The sampler in this figure is visibly trapped in one of the two modes.}
-    \label{fig:bimodal}
-\end{figure}
-
 To resolve this issue, PT constructs a sequence of $N$ distributions,  
@@
 is shown in Figure~\ref{fig:path}.
-
-\begin{figure}[t]
-    \centering
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
-      \caption*{$\pi_1$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
-      \caption*{$\pi_2$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
-      \caption*{$\pi_3$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
-      \caption*{$\pi_4$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
-      \caption*{$\pi_5$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
-      \caption*{$\pi_6$}
-    \end{minipage}
-    \caption{Heatmaps of six distributions lying on an annealing path 
-    from a unimodal reference distribution  
-    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
-    $\pi_N$, which is in this case the bimodal distribution from 
-    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
-    are not directly comparable because the densities of intermediate distributions 
-    are not normalized.)}
-    \label{fig:path}
-\end{figure}
 
@@
 incurring a potentially very high communication cost for large values of $d$.
+
+% NB: this is placed here just to make it appear in a different page
+\begin{figure}[t]
+    \centering
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
+      \caption*{$\pi_1$}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
+      \caption*{$\pi_2$}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
+      \caption*{$\pi_3$}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
+      \caption*{$\pi_4$}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
+      \caption*{$\pi_5$}
+    \end{minipage}
+    \begin{minipage}{0.15\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
+      \caption*{$\pi_6$}
+    \end{minipage}
+    \caption{Heatmaps of six distributions lying on an annealing path 
+    from a unimodal reference distribution  
+    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
+    $\pi_N$, which is in this case the bimodal distribution from 
+    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
+    are not directly comparable because the densities of intermediate distributions 
+    are not normalized.)}
+    \label{fig:path}
+\end{figure}
 
@@
 
-\begin{algorithm}[t]
-\footnotesize
-	\begin{algorithmic}[1]
-    \Require Initial state $x_0$, sequence of distributions $\{\pi_i\}_{i=1}^N$, 
-      number of iterations $T$, machine number $j$
-    \State $\texttt{chain} \gets j$ \Comment{current \texttt{chain} number}
-    \State $\texttt{dist\_array}[j] \gets j$ \Comment{\texttt{chain} $j$ is on machine $j$}
-
-    \For{$t$ {\bf in} 1, 2, \dots, $T$}
-
-     \State $x_t \gets \texttt{local\_exploration}(\pi_\texttt{chain}, x_{t-1})$
-    \State $c_t \gets \texttt{chain}$
-
-		  \If{$\texttt{is\_even}(t) = \texttt{is\_even}(\texttt{chain})$} 
-		    \State $\texttt{direction} \gets +1$
-		  \Else
-		    \State $\texttt{direction} \gets -1$
-		  \EndIf
-		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$
-
-      \If{$\texttt{partner} = 0$} 
-        \State $\texttt{partner} \gets 1$ \Comment{Encode boundaries as self-interaction}
-      \ElsIf{$\texttt{partner} = N + 1$}
-        \State $\texttt{partner} \gets N$
-      \EndIf
-	    \State $\texttt{my\_stat} \gets \texttt{swap\_stat}(x_t, $\texttt{chain}, $\texttt{partner} )$  
-      \State $\texttt{to\_machine} \gets  \texttt{permuted\_get}(\texttt{dist\_array}, \texttt{partner})$
-      \State $\texttt{partner\_stat} \gets \texttt{transmit}(\texttt{my\_stat}, \texttt{to\_machine})$
-      \If{$\text{swap\_accepted}(\texttt{my\_stat}, \texttt{partner\_stat})$}
-        \State $\texttt{chain} \gets \texttt{partner\_chain}$
-      \EndIf
-      \State $\texttt{permuted\_set!}(\texttt{dist\_array}, \texttt{chain}, j)$ 
-		\EndFor
-    \State \Return $\{(x_t, c_t)\}_{t=1}^T$
-	\end{algorithmic}
-  \caption{Distributed PT on machine $j$ (one replica per machine)}
-  \label{alg:distributed_PT}
-\end{algorithm}
-
 \subsubsection{Load balancing}
@@
     \centering
-    \begin{minipage}{0.4\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/speedup_distributed.pdf}
-    \end{minipage}
-    \begin{minipage}{0.4\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/speedup_parallel.pdf}
-    \end{minipage}
+    \includegraphics[width=0.33\textwidth]{../img/speedup_distributed.pdf}
+    \hspace{4em}
+    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
     \caption{

### 2025-05-13 workshop/TPG.tex
TPG section + reorder

\section{Towards tempered Particle Gibbs for Turing-complete PPLs}

Particle Gibbs (PG, \citealp{andrieu2010particle}) is an MCMC
algorithm that has been shown \citep{wood2014new} to be a natural sampler for
arbitrary programs written in Turing-complete probabilistic programming 
languages (PPLs). For example, the Julia package Turing.jl \citep{ge2018turing}
provides a PG sampler that can handle any program written in DynamicPPL
\citep{tarek2020dynamicppl}, its underlying PPL.
However, this generality comes at the expense of poor performance in moderately
complex models. This is due to the fact that PG is, essentially, a smart 
algorithm for selecting samples from the prior that are close to the posterior.
When the prior is far from the posterior, this process can take a long time to
produce a sample of acceptable quality.

But this is where Pigeons can help: within a Parallel Tempering (PT) run, PG can be
used to explore each distribution in an annealing sequence that interpolates between
the prior and posterior. We expect PG to work more effectively on tempered 
distributions closer to the prior. Such samples can then be transported towards 
the posterior chain via the PT process.
In order to construct this sequence of distributions for an arbitrary program in
a Turing-complete PPL, it suffices to inject an inverse temperature parameter 
$\beta\in[0,1]$ in the call to compute the log density of every \texttt{observe} 
statement in the program. Leveraging multiple dispatch in Julia, we have
successfully applied this technique to DynamicPPL programs in our package
TPGExplorers.jl\footnote{\url{https://github.com/Julia-Tempering/TPGExplorers.jl}}.

### 2025-05-13 workshop/PT.tex
compress

@@
 Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
-in \cite{syed2021nrpt}, which we outline in Algorithm~\ref{alg:distributed_PT}.
+in \cite{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
 This section gives both a brief overview of PT and some details of 
@@
 PT is a method that aims to address these issues.
-For a more in-depth review of PT, we refer readers to \cite{surjanovic2022vpt}.
+For a more in-depth review of PT, we refer readers to \citet{surjanovic2022vpt}.
 
@@
 For example, in a bimodal example such as the one illustrated in 
-Figure~\ref{fig:bimodal},
+\cref{fig:bimodal},
 traditional methods might remain in one of the two modes for an extremely 
@@
 of such a sequence of distributions, referred to as an \textit{annealing path},
-is shown in Figure~\ref{fig:path}.
+is shown in \cref{fig:path}.
 
@@
 for the discovery of new regions of the space of the target distribution such as 
-the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.
+the top-right mode of the distribution presented in \cref{fig:bimodal}.
 
@@
     $\pi_N$, which is in this case the bimodal distribution from 
-    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
+    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
     are not directly comparable because the densities of intermediate distributions 

### 2025-05-13 workshop/introduction.tex
compress

@@
 
-Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
-\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
-Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
-enables users to sample efficiently from challenging probability distributions 
-and solve integration problems by 
-implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
-that leverage distributed computation. Its simple user interface facilitates performing such computations 
+Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
+enables users to sample efficiently from challenging probability distributions.
+To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
+of non-reversible parallel tempering
+(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
+a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
+Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
+to provide a hyperparameter-free implementation.
+Pigeons.jl's simple user interface facilitates running PT
 single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
@@
 A typical instance of the class of problems that Pigeons.jl tackles can be described by
-a probability density\footnote{This density may also be a probability mass function 
-(discrete variables). We also allow for a combination of discrete/continuous variables.} 
+a probability density (or mass) function
+%\footnote{This density may also be a probability mass function 
+%(discrete variables). We also allow for a combination of discrete/continuous variables.} 
 $\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 

### 2025-05-13 workshop/TPG.tex
typo

@@
 
-But this is where Pigeons can help: within a Parallel Tempering (PT) run, PG can be
-used to explore each distribution in an annealing sequence that interpolates between
-the prior and posterior. We expect PG to work more effectively on tempered 
+But this is where Pigeons can help: within a PT run, PG can be
+used to explore each distribution in the annealing path. We expect PG to work 
+more effectively on tempered 
 distributions closer to the prior. Such samples can then be transported towards 

### 2025-05-13 workshop/new_developments.tex
new devels + fixes

\section{Recent developments}

\paragraph{autoMALA}
Since version v0.2.0, Pigeons.jl's default explorer for most models in continuous
space has been autoMALA \citep{biron2024automala}, a novel MCMC sampler
based on the Metropolis-adjusted Langevin algorithm (MALA). autoMALA continually
adapts its step size at each iteration based on the local geometry of the target
distribution. Such automatic adjustments are highly desirable in the local exploration 
phase of PT (see \cref{sec:PT} for a review of PT), since otherwise we would 
need to find optimal step sizes for each distribution in the annealing path---a 
prohibitively costly process.

\vspace{-1em}

\paragraph{Support for BUGS models}
The latest version of Pigeons.jl at the time of writing (v0.4.9) added support 
for sampling from programs written in BUGS \cite{lunn2000winbugs,lunn2009bugs,
lunn2013bugs}, by leveraging the Julia implementation of the 
language in JuliaBUGS.jl\footnote{\url{https://turinglang.org/JuliaBUGS.jl/stable/}}.

\vspace{-1em}

\paragraph{Tempered Particle Gibbs for Turing-complete PPLs}
Particle Gibbs (PG, \citealp{andrieu2010particle}) is an MCMC
algorithm that has been shown \citep{wood2014new} to be a natural sampler for
arbitrary programs written in Turing-complete probabilistic programming 
languages (PPLs). For example, the Julia package Turing.jl \citep{ge2018turing}
provides a PG sampler that can handle any program written in DynamicPPL
\citep{tarek2020dynamicppl}, its underlying PPL.
However, this generality comes at the expense of poor performance in moderately
complex models. The reason is that PG is, essentially, a smart 
way for selecting samples from the prior that are close to the posterior.
When these distributions are very different, PG can be inefficient.

With Pigeons.jl we can partially alleviate this issue by using PG as an explorer
in PT. Indeed, PG should be efficient for exploring distributions closer to the 
prior. Such samples can then be transported towards the posterior chain via the PT process.
In order to construct the annealing sequence of distributions for an arbitrary 
PPL program, it suffices to inject an inverse temperature parameter 
$\beta\in[0,1]$ in the call to compute the log density of every \texttt{observe} 
statement \citep[see][for an explanation of this syntax]{gordon2014probabilistic} 
in the program. Leveraging multiple dispatch in Julia, we have
successfully applied this technique to DynamicPPL programs in our package
TPGExplorers.jl\footnote{\url{https://github.com/Julia-Tempering/TPGExplorers.jl}}.

### 2025-05-23 workshop/PT.tex
ns edits

@@
 Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
-in \cite{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
+in \citet{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
 This section gives both a brief overview of PT and some details of 

### 2025-05-23 workshop/SPI.tex
compress
ns edits

@@
 \subsection{Distributed reduction and floating point non-associativity}
-After each round of PT, the machines need to exchange information such as the 
+After each round of PT, 
+\NS{I think we should assume that the readers don't know much about PT, so maybe 
+this part should be simplified}
+the machines need to exchange information such as the 
 average swap acceptance probabilities, statistics to adapt a variational 
@@
 of the \texttt{Replica}s/machines and is an example of a reduction scheme.
-
-To illustrate why distributed reduction with floating point values can violate 
-strong parallelism invariance if not properly implemented, we consider the 
-following toy example. 
-Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
-as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
-in the following examples.
-In this case, if our reduction procedure is to sum the floating point numbers, 
-we know that our final answer should be approximately $36x$. 
-However, depending on the exact order in which floating point addition is carried out, 
-the answers might not all be the same and exactly equal to $36x$.
-For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
-for eight machines would be given by 
-\[
-  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
-  \approx 978.5814582452562.
-\]
-In contrast, with two machines, one possible order of operations might be 
-\[
-  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
-  \approx 978.5814582452563.
-\]
-A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.
-
-To avoid the issue of non-associativity of 
-floating point arithmetic, we ensure that the order in which operations are performed 
-is exactly the same, irrespective of the number of processes/machines and threads. 
-This is achieved by making sure that every value to be added---if addition is our 
-reduction operation---is a leaf node in a reduction tree that is invariant to the 
-number of machines available to perform the computation. 
-For instance, if $N$ values are to be reduced, then the reduction tree would have 
-$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
-in such a way that the order of operations is as if there were $N$ machines available. 
-To do so, it may be necessary for a machine to ``communicate with itself'', 
-imitating the behaviour that would be present if there were $N$ machines available.
-\cref{fig:reduction_tree_colour} illustrates the reduction procedure for eight and 
-two machines.
 
@@
 
+To illustrate why distributed reduction with floating point values can violate 
+strong parallelism invariance if not properly implemented, we consider the 
+following toy example. 
+Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
+as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
+in the following examples.
+In this case, if our reduction procedure is to sum the floating point numbers, 
+we know that our final answer should be approximately $36x$. 
+However, depending on the exact order in which floating point addition is carried out, 
+the answers might not all be the same and exactly equal to $36x$.
+For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
+for eight machines would be given by 
+\[
+  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
+  \approx 978.5814582452562.
+\]
+In contrast, with two machines, one possible order of operations might be 
+\[
+  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
+  \approx 978.5814582452563.
+\]
+A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.
+
 \begin{figure*}[t]
@@
 
+To avoid the issue of non-associativity of 
+floating point arithmetic, we ensure that the order in which operations are performed 
+is exactly the same, irrespective of the number of processes/machines and threads. 
+This is achieved by making sure that every value to be added---if addition is our 
+reduction operation---is a leaf node in a reduction tree that is invariant to the 
+number of machines available to perform the computation. 
+For instance, if $N$ values are to be reduced, then the reduction tree would have 
+$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
+in such a way that the order of operations is as if there were $N$ machines available. 
+To do so, it may be necessary for a machine to ``communicate with itself'', 
+imitating the behaviour that would be present if there were $N$ machines available.
+\cref{fig:reduction_tree_colour} illustrates the reduction procedure for eight and 
+two machines.
+
 \subsection{Splittable random streams}
@@
 For this reason, with a different number of threads, the result is different:
+% NB: use minipage to avoid splitting the code across pages
+\begin{minipage}{\linewidth}
 \begin{lstlisting}[language = Julia]
@@
 \end{lstlisting}
+\end{minipage}
 
@@
 With one and eight threads, the code above outputs
+% NB: use minipage to avoid splitting the code across pages
+\begin{minipage}{\linewidth}
 \begin{lstlisting}[language = Julia]
@@
 Result: 0.4394633333251359
-\end{lstlisting}
-\begin{lstlisting}[language = Julia]
+
 Num. of threads: 8
@@
 \end{lstlisting}
+\end{minipage}

### 2025-05-23 workshop/abstract.tex
TPG section + reorder
new devels + fixes
ns edits

@@
 \begin{abstract}
+\NS{Feel free to place yourself as first author, Miguel.}
 Sampling from complicated probability distributions---such as 
@@
 software validation. We describe the key features of Pigeons.jl that enable a 
-distributed and randomized algorithm satisfying SPI.
+distributed and randomized algorithm satisfying SPI. Finally, we briefly discuss 
+some recent developments in Pigeons.jl.
 \end{abstract}

### 2025-05-23 workshop/examples.tex
ns edits

--- 2025-05-06 workshop/examples.tex
@@
 state-of-the-art sampling algorithm, to sample from the target distribution.
-PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
+PT comes with several tuning parameters and \citet{syed2021nrpt} describe how 
 to select these parameters effectively, which Pigeons.jl implements under the hood.
@@
 
-    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
+    \item \texttt{$\Lambda$}: the global communication barrier, as described in \citet{syed2021nrpt}, 
     which measures the inherent difficulty of the sampling problem.
     A rule of thumb to configure the number of PT chains is also 
-    given by \cite{syed2021nrpt}, where they suggest that stable performance 
+    given by \citet{syed2021nrpt}, where they suggest that stable performance 
     should be achieved when the number of chains is set to roughly 2$\Lambda$.
@@
 can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
-online documentation. For instance, we can also compute constant-memory estimates 
+online documentation. 
+\NS{We should link to the online documentation here and also earlier.}
+For instance, we can also compute constant-memory estimates 
 of extrema of our distribution.
@@
 We describe how to produce some key parallel tempering diagnostics from 
-\cite{syed2021nrpt}.
+\citet{syed2021nrpt}.
 
@@
 used to approximately inform the appropriate number of chains. 
-Based on \cite{syed2021nrpt}, stable PT performance should be achieved when  
+Based on \citet{syed2021nrpt}, stable PT performance should be achieved when  
 the number of chains is set to roughly $2\Lambda$. This can be achieved by 
@@
 that lies as close as possible to the target distribution in order to improve 
-the efficiency of PT. We refer readers to \cite{surjanovic2022vpt} for more details.
+the efficiency of PT. We refer readers to \citet{surjanovic2022vpt} for more details.
 When only continuous parameters are of interest, we encourage users to consider 

### 2025-05-23 workshop/introduction.tex
ns edits

@@
 single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
-We have stress-tested Pigeons successfully with up to 1000 processes running 
+We have stress-tested Pigeons successfully with up to 1,000 processes running 
 concurrently on a compute cluster (Sockeye) at the University of British Columbia.

### 2025-05-26 workshop/SPI.tex
NS commnets

--- 2025-05-07 workshop/SPI.tex
@@
 \label{sec:PI_causes}
-In this section we describe potential violations of 
-strong parallelism invariance (SPI) that can occur in a distributed setting. 
-We also explain how we avoid these issues by using special 
+In this section we describe potential violations of SPI that can occur in a 
+distributed setting. We also explain how we avoid these issues by using special 
 distributed reduction schemes and splittable random number generators. 
@@
 \subsection{Distributed reduction and floating point non-associativity}
-After each round of PT, the machines need to exchange information such as the 
-average swap acceptance probabilities, statistics to adapt a variational 
-reference and adjust annealing parameters, and so on. 
-For instance, suppose our \texttt{state} is univariate and real-valued and 
-that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
-is visited as well as the mean of the univariate \texttt{state}s from the target chain.
-To obtain the final estimate of the mean of the target distribution, we would like 
-to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
-the number of times that each \texttt{Replica} visited the target chain.   
-This process involves summing floating-point values that are located on each 
-of the \texttt{Replica}s/machines and is an example of a reduction scheme.
-
-To illustrate why distributed reduction with floating point values can violate 
-strong parallelism invariance if not properly implemented, we consider the 
-following toy example. 
-Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
-as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
-in the following examples.
-In this case, if our reduction procedure is to sum the floating point numbers, 
-we know that our final answer should be approximately $36x$. 
-However, depending on the exact order in which floating point addition is carried out, 
-the answers might not all be the same and exactly equal to $36x$.
-For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
-for eight machines would be given by 
-\[
-  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
-  \approx 978.5814582452562.
-\]
-In contrast, with two machines, one possible order of operations might be 
-\[
-  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
-  \approx 978.5814582452563.
-\]
-A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.
-
-To avoid the issue of non-associativity of 
-floating point arithmetic, we ensure that the order in which operations are performed 
-is exactly the same, irrespective of the number of processes/machines and threads. 
-This is achieved by making sure that every value to be added---if addition is our 
-reduction operation---is a leaf node in a reduction tree that is invariant to the 
-number of machines available to perform the computation. 
-For instance, if $N$ values are to be reduced, then the reduction tree would have 
-$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
-in such a way that the order of operations is as if there were $N$ machines available. 
-To do so, it may be necessary for a machine to ``communicate with itself'', 
-imitating the behaviour that would be present if there were $N$ machines available.
-\cref{fig:reduction_tree_colour} illustrates the reduction procedure for eight and 
-two machines.
+The execution of PT in Pigeons.jl is divided into \emph{rounds}, which are
+sequences of \emph{scans}---a single call to \cref{alg:distributed_PT}---of 
+exponentially increasing length. At the end of each round,
+\emph{replicas}---the fundamental unit of parallelization in PT (see \cref{sec:PT} for 
+details)---need to exchange information. This can be related to the output of the 
+program---e.g., the value of the normalization constant $Z$---as well as statistics 
+needed to adapt the sampler so that the next round runs more efficiently.
+Almost all of these cases involve summing floating-point numbers that are 
+located on different threads, processes, or compute nodes (in the following we use 
+the generic term \emph{machine}).
 
@@
 
+To illustrate why distributed reduction with floating point values can violate 
+strong parallelism invariance if not properly implemented, we consider the 
+following toy example. 
+Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
+as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
+in the following examples.
+In this case, if our reduction procedure is to sum the floating point numbers, 
+we know that our final answer should be approximately $36x$. 
+However, depending on the exact order in which floating point addition is carried out, 
+the answers might not all be the same and exactly equal to $36x$.
+For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
+for eight machines would be given by 
+\[
+  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
+  \approx 978.5814582452562.
+\]
+In contrast, with two machines, one possible order of operations might be 
+\[
+  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
+  \approx 978.5814582452563.
+\]
+\cref{fig:reduction_tree_2} shows a possible reduction tree for two machines.
+
 \begin{figure*}[t]
@@
 A first step to achieve this is to associate one random number 
-generator to each \texttt{Replica}. 
+generator to each replica. 
 To do so, we use our SplittableRandoms.jl package, which is a Julia implementation 

### 2025-05-26 workshop/abstract.tex
NS commnets

--- 2025-05-07 workshop/abstract.tex
@@
 multimodal posteriors arising in Bayesian inference and high-dimensional 
-distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
-a simple user interface to perform such computations single-threaded, multi-threaded,
-and/or distributed over thousands of MPI-communicating machines. In this paper,
-we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
-a guarantee that the output for a given seed is identical irrespective of the number 
-of threads and processes. This feature is crucial for scientific reproducibility and 
-software validation. We describe the key features of Pigeons.jl that enable a 
-distributed and randomized algorithm satisfying SPI.
+distributions in statistical mechanics---is a challenging problem. \emph{Pigeons.jl} 
+is a Julia package that implements \emph{Parallel Tempering}, a technique that 
+leverages parallelism to improve the effectiveness of \emph{Markov chain Monte Carlo} 
+algorithms. Pigeons.jl provides a simple user interface to perform such computations 
+using single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
+machines. In this paper, we discuss a feature of Pigeons.jl that we call \emph{strong 
+parallelism invariance} (SPI): a guarantee that the output for a given seed is identical 
+irrespective of the number of threads and processes. This feature is crucial for scientific 
+reproducibility and software validation. We describe the key features of Pigeons.jl that 
+enable a distributed and randomized algorithm satisfying SPI. Finally, we briefly discuss 
+some recent developments in Pigeons.jl.
 \end{abstract}

### 2025-05-26 workshop/examples.tex
NS commnets

@@
 can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
-online documentation. 
-\NS{We should link to the online documentation here and also earlier.}
+online documentation\footnote{\url{https://pigeons.run/dev/}}.
 For instance, we can also compute constant-memory estimates 

### 2025-05-26 workshop/introduction.tex
NS commnets

@@
 
-Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
+Pigeons.jl\footnote{Documentation available at \url{https://pigeons.run/dev/}. 
+Public repository: \url{https://github.com/Julia-Tempering/Pigeons.jl}.}
 enables users to sample efficiently from challenging probability distributions.
 To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
-of non-reversible parallel tempering
+of non-reversible \emph{parallel tempering}
 (PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 

### 2025-06-26 workshop/PT.tex
replace PT appendix with nutshell section

@@
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.25\textwidth]{../img/bimodal.pdf}
-    \caption{A simple bimodal distribution from which traditional 
-    MCMC methods may struggle to obtain samples. Blue lines display output 
-    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
-    The sampler in this figure is visibly trapped in one of the two modes.}
-    \label{fig:bimodal}
-\end{figure}
+%\begin{figure}[t]
+%    \centering
+%    \includegraphics[width=0.25\textwidth]{../img/bimodal.pdf}
+%    \caption{A simple bimodal distribution from which traditional 
+%    MCMC methods may struggle to obtain samples. Blue lines display output 
+%    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
+%    The sampler in this figure is visibly trapped in one of the two modes.}
+%    \label{fig:bimodal}
+%\end{figure}
 
@@
 
-% NB: this is placed here just to make it appear in a different page
-\begin{figure}[t]
-    \centering
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
-      \caption*{$\pi_1$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
-      \caption*{$\pi_2$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
-      \caption*{$\pi_3$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
-      \caption*{$\pi_4$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
-      \caption*{$\pi_5$}
-    \end{minipage}
-    \begin{minipage}{0.15\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
-      \caption*{$\pi_6$}
-    \end{minipage}
-    \caption{Heatmaps of six distributions lying on an annealing path 
-    from a unimodal reference distribution  
-    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
-    $\pi_N$, which is in this case the bimodal distribution from 
-    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
-    are not directly comparable because the densities of intermediate distributions 
-    are not normalized.)}
-    \label{fig:path}
-\end{figure}
+%% NB: this is placed here just to make it appear in a different page
+%\begin{figure}[t]
+%    \centering
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
+%      \caption*{$\pi_1$}
+%    \end{minipage}
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
+%      \caption*{$\pi_2$}
+%    \end{minipage}
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
+%      \caption*{$\pi_3$}
+%    \end{minipage}
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
+%      \caption*{$\pi_4$}
+%    \end{minipage}
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
+%      \caption*{$\pi_5$}
+%    \end{minipage}
+%    \begin{minipage}{0.15\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
+%      \caption*{$\pi_6$}
+%    \end{minipage}
+%    \caption{Heatmaps of six distributions lying on an annealing path 
+%    from a unimodal reference distribution  
+%    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
+%    $\pi_N$, which is in this case the bimodal distribution from 
+%    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
+%    are not directly comparable because the densities of intermediate distributions 
+%    are not normalized.)}
+%    \label{fig:path}
+%\end{figure}
 
@@
 
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=0.33\textwidth]{../img/speedup_distributed.pdf}
-    \hspace{4em}
-    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
-    \caption{
-        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
-        and parallel (Julia threads, {\bf right}) settings 
-        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
-        The blue dotted line represents a hypothetical speedup equal to the 
-        number of processes/threads. 
-        The red dashed line indicates a speedup of one for reference.
-    }
-    \label{fig:PT_scaling}
-\end{figure*}
+%\begin{figure*}[t]
+%    \centering
+%    \includegraphics[width=0.33\textwidth]{../img/speedup_distributed.pdf}
+%    \hspace{4em}
+%    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
+%    \caption{
+%        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
+%        and parallel (Julia threads, {\bf right}) settings 
+%        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
+%        The blue dotted line represents a hypothetical speedup equal to the 
+%        number of processes/threads. 
+%        The red dashed line indicates a speedup of one for reference.
+%    }
+%    \label{fig:PT_scaling}
+%\end{figure*}
 

### 2025-06-26 workshop/SPI.tex
acc + fix tree lengths
replace PT appendix with nutshell section

@@
-\section{Strong parallelism invariance}
-\label{sec:PI_causes}
-In this section we describe potential violations of SPI that can occur in a 
-distributed setting. We also explain how we avoid these issues by using special 
-distributed reduction schemes and splittable random number generators. 
-Insights provided in this section can be applied 
-to general distributed software beyond Julia.
+\section{Strong parallelism invariance}\label{sec:PI_causes}
+
+In this section we describe potential violations of \emph{strong 
+parallelism invariance} (SPI)---which we define as a guarantee that the output 
+for a given seed is identical irrespective of the number of threads and 
+processes---that can occur in a distributed setting. We also explain how we 
+avoid these issues by using special distributed reduction schemes and 
+splittable random number generators. Insights provided in this section 
+can be applied to general distributed software beyond Julia.
 
@@
 The execution of PT in Pigeons.jl is divided into \emph{rounds}, which are
-sequences of \emph{scans}---a single call to \cref{alg:distributed_PT}---of 
-exponentially increasing length. At the end of each round,
-\emph{replicas}---the fundamental unit of parallelization in PT (see \cref{sec:PT} for 
-details)---need to exchange information. This can be related to the output of the 
+sequences of \emph{scans}---an exploration phase followed by a communication phase---%
+of exponentially increasing length. At the end of each round,
+\emph{replicas}---the fundamental unit of parallelization in PT---need to exchange information. This can be related to the output of the 
 program---e.g., the value of the normalization constant $Z$---as well as statistics 
@@
 \begin{figure}[t]
+  \footnotesize
   \centering
@@
 \begin{figure*}[t]
+  \footnotesize
   \centering

### 2025-06-26 workshop/introduction.tex
replace PT appendix with nutshell section

@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.7\linewidth]{../img/speedup_distributed.pdf}
+%    \hspace{4em}
+%    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
+    \caption{
+        Speedup relative to single-thread execution for Pigeons.jl in the distributed 
+        (MPI) setting. Error bars denote approximate 95\% confidence intervals.
+        The blue dotted line represents a hypothetical speedup equal to the 
+        number of processes. The red dashed line indicates a speedup of one for reference.
+    }
+    \label{fig:PT_scaling}
+\end{figure}
+
 Pigeons.jl\footnote{Documentation available at \url{https://pigeons.run/dev/}. 
@@
 (PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
-a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
+a state-of-the-art MCMC algorithm.
 Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
@@
 \end{enumerate}
-
-\section{Overview of Pigeons.jl}
-A typical instance of the class of problems that Pigeons.jl tackles can be described by
-a probability density (or mass) function
-%\footnote{This density may also be a probability mass function 
-%(discrete variables). We also allow for a combination of discrete/continuous variables.} 
-$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
-density $\pi$ is typically known only up to a normalizing constant, 
-\[
-\label{eq:normalizing_constant}
-  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
-\]  
-where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
-Pigeons.jl takes as input the function $\gamma$.
-
-The output of Pigeons.jl can be used for two main tasks:
-\begin{enumerate}
-    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
-    For example, the choice $f(x) = x$ computes the mean and 
-    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
-    where $I[\cdot]$ denotes the indicator function.
-
-    \item Approximating the value of the normalization constant $Z$. For 
-    example, in Bayesian statistics, this corresponds to the 
-    marginal likelihood, which can be used for model selection. 
-\end{enumerate}
-The Pigeons package particularly shines compared to traditional sampling approaches in the 
-following scenarios:
-\begin{itemize}
-    \item When the target density $\pi$ is challenging due to a complex structure 
-    (e.g., high-dimensional, multi-modal, etc.).
-
-    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
-    the normalizing constant $Z$. 
-    Many existing tools focus on the former and struggle or fail to do the latter.
-
-    \item When the target distribution $\pi$ is defined over a non-standard space, 
-    e.g. a combinatorial object such as a phylogenetic tree.  
-\end{itemize}

### 2025-06-26 workshop/new_developments.tex
replace PT appendix with nutshell section

@@
 adapts its step size at each iteration based on the local geometry of the target
-distribution. Such automatic adjustments are highly desirable in the local exploration 
-phase of PT (see \cref{sec:PT} for a review of PT), since otherwise we would 
+distribution. Such automatic adjustments are highly desirable in the exploration 
+phase, since otherwise we would 
 need to find optimal step sizes for each distribution in the annealing path---a 

### 2025-06-26 workshop/nutshell.tex
replace PT appendix with nutshell section

\section{Parallel Tempering in a nutshell}

\begin{figure}[t]
  \centering
    \includegraphics[width=0.5\linewidth]{../img/bimodal.pdf}
    \caption{A bimodal distribution. Blue lines display output 
    from 1,000 iterations of a random walk Metropolis-Hastings sampler.}
    \label{fig:bimodal}
\end{figure}

Consider the problem of sampling from a probability density (or mass) function
$\pi(x)$, henceforth called the \emph{target}. Solving this task is often 
challenging, as the 
distribution $\pi$ can exhibit features which traditional MCMC methods---from
random walk Metropolis-Hastings to Hamiltonian Monte Carlo---struggle to
describe accurately. For example, in bimodal distributions such as the one 
illustrated in \cref{fig:bimodal}, traditional methods might remain stuck in one 
of the two modes for an extremely long period of time.

\begin{figure}[t]
  \centering
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_1.pdf}
%      \caption*{$\pi_1$}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_2.pdf}
%      \caption*{$\pi_2$}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_3.pdf}
%      \caption*{$\pi_3$}
    \end{minipage}\\
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_4.pdf}
%      \caption*{$\pi_4$}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_5.pdf}
%      \caption*{$\pi_5$}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
      \centering
      \includegraphics[width=\linewidth]{../img/heatmap_path_6.pdf}
%      \caption*{$\pi_6$}
    \end{minipage}
    \caption{Heatmaps of six distributions lying on an annealing path, 
    starting from a unimodal reference (top-left), and ending at a target
    distribution from \cref{fig:bimodal}.}
    \label{fig:path}
\end{figure}

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to the target $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$---%
called the \emph{reference} distribution---with the sampling difficulty increasing 
as one approaches $\pi_N$. An example of such a sequence of distributions, referred 
to as an \emph{annealing path}, is shown in \cref{fig:path}.

PT operates by first obtaining samples from each distribution on the path in parallel;
this is called an \emph{exploration} phase. Then, in the \emph{communication} phase, 
PT proposes swapping the samples between adjacent distributions. This step is crucial:
successive swaps along the path amount to transporting samples from the reference
to the target. This allows for the discovery of new regions of the space of the 
target distribution such as the top-right mode in \cref{fig:bimodal}.

### 2025-06-26 workshop/overview.tex
replace PT appendix with nutshell section

--- 2024-05-29 paper/introduction.tex
@@
-\section{Introduction}
-In many scientific application domains, the ability to obtain samples from a 
-probability distribution is of central interest. 
-For instance, sampling methods have been used to discover magnetic polarization 
-in the black hole of galaxy M87 \cite{akiyama2021seven}
-and to image the Sagittarius A* black hole \cite{akiyama2022first}.
-They have also been used to 
-model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
-infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
-and to identify gerrymandering in Georgia's 2021 congressional districting plan 
-\cite{zhao2022mathematically}.
-Similarly, evaluating high-dimensional integrals or sums over complicated 
-combinatorial spaces are related tasks that can also be solved with sampling 
-via Markov chain Monte Carlo (MCMC) methods. 
-However, such calculations can often be bottlenecks in the scientific process, with 
-simulations that can last days or even weeks to finish.
+\section{Overview of Pigeons.jl}\label{sec:overview}
 
-\medskip
-Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
-\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
-Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
-enables users to sample efficiently from challenging probability distributions 
-and solve integration problems by 
-implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
-that leverage distributed computation. Its simple user interface facilitates performing such computations 
-single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
-We have stress-tested Pigeons successfully with up to 1000 machines running 
-concurrently on a compute cluster (Sockeye) at the University of British Columbia.
-Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
-the output for a given seed is \emph{identical} irrespective 
-of the number of threads or processes. 
-Such a level of reproducibility is rare in distributed software but of 
-great use for the purposes of debugging in the context of sampling algorithms, 
-which produce stochastic output.
-Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
-\begin{enumerate}
-    \item one machine running on one thread;
-    \item one machine running on several threads;
-    \item several machines running, each using one thread, and
-    \item several machines running, each using several threads.
-\end{enumerate}
+Pigeons.jl exploits the fact that the exploration phase in PT is embarrassingly parallel
+to provide a multithreaded and distributed Julia implementation of the algorithm.
 
-\subsection{Latest documentation}
-
-More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.
-
-\subsection{Problem formulation}
-We describe the class of problems that can be approached using Pigeons.jl.
-Let $\pi(x)$ denote a probability density\footnote{This density may 
-also be a probability mass function 
-(discrete variables). We also allow for a combination of discrete/continuous variables.} 
-called the \emph{target}. 
-In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
-known only up to a normalizing constant, 
+In many problems, e.g. in Bayesian 
+statistics, the target $\pi$ is typically known only up to a normalizing constant, 
 \[
@@
 \end{itemize}
-
-\subsection{What is of interest to the general Julia developer?}
-Ensuring code correctness at the intersection of randomized, parallel, and 
-distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
-based on a principle that we refer to as \textit{strong parallelism invariance}
-(SPI).
-Namely, the output of Pigeons.jl is  invariant to the number of machines 
-and/or threads.
-Without explicitly keeping SPI in mind during software construction, 
-the (random) output of the algorithm is only guaranteed to have the same distribution.
-This is a much weaker guarantee that, in particular, makes debugging difficult.
-However, with our notion of SPI we make debugging and 
-software validation considerably easier. This is because the developer can 
-first focus on the fully serial randomized algorithm, and then use it as an 
-easy-to-compare gold-standard reference for parallel/distributed implementations. 
-This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
-testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
-incurs additional false negatives due to statistical error.
-
-\medskip 
-The general Julia developer will be interested in: 
-\begin{itemize}
-    \item The main causes of a violation of strong parallelism invariance that we have identified 
-    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
-    them in Pigeons.jl.
-
-    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
-    package that was developed by our team 
-    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
-\end{itemize}

### 2025-06-26 workshop/examples.tex
replace PT appendix with nutshell section
finish

@@
-\section{Examples}
+\section{Examples}\label{app:examples}
 In this section we present a set of minimal examples 
@@
     acceptance rates during the communication phase across the PT chains.
-    See \cref{sec:PT} for a description of PT communication.
 \end{itemize}

### 2025-06-26 workshop/overview.tex
finish

@@
 \end{itemize}
+
+There are two fundamental concepts in the implementation that we will use
+in the following sections. For a distribution $\pi_i$ in the path, we call its
+index the $i$-th \emph{chain}. During execution, we assign \emph{replicas} to 
+work on the distributions associated to the chains. The map between chains and
+replicas is one-to-one, and thus can be seen as a permutation of the 
+set $\{1,\dots,N\}$. Replicas are the fundamental unit of parallelization in PT:
+they may be distributed across threads, processes, and machines.
+
+In \cref{app:examples} we provide examples of how to use Pigeons.jl. For more details
+on the implementation, see \citet{surjanovic2025pigeons}.

### 2025-06-28 workshop/PT.tex
minor changes

@@
 $\pi$ can be challenging for traditional MCMC methods---such 
-as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
+as Metropolis--Hastings, slice sampling, and Hamiltonian Monte Carlo---
 because of its structure.

### 2025-06-28 workshop/nutshell.tex
minor changes

@@
-\section{Parallel Tempering in a nutshell}
+\section{Parallel tempering in a nutshell}
 
@@
     \caption{A bimodal distribution. Blue lines display output 
-    from 1,000 iterations of a random walk Metropolis-Hastings sampler.}
+    from 1,000 iterations of a random walk Metropolis--Hastings sampler.}
     \label{fig:bimodal}
@@
 distribution $\pi$ can exhibit features which traditional MCMC methods---from
-random walk Metropolis-Hastings to Hamiltonian Monte Carlo---struggle to
+random walk Metropolis--Hastings to Hamiltonian Monte Carlo---struggle to
 describe accurately. For example, in bimodal distributions such as the one 
@@
     \caption{Heatmaps of six distributions lying on an annealing path, 
-    starting from a unimodal reference (top-left), and ending at a target
+    starting from a unimodal reference (top-left), and ending at the target
     distribution from \cref{fig:bimodal}.}

### 2025-07-10 workshop/paper.tex
TPG section + reorder
new devels + fixes
NS commnets
replace PT appendix with nutshell section
finish
workshop submission file

@@
 \begin{icmlauthorlist}
+\icmlauthor{Miguel Biron-Lattes}{sfu}
 \icmlauthor{Nikola Surjanovic}{ubc}
-\icmlauthor{Miguel Biron-Lattes}{sfu}
 \icmlauthor{Paul Tiede}{harvard}
 \icmlauthor{Saifuddin Syed}{oxford}
+\icmlauthor{Trevor Campbell}{ubc}
 \icmlauthor{Alexandre Bouchard-C{\^o}t{\'e}}{ubc}
-\icmlauthor{Trevor Campbell}{ubc}
 \end{icmlauthorlist}
 
-\icmlaffiliation{harvard}{Center for Astrophysics, Harvard University, United States}
+\icmlaffiliation{harvard}{Center for Astrophysics and Black Hole Initiative, Harvard University, United States}
 \icmlaffiliation{oxford}{Department of Statistics, University of Oxford, United Kingdom}
@@
 
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+\printAffiliationsAndNotice{
+This is a summarized and revised version of \citet{surjanovic2025pigeons}.
+}  % leave blank if no need to mention equal contribution
 
@@
 \input{introduction.tex}
+\input{nutshell}
+\input{overview}
 \input{SPI.tex}
-\input{related.tex}
-%\input{conclusion.tex}
+\input{new_developments}
+%\input{related.tex}
+\input{conclusion}
 %\input{ack}
@@
 \bibliographystyle{icml2025}
-\bibliography{../paper/ref.bib}
+\bibliography{../paper/ref.bib,ref.bib}
 
@@
 \input{examples}
-\input{PT}
+%\input{PT}
 
